# Transportation Planning {#sec-transport}

## Introduction

Transportation is typically the single largest cost component in a supply chain, accounting for 40–70% of total logistics expenditure. Transportation planning sits at the interface of strategic network design and daily operations.

## Transportation Modes

The choice of transport mode fundamentally shapes a supply chain's responsiveness-efficiency profile.

### Value Density and Mode Selection

**Value density** (product value per unit weight or volume) is the primary driver of mode selection:

- **High value density** (electronics, pharmaceuticals, luxury goods): Can absorb air freight premiums
- **Low value density** (bulk chemicals, agricultural commodities): Sea or rail only
- **Medium** (consumer goods, machinery): Road or intermodal

```{r}
#| label: tbl-modes
#| tbl-cap: "Transport mode comparison"
library(kableExtra)

modes <- data.frame(
  Mode = c("Road (full truck)", "Rail (electric)", "Air cargo", "Sea (container)", "Inland waterway"),
  Speed = c("Medium", "Slow", "Fast", "Very slow", "Very slow"),
  `Cost_per_tkm` = c("€0.08–0.15", "€0.03–0.06", "€2–5", "€0.01–0.03", "€0.02–0.04"),
  `CO2_g_per_tkm` = c(62, 6, 500, 8, 30),
  check.names = FALSE
)

kbl(modes, booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## The Transportation Problem

### Problem Formulation

The **Transportation Problem (TP)** is a classic linear program that finds the minimum-cost allocation of flows from supply nodes to demand nodes.

**Given:**
- $m$ supply nodes $i = 1, \ldots, m$ with capacities $a_i$
- $n$ demand nodes $j = 1, \ldots, n$ with requirements $b_j$
- Unit transport cost $c_{ij}$ from $i$ to $j$

**Decision:** Flow $x_{ij} \geq 0$ (units shipped from $i$ to $j$)

$$\min \sum_{i=1}^m \sum_{j=1}^n c_{ij} x_{ij}$$

Subject to:
$$\sum_{j=1}^n x_{ij} \leq a_i \quad \forall i \quad \text{(supply)}$$
$$\sum_{i=1}^m x_{ij} = b_j \quad \forall j \quad \text{(demand)}$$
$$x_{ij} \geq 0$$

::: {.callout-note}
## Key Property: Integer Optimality
The TP constraint matrix is **totally unimodular**, which means the LP relaxation always yields an integer solution. We do not need integer programming — a standard LP solver suffices.
:::

### Vogel's Approximation Method (VAM)

VAM is a **regret-based greedy heuristic** that typically finds solutions within 5% of optimal.

**Algorithm:**
1. For each row and column, compute the **regret**: difference between 2nd cheapest and cheapest available cost
2. Select the row/column with maximum regret
3. Allocate maximum possible flow to the cheapest route in that row/column
4. Update remaining capacities/demands; eliminate satisfied row/column
5. Repeat until all flows assigned

### Solving with R

```{r}
#| label: tp-solve
#| echo: true
library(lpSolve)

# Pear Disk Drive example
cost_mat <- matrix(c(250, 420, 380, 280,
                     1280, 990, 1440, 1520,
                     1550, 1420, 1660, 1730),
                   nrow = 3, byrow = TRUE,
                   dimnames = list(c("Sunnyvale", "Dublin", "Bangkok"),
                                   c("Amarillo", "Teaneck", "Chicago", "Sioux Falls")))

supply <- c(45, 120, 95)
demand <- c(80, 78, 47, 55)

solution <- lp.transport(cost_mat, "min",
                         rep("<=", 3), supply,
                         rep(">=", 4), demand)

flow_mat <- matrix(solution$solution, nrow = 3,
                   dimnames = dimnames(cost_mat))

cat("Optimal total cost: $", solution$objval, "\n")
print(flow_mat)
```

## The Transshipment Problem

Real distribution networks often include **intermediate nodes** — cross-docks, ports, rail terminals — through which goods are routed but not consumed.

The Transshipment Problem extends the TP by adding a third tier of nodes $V_T$ between supply and demand.

### Transformation to Standard TP

Key insight: Duplicate each transshipment node $j \in V_T$ into supply copy $j$ and demand copy $j'$. Set artificial capacity $L = \sum_i a_i$. Assign cost 0 for self-routing and $M$ (large penalty) for infeasible routes.

The resulting $(|V_a| + |V_T|) \times (|V_T| + |V_d|)$ matrix can be solved as a standard TP.

## Exercises

1. Apply VAM to the Pear Disk Drive example. Show each iteration's regret calculation.

2. Dublin's plant capacity is unexpectedly cut to 80 (000 units). How does the optimal solution change? What is the cost penalty?

3. **Battery Logistics** (Case Study data): Using the Haversine distances, set up the TP for the 3-warehouse/11-plant configuration (Düren, Harsewinkel, Langenhagen). Solve with `lp.transport()`.

```{r}
#| eval: false
#| echo: true
# Haversine function for exercise
haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371  # Earth radius in km
  phi1 <- lat1 * pi / 180
  phi2 <- lat2 * pi / 180
  dphi <- (lat2 - lat1) * pi / 180
  dlambda <- (lon2 - lon1) * pi / 180
  a <- sin(dphi/2)^2 + cos(phi1) * cos(phi2) * sin(dlambda/2)^2
  2 * R * asin(sqrt(a))
}

# Load plant and warehouse data
plants <- read.csv("../../data/battery_plants.csv")
warehouses <- read.csv("../../data/battery_warehouses.csv")

# Subset to open warehouses
open_wh <- warehouses[warehouses$id %in% c("L2", "L4", "L5"), ]

# Build distance matrix
dist_mat <- outer(1:nrow(open_wh), 1:nrow(plants), Vectorize(function(i, j) {
  haversine(open_wh$latitude[i], open_wh$longitude[i],
            plants$latitude[j], plants$longitude[j])
}))

# Transport cost: €0.08 per km per MWh
cost_per_unit <- 0.08  # €/km/MWh
cost_mat_battery <- dist_mat * cost_per_unit
```
