# Sustainable Supply Chain Management {#sec-sustainability}

## Why Sustainability Matters in SCM

Approximately **70% of a company's carbon emissions** come from its supply chain (Scope 3 emissions under the GHG Protocol). This makes supply chain managers directly responsible for sustainability outcomes.

**Regulatory drivers**:
- EU Corporate Sustainability Reporting Directive (CSRD): Mandatory SC sustainability reporting from 2025
- EU Supply Chain Act (CSDDD): Due diligence on human rights and environment in supply chains
- EU ETS (Emissions Trading System): Carbon pricing increasingly internalized

## The Triple Bottom Line in SCM

| Dimension | Key Metrics | SCM Levers |
|-----------|-------------|------------|
| **Economic** | Cost, ROI, resilience | Efficiency optimization, risk management |
| **Environmental** | CO₂e, water, waste | Mode shift, green procurement, circular design |
| **Social** | Labor standards, safety | Supplier auditing, fair trade, EPR |

## Carbon Accounting in Transport

### Emission Factors by Mode

```{r}
#| label: tbl-emissions
#| tbl-cap: "CO₂ equivalent emission factors by transport mode"
library(kableExtra)

ef <- data.frame(
  Mode = c("Road truck (full)", "Rail (EU avg electric)", "Sea (container)", 
           "Air cargo (long haul)", "Inland waterway"),
  `CO2e (g/tkm)` = c(62, 6, 8, 500, 30),
  Source = c("UBA 2023", "UIC 2022", "IMO 2022", "ICAO 2022", "EEA 2023"),
  check.names = FALSE
)
kbl(ef, booktabs = TRUE) |> kable_styling(bootstrap_options = "striped")
```

### Carbon Cost Calculation

With EU ETS carbon price $p_C$ (currently ~€60–80/tCO₂):

$$\text{Carbon cost} = \text{distance} \times \text{volume} \times \text{emission factor} \times p_C$$

```{r}
#| label: carbon-calc
#| echo: true
# Annual carbon cost comparison for Hamburg → Munich (800 km, 10,000 t/yr)
route_km <- 800
volume_t <- 10000

modes <- data.frame(
  mode    = c("100% Road", "80% Rail + 20% Road", "100% Rail"),
  ef_road = c(62, 12.4, 0),   # road component (g/tkm)
  ef_rail = c(0,  4.8,  60),  # rail component (g/tkm)
  stringsAsFactors = FALSE
)

carbon_price <- 70  # €/tCO2

modes$total_co2_t <- (modes$ef_road + modes$ef_rail) * route_km * volume_t / 1e9 * 1e6
# grams to tonnes: g/tkm * km * t = g; / 1e6 = kg; / 1e3 = t
modes$total_co2_t <- (modes$ef_road + modes$ef_rail) * route_km * volume_t / 1e6

modes$carbon_cost_eur <- modes$total_co2_t * carbon_price

print(modes[, c("mode", "total_co2_t", "carbon_cost_eur")])
```

## Circular Supply Chains

### From Linear to Circular

```
Linear:   Extract → Produce → Use → Dispose
Circular: Design for longevity → Reuse → Remanufacture → Recycle → Recover
```

**Reverse logistics** is the operational backbone of circular supply chains:
- **Returns management**: Consumer/B2B returns processing
- **Remanufacturing**: Restore used products to "as-new" condition (automotive, electronics)
- **Recycling flows**: Material recovery
- **Take-back programs**: EPR compliance (packaging, batteries, WEEE)

### Business Case for Circularity

Circular design creates value through:
1. **Retained material value**: Remanufactured parts sell at 30–70% of new part prices but cost 40–60% less to produce
2. **Customer loyalty**: Take-back programs build brand relationships
3. **Regulatory compliance**: Avoid EPR fees and penalties
4. **Resource security**: Reduce dependence on primary raw materials

## Green Procurement

Key elements:
- **Environmental criteria in supplier selection**: ISO 14001 certification, carbon footprint disclosure
- **Life Cycle Assessment (LCA)**: Evaluate environmental impact cradle-to-grave
- **Green supplier development**: Build supplier capability vs. pure selection
- **Total cost of ownership**: Include environmental and social costs

## Exercise

A manufacturer ships 50,000 tonnes of components annually from Hamburg to five European destinations. Current split: 100% road truck.

Proposed: Shift 60% to rail (electric) for destinations ≥ 400 km.

1. Using the emission factors from the table, calculate the annual CO₂ reduction.
2. If rail costs €0.05/tkm more than road and the average road distance is 650 km — what is the annual cost premium?
3. At what carbon price does the modal shift break even financially?
4. With EU ETS at €70/tCO₂, is the modal shift economically justified?
