

```{r}
#| echo: false
#| warning: false
#| message: false

# Load functions from scm_functions.R
source("../R/scm_functions.R")

library(tidyverse)
library(leaflet)
library(readxl)
library(dplyr)
library(plotly)
library(knitr)
library(kableExtra)
library(DiagrammeR)
library(osrm)


```

---

## Strategic Design of Supply Chains {#slide-92}

- Multi-criteria facility location planning
- Continuous facility location planning
- **Discrete facility location planning**

---

## Learning objectives {#learning-objectives-location-discrete}

After this unit, you will be able to …

- **formulate** discrete facility location and network design problems,
- **model** transport, transshipment and location decisions in a network,
- **interpret** key decision variables, objective functions and constraints,
- **evaluate** the results of discrete network models for strategic location decisions.

**Further reading**

- **Transportation, transshipment and network models:** [@guenther2012produktion, Chapter 4].
- **Discrete location models, algorithms and applications:** Chapters 1–4 in [@daskin2013network].
- **Theory and applications of facility location:** [@drezner2002facility].

---

## Planning Distribution Networks {#slide-93}
::: slide-subtitle
Context
:::

::: r-stretch
```{r}
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"
#| fig-height: 1.1

top_boxes <- data.frame(
  label = c("Procurement", "Production", "Distribution", "Sales"),
  x0 = c(0.00, 2.00, 4.00, 6.00),
  x1 = c(2.00, 4.00, 6.00, 8.00),
  y0 = 9.25,
  y1 = 9.75
)


shapes <- c(
  lapply(seq_len(nrow(top_boxes)), function(i) {
    shape_path(
      chevron_path(top_boxes$x0[i], top_boxes$x1[i], top_boxes$y0[i], top_boxes$y1[i]),
      fill = cols$top
    )
  })
)


fig <- plot_ly()

fig <- fig %>%
  add_text(
    x = mid(top_boxes$x0, top_boxes$x1),
    y = mid(top_boxes$y0, top_boxes$y1),
    text = top_boxes$label,
    textfont = list(size = 18, color = "white", family = "Arial"),
    hoverinfo = "skip", showlegend = FALSE
  )  %>%
  layout(
    xaxis = list(visible = FALSE, range = c(0, 8), fixedrange = T),
    yaxis = list(visible = FALSE, range = c(9.1, 9.8), fixedrange = T),
    plot_bgcolor = cols$bg,
    paper_bgcolor = cols$bg,
    margin = list(l = 10, r = 10, t = 10, b = 10),
    shapes = shapes,
    width = 1500,
    height = 100
  )

fig

```

```{r}
#| echo: false
#| warning: false
#| message: false
#| out-width: "90%"
#| fig-height: 5.5


grViz("
digraph network {

  graph [
    rankdir = LR,
    bgcolor = 'transparent',
    nodesep = 0.35,
    ranksep = 0.75,
    splines = line
  ]

  node [
    shape = box,
    style = filled,
    fillcolor = '#DCEAF0',
    color = '#4B5563',
    fontname = Arial,
    fontsize = 16,
    width = 0.55,
    height = 0.38,
    fixedsize = true
  ]

  edge [
    color = '#4B4B4B',
    penwidth = 1.3,
    arrowsize = 0.8
  ]

  subgraph cluster_beschaffung {
    label = ''
    style = filled
    fillcolor = '#F3E7AE'
    color = '#183A43'
    penwidth = 2

    { rank = same; '7A'; '7B'; '7C'; '7D'; '7E' }
    { rank = same; '6A'; '6B'; '6C'; '6D' }
  }

  { rank = same; '5A'; '5B'; '5C' }
  { rank = same; '4A'; '4B'; '4C' }
  { rank = same; '3A'; '3B'; '3C' }
  { rank = same; '2A'; '2B'; '2C'; '2D' }
  { rank = same; '1A'; '1B'; '1C'; '1D'; '1E' }

  '7A' -> '6A'
  '7B' -> '6A'
  '7B' -> '6B'
  '7C' -> '6B'
  '7C' -> '6C'
  '7D' -> '6C'
  '7D' -> '6D'
  '7E' -> '6D'

  '6A' -> '5A'
  '6B' -> '5A'
  '6B' -> '5B'
  '6C' -> '5B'
  '6C' -> '5C'
  '6D' -> '5C'

  '5A' -> '4A'
  '5B' -> '4A'
  '5B' -> '4B'
  '5C' -> '4B'
  '5C' -> '4C'

  '4A' -> '4B'
  '4B' -> '4C'

  '4A' -> '3A'
  '4A' -> '3B'
  '4A' -> '3C'
  '4B' -> '3B'
  '4B' -> '3C'
  '4C' -> '3C'

  '3A' -> '2A'
  '3A' -> '2B'
  '3B' -> '2B'
  '3B' -> '2C'
  '3C' -> '2C'
  '3C' -> '2D'

  '2B' -> '2A'

  '2A' -> '1A'
  '2A' -> '1B'
  '2B' -> '1C'
  '2C' -> '1C'
  '2D' -> '1D'
  '2D' -> '1E'
}
")
```
:::

$\rightarrow$ Select specific locations in the network $\rightarrow$ usually a limited view of the SC




---

## Case study – SushiFresh GmbH {#slide-cs1}
::: slide-subtitle
Initial situation
:::

::: columns
::: {.column width="48%"}

```{r karte+data}
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"
#| fig-height: 6.5

# Stores (demand nodes): six supermarket stores in the Rhine-Main area
werke <- data.frame(
  ID     = paste0("C",1:6),
  Ort    = c("Wiesbaden", "Mainz", "Frankfurt", "Rüsselsheim",
             "Darmstadt", "Offenbach"),
  Lon    = c(8.2398, 8.2711, 8.6821, 8.4116, 8.6512, 8.7761),
  Lat    = c(50.0782, 50.0000, 50.1109, 49.9904, 49.8728, 50.0956),
  Bedarf = c(120, 150, 280, 90, 110, 95)
)



standorte <- data.frame(
  ID       = c("S1","S2","S3", "S4", "S5"),
  Standort = c("Wiesbaden-Erbenheim","Mainz-Hechtsheim","Raunheim","Höchst", "Isenburg"),
  Lon      = c(8.2999, 8.2831, 8.4650, 8.54218, 8.7072),
  Lat      = c(50.0632, 49.9698, 50.0000, 50.11146, 50.0543),
  Kosten   = c(180, 210, 260, 240, 200)/5
)



# Candidate locations for production/distribution (from Chapter 4)
lager_data <- standorte

# Haversine distances (from scm_functions.R), e.g. to the Frankfurt store
lager_data$dist_frankfurt <- mapply(
  haversine,
  lager_data$Lat, lager_data$Lon,
  MoreArgs = list(lat2 = 50.1109, lon2 = 8.6821)
)

# Leaflet map
map <- leaflet(options = leafletOptions(zoomControl = TRUE)) %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  setView(lng = 8.5, lat = 50.02, zoom = 10) %>%

  # Stores (blue)
  addAwesomeMarkers(
    data        = werke,
    lng         = ~Lon,
    lat         = ~Lat,
    label       = ~Ort,
    icon        = awesomeIcons(
      icon        = "shopping-cart",
      library     = "fa",
      markerColor = "blue",
      iconColor   = "white"
    ),
    group = "Stores"
  ) %>%

  # Candidate locations (green)
  addAwesomeMarkers(
    data        = lager_data,
    lng         = ~Lon,
    lat         = ~Lat,
    label       = ~Standort,
    icon        = awesomeIcons(
      icon        = "home",
      library     = "fa",
      markerColor = "green",
      iconColor   = "white"
    ),
    group = "Candidate locations"
  ) %>%

  addLayersControl(
    overlayGroups = c("Stores", "Candidate locations"),
    options       = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomleft",
    colors   = c("#2A7ACA", "#3DB843"),
    labels   = c("Store (demand)", "Candidate location"),
    title    = "Legend"
  )

map
```

:::
::: {.column width="48%"}

::: {.small-text}

::: {.callout-note appearance="simple"}

## Problem statement
**SushiFresh GmbH** supplies six supermarket stores in the Rhine-Main area daily with freshly produced sushi boxes. In Chapter 4, a **qualitative pre-selection of locations using AHP** was carried out (result: Mainz-Hechtsheim narrowly ahead), as well as a **continuous location optimisation** (centre of gravity/Weiszfeld). The question now is: does this result also hold up under a cost-based assessment?
:::

::: {.callout-tip appearance="simple"}

## Review of Chapter 4
- **AHP evaluation:** Raunheim narrowly ahead
- **Centre of gravity / Weiszfeld:** optimal point in the middle of the Rhine-Main area, near Raunheim
- **Open question:** does a cost optimisation (fixed costs + transport) confirm this picture?
:::
:::

:::
:::

---

## Case study – SushiFresh GmbH {#slide-cs2}
::: slide-subtitle
Decision situation
:::

::: {.callout-note appearance="simple"}

## Decision situation

SushiFresh must decide which of the three candidate locations will be developed for production and distribution, and which store should be supplied from which location. The company wants to make this decision so that the annual total costs of fixed costs and transport are minimised. The following information is known:
:::

::: columns
::: {.column width="48%"}

::: small-text
**Candidate locations and annual fixed costs**

```{r lager-tabelle}
#| echo: false
#| warning: false
#| message: false

kbl(standorte,
    col.names = c("ID","Location","Longitude","Latitude",
                  "Fixed costs/year (€1,000)"),
    align = c("c","l","r","r","r")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 20) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

**Transport distances in km**

```{r dist-tabelle}
#| echo: false
#| warning: false
#| message: false

##########################################
# For direct calculation via OSM
# matrix_result <- osrmTable(
#   src = standorte[, c("Lon", "Lat")],
#   dst = werke[, c("Lon", "Lat")],
#   measure = "distance" # Can be "duration", "distance", or c("duration", "distance")
# )
#
# dist_mat <- matrix(matrix_result$distances/1000,
#   nrow = nrow(standorte), byrow = TRUE,
#   dimnames = list(standorte$ID, werke$ID)
# )
##########################################

dist_mat <- matrix(
  c(
    0.12348,	0.41478,	0.47474,	0.55152,	0.85996,	0.18220,
    0.09206,	0.51490,	0.59168,	0.88578,	0.73856,	0.78004,
  	0.48168,	0.27202,	0.17036,	0.54650,	0.44772,	0.12592,
  	0.61176,	0.66708,	0.89392,	0.33502,	0.68526,	0.74242,
  	0.50226,	0.87512,	0.90576,	0.60740,	0.32042,	0.11400),
  nrow = nrow(standorte), byrow = TRUE,dimnames = list(standorte$ID, werke$ID))/.02


df_dist <- as.data.frame(dist_mat)

kbl(df_dist,
    col.names = c(paste0("C",1:6)),
    align = c( rep("r",6)),
    escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 20) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```
:::

:::
::: {.column width="50%"}

::: small-text
**Stores and daily demand**

```{r werke-tabelle}
#| echo: false
#| warning: false
#| message: false



kbl(werke,
    col.names = c("ID","Store",
                  "Longitude","Latitude",
                  "Demand $\\mu$ (boxes/day)"),
    align = c("c","l","l","r","r","r"),
    escape = FALSE,
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 20) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white", width = "1em") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```
:::

:::
:::

<!-- ## Case study - SushiFresh GmbH -->
<!-- ::: slide-subtitle -->
<!-- Task -->
<!-- ::: -->

<!-- ::: {.callout-note appearance="simple"} -->
<!-- ## Objective -->

<!-- You have been commissioned by the management to prepare a decision proposal on how the production and distribution of the sushi boxes should be organised. The proposal should include an assignment of the stores to supply locations. Based on this assignment, the management expects an estimate of the annual total costs that would result from your proposal – and an assessment of whether this confirms the AHP recommendation from Chapter 4 (Raunheim). -->
<!-- ::: -->

<!-- ::: {.fragment .fade-up} -->
<!-- ### Tasks -->

<!-- 1. Formulate an approach for calculating the annual total costs. -->
<!-- 2. Formulate an approach for calculating the transport costs for the distribution of the sushi boxes. -->
<!-- 3. Use the given transport cost rate of €0.02/(box·km). -->
<!-- 4. Find an assignment of the stores to the candidate locations and determine the total costs of your solution. -->
<!-- ::: -->

---

## Warehouse Location Problem {#slide-wlp-1}
::: slide-subtitle
Problem description
:::

::: {.callout-note appearance="minimal" icon=false}
**Problem statement**

Consider a set of **customer locations** and a set of (potential) **warehouse locations**. Each customer location has a **demand for goods** (e.g. in units per month/year) that must be supplied by the warehouses. Transporting the goods incurs distance-dependent **transport costs**. Opening a warehouse incurs **set-up costs**.

The decision is at which locations warehouses should be set up and which customers they should supply.
:::

::: columns
::: {.column width="55%"}

::: {.smaller-text}
::: {.fragment .fade-up}
**Notation**

- Set of warehouses $i \in I$
- Set of customer locations $j \in J$
- Transport links $E \subseteq I \times J$
- Opening costs at warehouse location $i$: $f_i$
- Transport costs (e.g. in €/month) from location $i$ to customer $j$: $c_{ij}$
- Demand of customer $j$: $a_j$
:::
:::
:::

::: {.column width="42%"}
::: {.fragment .fade-up}

```{r wlp-graph}
#| echo: false
#| warning: false
#| message: false
#| out-width: "90%"
#| fig-height: 5

# --- Coordinates (normalised 0..1) ---
lager_nodes <- data.frame(
  id  = c("W1","W2","W3"),
  lbl = c("1","2","3"),
  fi  = c("f<sub>1</sub>", "...", "f<sub>3</sub>"),
  x   = 0.20,
  y   = c(0.15, 0.50, 0.85)
)

kunden_nodes <- data.frame(
  id  = c("A","B","C","D"),
  lbl = c("A","B","C","D"),
  aj  = c("a<sub>A</sub>", "", "...", "a<sub>D</sub>"),
  x   = 0.75,
  y   = c(0.85, 0.60, 0.35, 0.10)
)

lxy <- setNames(lager_nodes[, c("x","y")],   c("x","y")); rownames(lxy) <- lager_nodes$id
kxy <- setNames(kunden_nodes[, c("x","y")],  c("x","y")); rownames(kxy) <- kunden_nodes$id

# Node radius in plot coordinates (adjusted to fig size)
r <- 0.058   # <-- adjust slightly if size changes

edges <- data.frame(
  from = c("W3","W3","W2","W2","W2","W1"),
  to   = c("A", "B", "B", "C", "D", "D"),
  lbl  = c("c<sub>1A</sub>","","","","","c<sub>3D</sub>"),
  stringsAsFactors = FALSE
)

fig <- plot_ly(type = "scatter", mode = "markers")

# Edges: move end points to the node boundary
for (i in seq_len(nrow(edges))) {
  x0 <- lxy[edges$from[i], "x"];  y0 <- lxy[edges$from[i], "y"]
  x1 <- kxy[edges$to[i],   "x"];  y1 <- kxy[edges$to[i],   "y"]

  dx <- x1 - x0;  dy <- y1 - y0
  d  <- sqrt(dx^2 + dy^2)

  # Start point: r away from the warehouse centre towards the customer
  ax <- x0 + r * dx / d
  ay <- y0 + r * dy / d
  # End point:   r away from the customer centre towards the warehouse
  ex <- x1 - r * dx / d
  ey <- y1 - r * dy / d

  fig <- fig %>% add_annotations(
    x = ex, y = ey, ax = ax, ay = ay,
    xref = "x", yref = "y", axref = "x", ayref = "y",
    showarrow = TRUE, arrowhead = 2, arrowsize = 1,
    arrowwidth = 1.5, arrowcolor = "#4B4B4B", text = ""
  )

  # Edge label centred
  if (nchar(edges$lbl[i]) > 0) {
    fig <- fig %>% add_annotations(
      x = (ax + ex) / 2 + 0.02, y = (ay + ey) / 2 + 0.025,
      xref = "x", yref = "y",
      text = edges$lbl[i], showarrow = FALSE,
      font = list(size = 13, color = "#1F2933", family = "Arial")
    )
  }
}

# Warehouse nodes (hexagon)
fig <- fig %>% add_trace(
  x = lager_nodes$x, y = lager_nodes$y, text = lager_nodes$lbl,
  mode = "markers+text",
  marker = list(symbol = "hexagon", size = 40,
                color = "#BFDBF7", line = list(color = "#4B5563", width = 1.5)),
  textfont = list(size = 16, color = "#1F2933", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# f_i on the left
fig <- fig %>% add_annotations(
  x = lager_nodes$x - 0.10, y = lager_nodes$y,
  xref = "x", yref = "y",
  text = lager_nodes$fi, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "right"
)

# Customer nodes (circle)
fig <- fig %>% add_trace(
  x = kunden_nodes$x, y = kunden_nodes$y, text = kunden_nodes$lbl,
  mode = "markers+text",
  marker = list(symbol = "circle", size = 40,
                color = "#2D9E8E", line = list(color = "#2F3E46", width = 1.5)),
  textfont = list(size = 16, color = "white", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# a_j on the right
fig <- fig %>% add_annotations(
  x = kunden_nodes$x + 0.08, y = kunden_nodes$y,
  xref = "x", yref = "y",
  text = kunden_nodes$aj, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "left"
)

# Set labels
fig <- fig %>% add_annotations(
  x = c(0.20, 0.75), y = c(1.00, 1.00),
  xref = "x", yref = "y",
  text = c("<b>Set <i>I</i></b>", "<b>Set <i>J</i></b>"),
  showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial")
)

fig <- fig %>% layout(
  xaxis = list(visible = FALSE, range = c(0, 1), fixedrange = TRUE),
  yaxis = list(visible = FALSE, range = c(-0.05, 1.08), fixedrange = TRUE),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 40, r = 40, t = 10, b = 10),
  height = 380,
  width = 650
)

fig
```
:::
:::
:::


---

## Warehouse Location Problem {#slide-wlp-model}
::: slide-subtitle
Mathematical model
:::

::: columns
::: {.column width="55%"}

::: {.callout-tip appearance="simple" icon=false}

#### Mixed-integer linear programming model
Minimise the total costs of transport and opening locations

$$\min \rightarrow \sum_{i \in I} y_i \cdot f_i + \sum_{(i,j) \in E} x_{ij} \cdot c_{ij}$$
:::

::: {.smaller-text}

**Constraints**
$$
\begin{align}
  \sum_{(i,j) \in E} x_{ij} &= a_j          & \forall j \in J      &\quad\rightarrow \text{satisfy demand} \\
  x_{ij}                    &\leq y_i \cdot a_j & \forall (i,j) \in E &\quad\rightarrow \text{transport only from opened locations} \\
  x_{ij}                    &\geq 0          & \forall (i,j) \in E &\quad\rightarrow \text{transport quantities non-negative} \\
  y_i                       &\in \{0,1\}     & \forall i \in I     &\quad\rightarrow \text{location is opened or not}
\end{align}
$$
:::

:::
::: {.column width="43%"}

```{r wlp-graph2}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 5

# --- Coordinates (normalised 0..1) ---
lager <- data.frame(
  id  = c("W1","W2","W3"),
  lbl = c("1","2","3"),
  fi  = c("f<sub>1</sub>", "...", "f<sub>3</sub>"),
  x   = 0.20,
  y   = c(0.15, 0.50, 0.85)
)

kunden <- data.frame(
  id  = c("A","B","C","D"),
  lbl = c("A","B","C","D"),
  aj  = c("a<sub>A</sub>", "", "...", "a<sub>D</sub>"),
  x   = 0.75,
  y   = c(0.85, 0.60, 0.35, 0.10)
)

lxy <- setNames(lager[, c("x","y")],   c("x","y")); rownames(lxy) <- lager$id
kxy <- setNames(kunden[, c("x","y")],  c("x","y")); rownames(kxy) <- kunden$id

# Node radius in plot coordinates (adjusted to fig size)
r <- 0.058   # <-- adjust slightly if size changes

edges <- data.frame(
  from = c("W3","W3","W2","W2","W2","W1"),
  to   = c("A", "B", "B", "C", "D", "D"),
  lbl  = c("c<sub>1A</sub>","","","","","c<sub>3D</sub>"),
  stringsAsFactors = FALSE
)

fig <- plot_ly(type = "scatter", mode = "markers")

# Edges: move end points to the node boundary
for (i in seq_len(nrow(edges))) {
  x0 <- lxy[edges$from[i], "x"];  y0 <- lxy[edges$from[i], "y"]
  x1 <- kxy[edges$to[i],   "x"];  y1 <- kxy[edges$to[i],   "y"]

  dx <- x1 - x0;  dy <- y1 - y0
  d  <- sqrt(dx^2 + dy^2)

  # Start point: r away from the warehouse centre towards the customer
  ax <- x0 + r * dx / d
  ay <- y0 + r * dy / d
  # End point:   r away from the customer centre towards the warehouse
  ex <- x1 - r * dx / d
  ey <- y1 - r * dy / d

  fig <- fig %>% add_annotations(
    x = ex, y = ey, ax = ax, ay = ay,
    xref = "x", yref = "y", axref = "x", ayref = "y",
    showarrow = TRUE, arrowhead = 2, arrowsize = 1,
    arrowwidth = 1.5, arrowcolor = "#4B4B4B", text = ""
  )

  # Edge label centred
  if (nchar(edges$lbl[i]) > 0) {
    fig <- fig %>% add_annotations(
      x = (ax + ex) / 2 + 0.02, y = (ay + ey) / 2 + 0.025,
      xref = "x", yref = "y",
      text = edges$lbl[i], showarrow = FALSE,
      font = list(size = 13, color = "#1F2933", family = "Arial")
    )
  }
}

# Warehouse nodes (hexagon)
fig <- fig %>% add_trace(
  x = lager$x, y = lager$y, text = lager$lbl,
  mode = "markers+text",
  marker = list(symbol = "hexagon", size = 40,
                color = "#BFDBF7", line = list(color = "#4B5563", width = 1.5)),
  textfont = list(size = 16, color = "#1F2933", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# f_i on the left
fig <- fig %>% add_annotations(
  x = lager$x - 0.10, y = lager$y,
  xref = "x", yref = "y",
  text = lager$fi, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "right"
)

# Customer nodes (circle)
fig <- fig %>% add_trace(
  x = kunden$x, y = kunden$y, text = kunden$lbl,
  mode = "markers+text",
  marker = list(symbol = "circle", size = 40,
                color = "#2D9E8E", line = list(color = "#2F3E46", width = 1.5)),
  textfont = list(size = 16, color = "white", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# a_j on the right
fig <- fig %>% add_annotations(
  x = kunden$x + 0.08, y = kunden$y,
  xref = "x", yref = "y",
  text = kunden$aj, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "left"
)

# Set labels
fig <- fig %>% add_annotations(
  x = c(0.20, 0.75), y = c(1.00, 1.00),
  xref = "x", yref = "y",
  text = c("<b>Set <i>I</i></b>", "<b>Set <i>J</i></b>"),
  showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial")
)

fig <- fig %>% layout(
  xaxis = list(visible = FALSE, range = c(0, 1), fixedrange = TRUE),
  yaxis = list(visible = FALSE, range = c(-0.05, 1.08), fixedrange = TRUE),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 40, r = 40, t = 10, b = 10),
  height = 380,
  width = 650
)

fig
```

:::
:::

---

## Warehouse Location Problem {#slide-wlp-example}
::: slide-subtitle
Case example
:::

```{r wlp-example-data}
#| echo: false
#| warning: false
#| message: false

# Problem data: 3 candidate locations, 6 stores (SushiFresh GmbH)
wlp_lager <- data.frame(
  ID      = standorte$ID,
  fi      = standorte$Kosten*1000,   # Fixed costs (EUR/year)
  stringsAsFactors = FALSE
)

wlp_kunden <- data.frame(
  ID  = werke$ID,
  aj  = werke$Bedarf*1000,   # Demand (boxes/year)
  stringsAsFactors = FALSE
)

# Transport costs cij (EUR per box, from distance x 0.02 EUR/(box*km))
wlp_cij <-df_dist*.02
```

::: columns
::: {.column width="38%"}

::: small-text
**Locations – fixed costs $f_i$ (€/year)**

```{r}
#| echo: false
kbl(wlp_lager,
    col.names = c("Location i", "f_i (€)"),
    align = c("c","r"), escape = FALSE,
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

**Stores – demand $a_j$ (boxes/year)**

```{r}
#| echo: false
kbl(wlp_kunden,
    col.names = c("Store j", "a_j"),
    align = c("c","r"), escape = FALSE,
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2D9E8E", color = "white") |>
  row_spec(0, background = "#2D9E8E", color = "white", bold = TRUE)
```
:::

:::
::: {.column width="59%"}

::: small-text
**Transport costs $c_{ij}$ (€/box)**

```{r}
#| echo: false
kbl(round(as.data.frame(wlp_cij), 3),
    align = rep("r", 6), escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

:::

:::
:::

---

## Warehouse Location Problem {#slide-wlp-ompr}
::: slide-subtitle
Implementation with R/OMPR
:::

::: columns
::: {.column width="55%"}

```{r wlp-ompr-code}
#| echo: true
#| eval: true
#| warning: false
#| message: false
#| code-line-numbers: true

library(ompr)
library(ompr.roi)
library(ROI.plugin.glpk)

# Problem data
n_I <- 5; n_J <- 6
f   <- c(36000, 42000, 52000, 48000, 40000)
a   <- c(120000, 150000, 280000, 90000, 110000, 95000)
c_ij <- matrix(
  c(
    0.12348,	0.41478,	0.47474,	0.55152,	0.85996,	0.18220,
    0.09206,	0.51490,	0.59168,	0.88578,	0.73856,	0.78004,
  	0.48168,	0.27202,	0.17036,	0.54650,	0.44772,	0.12592,
  	0.61176,	0.66708,	0.89392,	0.33502,	0.68526,	0.74242,
  	0.50226,	0.87512,	0.90576,	0.60740,	0.32042,	0.11400),
  nrow = n_I, byrow = TRUE)

# MILP model
model <- MIPModel() |>
  # Variables
  add_variable(y[i], i = 1:n_I, type = "binary") |>
  add_variable(x[i,j], i = 1:n_I, j = 1:n_J,
               type = "continuous", lb = 0) |>
  # Objective function
  set_objective(
    sum_expr(f[i] * y[i], i = 1:n_I) +
    sum_expr(c_ij[i,j] * x[i,j],
             i = 1:n_I, j = 1:n_J),
    sense = "min") |>
  # C1: satisfy demand completely
  add_constraint(
    sum_expr(x[i,j], i = 1:n_I) == a[j],
    j = 1:n_J) |>
  # C2: transport only from opened warehouses
  add_constraint(
    x[i,j] <= y[i] * a[j],
    i = 1:n_I, j = 1:n_J)

result <- solve_model(model,
           with_ROI(solver = "glpk", verbose = FALSE))
```

:::
::: {.column width="42%"}

::: {.callout-note appearance="simple"}

## OMPR – workflow

1. `MIPModel()` → empty model
2. `add_variable()` → decision variables
3. `set_objective()` → objective function
4. `add_constraint()` → constraints
5. `solve_model()` → call solver
6. `get_solution()` → extract solution
:::

```{r wlp-ompr-result}
#| echo: true
#| eval: true
#| warning: false
#| message: false

# Extract optimal solution
y_sol <- get_solution(result, y[i])
x_sol <- get_solution(result, x[i,j])

# Opened warehouses
offene_lager <- y_sol$i[y_sol$value > 0.5]

# Optimal objective value
obj_val <- objective_value(result)
```


:::
:::

---

## Warehouse Location Problem  {#slide-wlp-solution}
::: slide-subtitle
Case example – optimal solution
:::

::: columns
::: {.column width="48%"}

```{r wlp-solution-table}
#| echo: false
#| warning: false
#| message: false

# Prepare assignment table
zuordnung <- x_sol |>
  dplyr::filter(value > 0.001) |>
  dplyr::mutate(
    Lager  = paste0("S", i),
    Kunde  = paste0("C", j),
    Menge  = value,
    Kosten = round(c_ij[cbind(i, j)] * value, 1)
  ) |>
  dplyr::select(Lager, Kunde, Menge, Kosten)

kbl(zuordnung,
    col.names = c("Location $i$","Store $j$",
                  "Quantity $x_{ij}$ (boxes)","Transport costs (€)"),
    align = c("c","c","r","r"), escape = F,
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  column_spec(2, bold = TRUE, background = "#2D9E8E", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

```{r wlp-cost-breakdown}
#| echo: false
#| warning: false
#| message: false

eroeffnung  <- sum(f[offene_lager])
transport   <- sum(zuordnung$Kosten)

kosten_df <- data.frame(
  Kostenart = c("Fixed costs", "Transport costs", "Total costs Z*"),
  Wert      = c(eroeffnung, transport, round(obj_val, 1))
)

kbl(kosten_df,
    col.names = c("Cost type", "€/year"),
    align = c("l","r"),
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed"),
                full_width = TRUE, font_size = 24) |>
  row_spec(3, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

:::
::: {.column width="50%"}

```{r wlp-solution-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 5.5

lager_pos <- data.frame(
  id = paste0("S", 1:nrow(standorte)),
  x  = 0.18,
  y  =  seq(0.04 , 0.9, length.out = nrow(standorte)), #c(0.15, 0.50, 0.85),
  open = (1:nrow(standorte)) %in% offene_lager
)

kunden_pos <- data.frame(
  id = paste0("C", 1:6),
  x  = 0.80,
  y  = c(0.92, 0.74, 0.56, 0.38, 0.20, 0.04)
)

lxy2 <- setNames(lager_pos[, c("x","y")], c("x","y"))
rownames(lxy2) <- lager_pos$id
kxy2 <- setNames(kunden_pos[, c("x","y")], c("x","y"))
rownames(kxy2) <- kunden_pos$id

r2 <- 0.048

fig2 <- plot_ly(type = "scatter", mode = "markers")

# Active edges
active_edges <- x_sol |> dplyr::filter(value > 0.001) |>
  dplyr::mutate(from = paste0("S", i), to = paste0("C", j))

# All possible edges (thin, grey)
for (ii in 1:nrow(standorte)) for (jj in 1:nrow(werke)) {
  wi <- paste0("S", ii); cj <- paste0("C", jj)
  x0 <- lxy2[wi,"x"]; y0 <- lxy2[wi,"y"]
  x1 <- kxy2[cj,"x"]; y1 <- kxy2[cj,"y"]
  dx <- x1-x0; dy <- y1-y0; dd <- sqrt(dx^2+dy^2)
  ax <- x0 + r2*dx/dd; ay <- y0 + r2*dy/dd
  ex <- x1 - r2*dx/dd; ey <- y1 - r2*dy/dd
  fig2 <- fig2 %>% add_annotations(
    x=ex,y=ey,ax=ax,ay=ay,
    xref="x",yref="y",axref="x",ayref="y",
    showarrow=TRUE,arrowhead=0,arrowsize=0.8,
    arrowwidth=0.8,arrowcolor="#D0D5DD",text="")
}

# Active edges (bold, coloured)
for (k in seq_len(nrow(active_edges))) {
  wi <- active_edges$from[k]; cj <- active_edges$to[k]
  x0 <- lxy2[wi,"x"]; y0 <- lxy2[wi,"y"]
  x1 <- kxy2[cj,"x"]; y1 <- kxy2[cj,"y"]
  dx <- x1-x0; dy <- y1-y0; dd <- sqrt(dx^2+dy^2)
  ax <- x0 + r2*dx/dd; ay <- y0 + r2*dy/dd
  ex <- x1 - r2*dx/dd; ey <- y1 - r2*dy/dd
  fig2 <- fig2 %>% add_annotations(
    x=ex,y=ey,ax=ax,ay=ay,
    xref="x",yref="y",axref="x",ayref="y",
    showarrow=TRUE,arrowhead=2,arrowsize=1,
    arrowwidth=2.2,arrowcolor="#01696f",text="")
  mx <- (ax+ex)/2 + 0.025; my <- (ay+ey)/2 + 0.03
  fig2 <- fig2 %>% add_annotations(
    x=mx,y=my,xref="x",yref="y",
    text=paste0("x=",active_edges$value[k]),
    showarrow=FALSE,
    font=list(size=11,color="#01696f",family="Arial"))
}

# Warehouse points
lager_col <- ifelse(lager_pos$open, "#BFDBF7", "#E8E8E8")
lager_brd <- ifelse(lager_pos$open, "#4B5563", "#B0B0B0")
fig2 <- fig2 %>% add_trace(
  x=lager_pos$x, y=lager_pos$y, text=lager_pos$id,
  mode="markers+text",
  marker=list(symbol="hexagon",size=38,
              color=lager_col,
              line=list(color=lager_brd,width=2)),
  textfont=list(size=14,color="#1F2933",family="Arial"),
  textposition="middle center",
  hoverinfo="skip",showlegend=FALSE)

# Opening label
fig2 <- fig2 %>% add_annotations(
  x=lager_pos$x - 0.09, y=lager_pos$y,
  xref="x",yref="y",
  text=ifelse(lager_pos$open,
              paste0("f=",f[1:3],"✓"),
              paste0("f=",f[1:3])),
  showarrow=FALSE,
  font=list(size=11,
            color=ifelse(lager_pos$open,"#01696f","#9CA3AF"),
            family="Arial"),
  xanchor="right")

# Customer points
fig2 <- fig2 %>% add_trace(
  x=kunden_pos$x, y=kunden_pos$y, text=kunden_pos$id,
  mode="markers+text",
  marker=list(symbol="circle",size=38,
              color="#2D9E8E",line=list(color="#2F3E46",width=1.5)),
  textfont=list(size=14,color="white",family="Arial"),
  textposition="middle center",
  hoverinfo="skip",showlegend=FALSE)

fig2 <- fig2 %>% add_annotations(
  x=kunden_pos$x+0.07, y=kunden_pos$y,
  xref="x",yref="y",
  text=paste0("a=",a),
  showarrow=FALSE,
  font=list(size=11,color="#1F2933",family="Arial"),
  xanchor="left")

fig2 <- fig2 %>% layout(
  xaxis=list(visible=FALSE,range=c(0,1),fixedrange=TRUE),
  yaxis=list(visible=FALSE,range=c(-0.05,1.05),fixedrange=TRUE),
  plot_bgcolor="transparent",
  paper_bgcolor="transparent",
  margin=list(l=40,r=20,t=10,b=10),
  height=420,
  width = 700)

fig2
```

:::
:::

---

## Warehouse Location Problem {#slide-heuristics-intro}
::: slide-subtitle
Heuristic solution methods – basic ideas
:::

::: columns
::: {.column width="55%"}

::: {.callout-warning appearance="simple"}

## Why heuristics?

The WLP is **NP-hard**: exact solvers (MILP) do find optimal solutions, but the computing time grows exponentially with $|I|$ and $|J|$. For large real-world instances, **constructive heuristics** are often the only practicable approach.
:::

:::: {.fragment .fade-up}
::: {.small-text}
**Basic idea**

- start with a feasible solution and try to improve it step by step
- Each step: **greedy** – choose the change with the largest cost reduction
- Stop when no further improvement is possible
- Customers are always assigned to the **cheapest open warehouse**
:::
::::

:::
::: {.column width="42%"}

::: {.fragment .fade-up}
::: {.smaller-text}
**Two classic approaches**

| Method | Starting point | Direction |
|---|---|---|
| **Add heuristic** | No warehouses open | Add step by step |
| **Drop heuristic** | All warehouses open | Close step by step |
:::
:::

::: {.fragment .fade-up}
::: {.smaller-text}
**Basic ideas of add/drop – graphical**
:::
```{r heuristic-overview-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 5

# Cost curve add vs. drop progression (actual values from the SushiFresh case example,
# k€/year)
fig_h <- plot_ly()

# Add curve: starts empty, opens S1 (Wiesbaden-Erbenheim) directly -> optimum
fig_h <- fig_h %>% add_trace(
  x = c(1, 2), y = c(269, 262),
  type = "scatter", mode = "lines+markers",
  name = "Add heuristic",
  line   = list(color = "#01696f", width = 2.5),
  marker = list(color = "#01696f", size = 10)
)

# Drop curve: starts with all 3 locations open, closes Raunheim, then
# Mainz-Hechtsheim -> same optimum
fig_h <- fig_h %>% add_trace(
  x = c(5,4,3, 2), y = c(388, 340 ,300, 262),
  type = "scatter", mode = "lines+markers",
  name = "Drop heuristic",
  line   = list(color = "#da7101", width = 2.5, dash = "dash"),
  marker = list(color = "#da7101", size = 10)
)

# Optimum
fig_h <- fig_h %>% add_trace(
  x = c(1), y = c(262),
  type = "scatter", mode = "lines",
  name = "Optimum (Z* = 262)",
  line = list(color = "#a12c7b", width = 1.5, dash = "dot")
)

fig_h <- fig_h %>% layout(
  xaxis = list(title = "Number of open locations", dtick = 1,
               range = c(0.5, 5.5)),
  yaxis = list(title = "Total costs (k€/year)",
               range = c(250, 760)),
  legend = list(orientation = "h", y = -0.2),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 20, t = 10, b = 60),
  height = 340
)

fig_h
```
:::

:::
:::

---

## Warehouse Location Problem {#slide-add-heuristic}
::: slide-subtitle
Add heuristic
:::

::: columns
::: {.column width="52%"}

::: {.callout-note appearance="minimal" icon=false}
**Idea:** Start with no open warehouse. In each step, add the warehouse $i^*$ that **reduces** the total costs the most.
:::

::: {.fragment .fade-up}

**Algorithm**

::: {.smaller-text}

**Initialisation:** $S \leftarrow \emptyset$

**Repeat:**

1. For each warehouse not yet opened $i \notin S$, calculate the cost contribution:
$$\Delta_i = Z(S) - Z(S \cup \{i\})$$

2. Choose $i^* = \arg\max_{i \notin S} \Delta_i$

3. **If** $\Delta_{i^*} > 0$: open $i^*$, set $S \leftarrow S \cup \{i^*\}$

4. **Otherwise:** stop

**Assignment rule:** Each customer $j$ is assigned to the cheapest open warehouse:
$\sigma(j) = \arg\min_{i \in S} c_{ij}$
:::
:::

:::
::: {.column width="46%"}

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## Cost calculation

$$Z(S) = \underbrace{\sum_{i \in S} f_i}_{\text{Opening}} + \underbrace{\sum_{j \in J} \min_{i \in S} c_{ij} \cdot a_j}_{\text{Transport}}$$

- **Properties:**
  - Greedy, not guaranteed to be optimal
  - Transport costs decrease from step to step
  - A new warehouse is opened if opening costs < transport cost savings
:::
:::

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}

## Stopping condition

The heuristic stops as soon as **no** further warehouse reduces the total costs.
:::
:::

:::
:::

---

## Warehouse Location Problem {#slide-drop-heuristic}
::: slide-subtitle
Drop heuristic
:::

::: columns
::: {.column width="52%"}

::: {.callout-note appearance="minimal" icon=false}
**Idea:** Start with **all** warehouses open. In each step, close the warehouse $i^*$ whose closure **reduces** the total costs the most.
:::

::: {.fragment .fade-up}
**Algorithm**

::: {.smaller-text}
**Initialisation:** $S \leftarrow I$

**Repeat:**

1. For each warehouse still open $i \in S$, calculate the cost contribution:
$$\Delta_i = Z(S) - Z(S \setminus \{i\})$$

2. Choose $i^* = \arg\max_{i \in S} \Delta_i$

3. **If** $\Delta_{i^*} > 0$: close $i^*$, set $S \leftarrow S \setminus \{i^*\}$

4. **Otherwise:** stop

**Assignment rule:** $\sigma(j) = \arg\min_{i \in S} c_{ij}$
:::
:::
:::
::: {.column width="46%"}

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## Comparison with add heuristic

| Feature | Add | Drop |
|---|---|---|
| **Start** | $S = \emptyset$ | $S = I$ |
| **Direction** | constructive | destructive |
| **Starting costs** | $\infty$ (no warehouse) | $\sum_i f_i + \ldots$ |
| **Running time** | $\mathcal{O}(|I|^2 \cdot |J|)$ | $\mathcal{O}(|I|^2 \cdot |J|)$ |
| **Tendency** | few warehouses | more warehouses |

:::
:::

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}

## Important

Neither heuristic is **guaranteed to be optimal**. They deliver feasible solutions; the solution quality depends heavily on the cost structure of the instance.
:::
:::

:::
:::

---

## Warehouse Location Problem {#slide-add-example}
::: slide-subtitle
Add heuristic – case example
:::

```{r add-heuristic-compute}
#| echo: false
#| warning: false
#| message: false

# Problem data (identical to slide-wlp-example)
n_I  <- nrow(standorte); n_J <- nrow(werke)
f_h  <- standorte$Kosten*1000 # c(180000, 210000, 260000)
a_h  <- werke$Bedarf*1000 # c(43800, 54750, 102200, 32850, 40150, 34675)
c_h  <- c_ij
#   matrix(
#   c(0.092044, 0.146448, 0.555583, 0.227330, 0.657125, 0.683408,
#     0.248882, 0.069319, 0.650567, 0.189390, 0.569494, 0.757793,
#     0.365663, 0.277179, 0.396128, 0.079272, 0.388659, 0.492523),
#   nrow = n_I, byrow = TRUE,
#   dimnames = list(paste0("S", 1:n_I), paste0("C", 1:n_J))
# )

total_cost_h <- function(S) {
  if (length(S) == 0) return(Inf)
  sum(f_h[S]) + sum(sapply(1:n_J, function(j) min(c_h[S, j]) * a_h[j]))
}

best_assign <- function(S) {
  sapply(1:n_J, function(j) S[which.min(c_h[S, j])])
}

# Run add heuristic
S_add_h  <- integer(0)
rem_h    <- 1:n_I
add_log  <- list()

repeat {
  cur_cost  <- total_cost_h(S_add_h)
  cand_cost <- sapply(rem_h, function(i) total_cost_h(c(S_add_h, i)))
  gains     <- cur_cost - cand_cost
  # Selection based on the resulting total costs (robust even if
  # cur_cost = Inf, i.e. in the very first step: there all
  # differences to Inf are equal to "Inf" and which.max(gains) would only
  # be a coincidence of the order in rem_h)
  best_idx  <- which.min(cand_cost)
  best_gain <- gains[best_idx]
  best_i    <- rem_h[best_idx]
  if (is.finite(best_gain) && best_gain <= 0) break
  S_add_h  <- c(S_add_h, best_i)
  rem_h    <- setdiff(rem_h, best_i)
  asgn     <- best_assign(S_add_h)
  fix_cost <- sum(f_h[S_add_h])
  tr_cost  <- total_cost_h(S_add_h) - fix_cost
  add_log[[length(add_log)+1]] <- list(
    step      = length(add_log)+1,
    added     = best_i,
    open      = S_add_h,
    gain      = best_gain,
    fix       = fix_cost,
    transp    = tr_cost,
    total     = total_cost_h(S_add_h),
    assign    = asgn
  )
}
```

::: columns
::: {.column width="52%"}

::: small-text

```{r add-steps-table}
#| echo: false
#| warning: false
#| message: false

add_df <- do.call(rbind, lapply(add_log, function(s) {
  data.frame(
    Schritt    = s$step,
    Aktion     = paste0("+S", s$added),
    Offen      = paste0("{", paste0("S", sort(s$open), collapse=", "), "}"),
    Fix        = s$fix,
    Transport  = round(s$transp, 1),
    Gesamt     = round(s$total, 1),
    Zuordnung  = paste0(paste0("C", 1:n_J, "→S", s$assign), collapse=", ")
  )
}))

kbl(add_df[, c("Schritt","Aktion","Offen","Fix","Transport","Gesamt")],
    col.names = c("Step","Action","Open locations",
                  "Fixed costs (€)","Transport (€)","Total (€)"),
    align = c("c","c","l","r","r","r"),
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  column_spec(2, bold = TRUE, color = "#01696f") |>
  column_spec(6, bold = TRUE) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

**Store assignment per step:**

```{r add-assign-table}
#| echo: false
#| warning: false
#| message: false

assign_df <- do.call(rbind, lapply(add_log, function(s) {
  row <- c(Schritt = s$step,
           setNames(paste0("S", s$assign), paste0("C", 1:n_J)))
  as.data.frame(t(row), stringsAsFactors = FALSE)
}))

kbl(assign_df,
    col.names = c("Step", paste0("C", 1:n_J)),
    align = c("c", rep("c", n_J))) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```

::: {.callout-tip appearance="simple"}
**Add heuristic result:** location $\{`r paste0("S", sort(S_add_h), collapse=", ")`\}$ opened, $Z_\text{Add} = `r format(round(total_cost_h(S_add_h), 1), big.mark=" ")`$ €/year
:::
:::

:::
::: {.column width="46%"}

```{r add-solution-plot}
#| cache: false
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 5.5

final_open_add <- S_add_h
final_asgn_add <- best_assign(final_open_add)

lp_add <- data.frame(
  id = paste0("S", 1:nrow(standorte)),
  x  = 0.18,
  y  =  seq(0.04 , 0.9, length.out = nrow(standorte)), #c(0.15, 0.50, 0.85),
  open = seq_len(n_I) %in% final_open_add
)
kp_add <- data.frame(
  id = paste0("C", 1:n_J),
  x  = 0.80,
  y  = c(0.92, 0.74, 0.56, 0.38, 0.20, 0.04)
)
lxy_a <- setNames(lp_add[, c("x","y")], c("x","y")); rownames(lxy_a) <- lp_add$id
kxy_a <- setNames(kp_add[, c("x","y")], c("x","y")); rownames(kxy_a) <- kp_add$id
r_a <- 0.048

fig_a <- plot_ly(type = "scatter", mode = "markers")

for (ii in 1:n_I) for (jj in 1:n_J) {
  wi <- paste0("S", ii); cj <- paste0("C", jj)
  x0 <- lxy_a[wi,"x"]; y0 <- lxy_a[wi,"y"]
  x1 <- kxy_a[cj,"x"]; y1 <- kxy_a[cj,"y"]
  dx <- x1-x0; dy <- y1-y0; dd <- sqrt(dx^2+dy^2)
  ax <- x0+r_a*dx/dd; ay <- y0+r_a*dy/dd
  ex <- x1-r_a*dx/dd; ey <- y1-r_a*dy/dd
  fig_a <- fig_a %>% add_annotations(
    x=ex,y=ey,ax=ax,ay=ay,xref="x",yref="y",axref="x",ayref="y",
    showarrow=TRUE,arrowhead=0,arrowsize=0.8,arrowwidth=0.8,
    arrowcolor="#D0D5DD",text="")
}

for (jj in 1:n_J) {
  wi <- paste0("S", final_asgn_add[jj]); cj <- paste0("C", jj)
  x0 <- lxy_a[wi,"x"]; y0 <- lxy_a[wi,"y"]
  x1 <- kxy_a[cj,"x"]; y1 <- kxy_a[cj,"y"]
  dx <- x1-x0; dy <- y1-y0; dd <- sqrt(dx^2+dy^2)
  ax <- x0+r_a*dx/dd; ay <- y0+r_a*dy/dd
  ex <- x1-r_a*dx/dd; ey <- y1-r_a*dy/dd
  fig_a <- fig_a %>% add_annotations(
    x=ex,y=ey,ax=ax,ay=ay,xref="x",yref="y",axref="x",ayref="y",
    showarrow=TRUE,arrowhead=2,arrowsize=1,arrowwidth=2.2,
    arrowcolor="#01696f",text="")
}

lc <- ifelse(lp_add$open, "#BFDBF7", "#E8E8E8")
lb <- ifelse(lp_add$open, "#4B5563", "#B0B0B0")
fig_a <- fig_a %>% add_trace(
  x=lp_add$x, y=lp_add$y, text=lp_add$id,
  mode="markers+text",
  marker=list(symbol="hexagon",size=38,color=lc,line=list(color=lb,width=2)),
  textfont=list(size=14,color="#1F2933",family="Arial"),
  textposition="middle center",hoverinfo="skip",showlegend=FALSE)
fig_a <- fig_a %>% add_annotations(
  x=lp_add$x-0.09, y=lp_add$y, xref="x",yref="y",
  text=ifelse(lp_add$open, paste0("f=",round(f_h/1000),"k€ ✓"), paste0("f=",round(f_h/1000),"k€")),
  showarrow=FALSE,
  font=list(size=11,color=ifelse(lp_add$open,"#01696f","#9CA3AF"),family="Arial"),
  xanchor="right")
fig_a <- fig_a %>% add_trace(
  x=kp_add$x, y=kp_add$y, text=kp_add$id,
  mode="markers+text",
  marker=list(symbol="circle",size=38,color="#2D9E8E",
              line=list(color="#2F3E46",width=1.5)),
  textfont=list(size=14,color="white",family="Arial"),
  textposition="middle center",hoverinfo="skip",showlegend=FALSE)
fig_a <- fig_a %>% add_annotations(
  x=kp_add$x+0.07, y=kp_add$y, xref="x",yref="y",
  text=paste0("μ=",c(120,150,280,90,110,95)),
  showarrow=FALSE,font=list(size=11,color="#1F2933",family="Arial"),
  xanchor="left")
fig_a <- fig_a %>% layout(
  xaxis=list(visible=FALSE,range=c(0,1),fixedrange=TRUE),
  yaxis=list(visible=FALSE,range=c(-0.05,1.05),fixedrange=TRUE),
  plot_bgcolor="transparent",paper_bgcolor="transparent",
  margin=list(l=40,r=20,t=10,b=10),height=420, width = 650)
fig_a
```

:::
:::

---

## Warehouse Location Problem {#slide-drop-example}
::: slide-subtitle
Drop heuristic – case example
:::

```{r drop-heuristic-compute}
#| echo: false
#| warning: false
#| message: false

S_drop_h  <- 1:n_I
drop_log  <- list()

# Starting row
fix0   <- sum(f_h[S_drop_h])
tr0    <- total_cost_h(S_drop_h) - fix0
drop_log[[1]] <- list(
  step    = 0,
  removed = NA,
  open    = S_drop_h,
  gain    = NA,
  fix     = fix0,
  transp  = tr0,
  total   = total_cost_h(S_drop_h),
  assign  = best_assign(S_drop_h)
)

repeat {
  if (length(S_drop_h) <= 1) break
  cur_cost <- total_cost_h(S_drop_h)
  gains <- sapply(S_drop_h, function(i)
    cur_cost - total_cost_h(setdiff(S_drop_h, i)))
  best_idx  <- which.max(gains)
  best_gain <- gains[best_idx]
  best_i    <- S_drop_h[best_idx]
  if (best_gain <= 0) break
  S_drop_h <- setdiff(S_drop_h, best_i)
  fix_cost <- sum(f_h[S_drop_h])
  tr_cost  <- total_cost_h(S_drop_h) - fix_cost
  asgn     <- best_assign(S_drop_h)
  drop_log[[length(drop_log)+1]] <- list(
    step    = length(drop_log),
    removed = best_i,
    open    = S_drop_h,
    gain    = best_gain,
    fix     = fix_cost,
    transp  = tr_cost,
    total   = total_cost_h(S_drop_h),
    assign  = asgn
  )
}
```

::: columns
::: {.column width="52%"}

::: small-text

```{r drop-steps-table}
#| echo: false
#| warning: false
#| message: false

drop_df <- do.call(rbind, lapply(drop_log, function(s) {
  data.frame(
    Schritt   = s$step,
    Aktion    = ifelse(is.na(s$removed), "Start",
                       paste0("-S", s$removed)),
    Offen     = paste0("{", paste0("S", sort(s$open), collapse=", "), "}"),
    Fix       = s$fix,
    Transport = round(s$transp, 1),
    Gesamt    = round(s$total, 1)
  )
}))

kbl(drop_df,
    col.names = c("Step","Action","Open locations",
                  "Fixed costs (€)","Transport (€)","Total (€)"),
    align = c("c","c","l","r","r","r"),
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  column_spec(2, bold = TRUE, color = "#da7101") |>
  column_spec(6, bold = TRUE) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(1, italic = TRUE, color = "#6B7280")
```

**Store assignment per step:**

```{r drop-assign-table}
#| echo: false
#| warning: false
#| message: false

assign_drop_df <- do.call(rbind, lapply(drop_log, function(s) {
  row <- c(Schritt = s$step,
           setNames(paste0("S", s$assign), paste0("C", 1:n_J)))
  as.data.frame(t(row), stringsAsFactors = FALSE)
}))

kbl(assign_drop_df,
    col.names = c("Step", paste0("C", 1:n_J)),
    align = c("c", rep("c", n_J))) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(1, italic = TRUE, color = "#6B7280")
```

::: {.callout-tip appearance="simple"}
**Drop heuristic result:** location $\{`r paste0("S", sort(S_drop_h), collapse=", ")`\}$ opened, $Z_\text{Drop} = `r format(round(total_cost_h(S_drop_h), 1), big.mark=" ")`$ €/year — **identical to the result of the add heuristic!**
:::
:::

:::
::: {.column width="46%"}

```{r drop-solution-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 5.5

final_open_drop <- S_drop_h
final_asgn_drop <- best_assign(final_open_drop)

lp_drop <- data.frame(
  id = paste0("S", 1:nrow(standorte)),
  x  = 0.18,
  y  =  seq(0.04 , 0.9, length.out = nrow(standorte)), #c(0.15, 0.50, 0.85),
  open = seq_len(n_I) %in% final_open_drop
)
kp_drop <- data.frame(
  id = paste0("C", 1:n_J),
  x  = 0.80,
  y  = c(0.92, 0.74, 0.56, 0.38, 0.20, 0.04)
)
lxy_d <- setNames(lp_drop[, c("x","y")], c("x","y")); rownames(lxy_d) <- lp_drop$id
kxy_d <- setNames(kp_drop[, c("x","y")], c("x","y")); rownames(kxy_d) <- kp_drop$id
r_d <- 0.048

fig_d <- plot_ly(type = "scatter", mode = "markers")

for (ii in 1:n_I) for (jj in 1:n_J) {
  wi <- paste0("S", ii); cj <- paste0("C", jj)
  x0 <- lxy_d[wi,"x"]; y0 <- lxy_d[wi,"y"]
  x1 <- kxy_d[cj,"x"]; y1 <- kxy_d[cj,"y"]
  dx <- x1-x0; dy <- y1-y0; dd <- sqrt(dx^2+dy^2)
  ax <- x0+r_d*dx/dd; ay <- y0+r_d*dy/dd
  ex <- x1-r_d*dx/dd; ey <- y1-r_d*dy/dd
  fig_d <- fig_d %>% add_annotations(
    x=ex,y=ey,ax=ax,ay=ay,xref="x",yref="y",axref="x",ayref="y",
    showarrow=TRUE,arrowhead=0,arrowsize=0.8,arrowwidth=0.8,
    arrowcolor="#D0D5DD",text="")
}

for (jj in 1:n_J) {
  wi <- paste0("S", final_asgn_drop[jj]); cj <- paste0("C", jj)
  x0 <- lxy_d[wi,"x"]; y0 <- lxy_d[wi,"y"]
  x1 <- kxy_d[cj,"x"]; y1 <- kxy_d[cj,"y"]
  dx <- x1-x0; dy <- y1-y0; dd <- sqrt(dx^2+dy^2)
  ax <- x0+r_d*dx/dd; ay <- y0+r_d*dy/dd
  ex <- x1-r_d*dx/dd; ey <- y1-r_d*dy/dd
  fig_d <- fig_d %>% add_annotations(
    x=ex,y=ey,ax=ax,ay=ay,xref="x",yref="y",axref="x",ayref="y",
    showarrow=TRUE,arrowhead=2,arrowsize=1,arrowwidth=2.2,
    arrowcolor="#da7101",text="")
}

lc_d <- ifelse(lp_drop$open, "#BFDBF7", "#E8E8E8")
lb_d <- ifelse(lp_drop$open, "#4B5563", "#B0B0B0")
fig_d <- fig_d %>% add_trace(
  x=lp_drop$x, y=lp_drop$y, text=lp_drop$id,
  mode="markers+text",
  marker=list(symbol="hexagon",size=38,color=lc_d,line=list(color=lb_d,width=2)),
  textfont=list(size=14,color="#1F2933",family="Arial"),
  textposition="middle center",hoverinfo="skip",showlegend=FALSE)
fig_d <- fig_d %>% add_annotations(
  x=lp_drop$x-0.09, y=lp_drop$y, xref="x",yref="y",
  text=ifelse(lp_drop$open, paste0("f=",round(f_h/1000),"k€ ✓"), paste0("f=",round(f_h/1000),"k€")),
  showarrow=FALSE,
  font=list(size=11,color=ifelse(lp_drop$open,"#da7101","#9CA3AF"),family="Arial"),
  xanchor="right")
fig_d <- fig_d %>% add_trace(
  x=kp_drop$x, y=kp_drop$y, text=kp_drop$id,
  mode="markers+text",
  marker=list(symbol="circle",size=38,color="#2D9E8E",
              line=list(color="#2F3E46",width=1.5)),
  textfont=list(size=14,color="white",family="Arial"),
  textposition="middle center",hoverinfo="skip",showlegend=FALSE)
fig_d <- fig_d %>% add_annotations(
  x=kp_drop$x+0.07, y=kp_drop$y, xref="x",yref="y",
  text=paste0("μ=",c(120,150,280,90,110,95)),
  showarrow=FALSE,font=list(size=11,color="#1F2933",family="Arial"),
  xanchor="left")
fig_d <- fig_d %>% layout(
  xaxis=list(visible=FALSE,range=c(0,1),fixedrange=TRUE),
  yaxis=list(visible=FALSE,range=c(-0.05,1.05),fixedrange=TRUE),
  plot_bgcolor="transparent",paper_bgcolor="transparent",
  margin=list(l=40,r=20,t=10,b=10),height=420, width = 650)
fig_d
```

:::
:::

---

## Warehouse Location Problem {#slide-heuristic-comparison}
::: slide-subtitle
Comparison of solutions
:::

```{r comparison-compute}
#| echo: false
#| warning: false
#| message: false

# Merge all solution data
cost_opt_val  <- total_cost_h(offene_lager)
cost_add_val  <- total_cost_h(S_add_h)
cost_drop_val <- total_cost_h(S_drop_h)

fix_opt  <- sum(f_h[c(1)])
tr_opt   <- cost_opt_val  - fix_opt
fix_add  <- sum(f_h[S_add_h])
tr_add   <- cost_add_val  - fix_add
fix_drop <- sum(f_h[S_drop_h])
tr_drop  <- cost_drop_val - fix_drop

asgn_opt  <- best_assign(c(1))
asgn_add  <- best_assign(S_add_h)
asgn_drop <- best_assign(S_drop_h)

gap_add  <- round((cost_add_val  / cost_opt_val - 1) * 100, 2)
gap_drop <- round((cost_drop_val / cost_opt_val - 1) * 100, 2)
```

::: columns
::: {.column width="50%"}

```{r comparison-table}
#| echo: false
#| warning: false
#| message: false

comp_df <- data.frame(
  Kennzahl = c(
    "Opened locations",
    "Number of locations",
    "Fixed costs (€)",
    "Transport costs (€)",
    "Total costs Z (€)",
    "Gap to optimum (%)"
  ),
  Optimum = c(
    paste0("S", sort(offene_lager), collapse=", "),
    length(offene_lager),
    fix_opt,
    round(tr_opt, 1),
    round(cost_opt_val, 1),
    "–"
  ),
  Add = c(
    paste0("S", sort(S_add_h), collapse=", "),
    length(S_add_h),
    fix_add,
    round(tr_add, 1),
    round(cost_add_val, 1),
    paste0(gap_add, "%")
  ),
  Drop = c(
    paste0("S", sort(S_drop_h), collapse=", "),
    length(S_drop_h),
    fix_drop,
    round(tr_drop, 1),
    round(cost_drop_val, 1),
    paste0(gap_drop, "%")
  )
)

kbl(comp_df,
    col.names = c("Metric", "Optimum (MILP)", "Add heuristic", "Drop heuristic"),
    align = c("l","c","c","c"), escape = FALSE,
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 24) |>
  column_spec(1, bold = FALSE) |>
  column_spec(2, bold = TRUE, background = "#a12c7b", color = "white") |>
  column_spec(3, background = "#E8F4F1") |>
  column_spec(4, background = "#FEF3E2") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(5, bold = TRUE)
```

::: {.callout-note appearance="simple"}
In this case example, the **add heuristic and the drop heuristic independently** find the same, **provably optimal solution** ($Z = `r format(round(cost_add_val,1), big.mark=" ")`$ €/year, only location S1 = Wiesbaden-Erbenheim). This is by no means a given, but here it shows the clear cost advantage of this location.
:::

:::
::: {.column width="48%"}

**Cost comparison**

```{r comparison-bar}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4.5

methoden  <- c("Optimum (MILP)", "Add heuristic", "Drop heuristic")
fix_vals  <- round(c(fix_opt,  fix_add,  fix_drop) / 1000, 1)
tr_vals   <- round(c(tr_opt, tr_add, tr_drop) / 1000, 1)
farben_fix <- c("#a12c7b", "#01696f", "#da7101")
farben_tr  <- c("#d163a7", "#4f98a3", "#fdab43")

fig_bar <- plot_ly()

fig_bar <- fig_bar %>% add_trace(
  x = methoden, y = fix_vals, type = "bar", name = "Fixed costs",
  marker = list(color = farben_fix),
  text = paste0(fix_vals, " k€"), textposition = "inside",
  insidetextanchor = "middle",
  textfont = list(color = "white", size = 13, family = "Arial")
)

fig_bar <- fig_bar %>% add_trace(
  x = methoden, y = tr_vals, type = "bar", name = "Transport costs",
  marker = list(color = farben_tr),
  text = paste0(tr_vals, " k€"), textposition = "inside",
  insidetextanchor = "middle",
  textfont = list(color = "white", size = 13, family = "Arial")
)

# Total cost label on top
total_vals <- fix_vals + tr_vals
fig_bar <- fig_bar %>% add_annotations(
  x = methoden, y = total_vals + 12,
  text = paste0("<b>", total_vals, " k€</b>"),
  showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial")
)

fig_bar <- fig_bar %>% layout(
  barmode = "stack",
  xaxis   = list(title = ""),
  yaxis   = list(title = "Costs (k€/year)", range = c(0, max(total_vals) * 1.15)),
  legend  = list(orientation = "h", y = -0.18),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 20, t = 30, b = 60),
  height = 380, width = 650
)

fig_bar
```

:::
:::

---

## Qualitative vs. Quantitative Location Assessment {#slide-ahp-vs-wlp}
::: slide-subtitle
Discussion
:::

::: columns
::: {.column width="52%"}

::: {.callout-warning appearance="simple"}

## A contradiction with Chapter 4?

The **AHP analysis** in Chapter 4 narrowly favoured **Mainz-Hechtsheim**. The **quantitative WLP analysis** in this chapter favours **Wiesbaden-Erbenheim** and **Raunheim**. How do these fit together?
:::

::: {.fragment .fade-up}
```{r ahp-vs-wlp-tabelle}
#| echo: false
#| warning: false
#| message: false

vergleich_ahp <- data.frame(
  Standort = c("Raunheim", "Wiesbaden-Erbenheim"),
  Fixkosten = c(260000, 180000),
  Transport = c(round(.wlp_cost(3, f_h, c_h, a_h)$transport, 1),
                round(.wlp_cost(1, f_h, c_h, a_h)$transport, 1))
) |>
  dplyr::mutate(Gesamt = Fixkosten + Transport)

kbl(vergleich_ahp,
    col.names = c("Location (single site)", "Fixed costs (€)",
                  "Transport (€/year)", "Total (€/year)"),
    align = c("l","r","r","r"),
    format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 22) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```
:::

::: {.fragment .fade-up}
Raunheim and Wiesbaden-Erbenheim have lower transport costs; the savings compensate for the higher fixed costs.
:::

:::
::: {.column width="44%"}

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## Conclusion

- **AHP (Chapter 4):** good for an initial **pre-selection** based on soft criteria that are hard to quantify
- **WLP (Chapter 5):** delivers the robust **cost-based decision** once hard figures are available
- The two methods **complement** each other – neither replaces the other
:::
:::

::: {.fragment .fade-up}
::: {.callout-note appearance="simple"}

## SushiFresh decision

Consolidation of the production site at **Wiesbaden-Erbenheim**. This decision is revisited in Chapter 6 (risk pooling) and Chapter 8 (bullwhip effect).
:::
:::

:::
:::
