

```{r}
#| echo: false
#| warning: false
#| message: false


library(quarto)
library(readr)
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(planr)
library(dplyr)
library(MASS)

source("../R/scm_functions.R")
```


# Uncertainty and Risk Management in Supply Chains

## Uncertainty in Supply Chains

- **Pooling**
- Newsvendor model
- Bullwhip effect

## Learning objectives {#learning-objectives-uncertainty-pooling}

After this unit you will be able to …

- **explain** sources and effects of uncertainty in supply chains,
- **distinguish** between demand uncertainty, implied uncertainty and responsiveness,
- **analyse** the effect of risk pooling on demand variance and safety stocks,
- **compare and assess** centralised and decentralised inventory structures under uncertainty.

**Further reading**

- **Pooling, demand variability and inventory decisions:** [@cachon2024matching, Chapters 10–13].
- **Safety stocks, service levels and inventory management:** [@silver2017inventory, Chapters 6–8].

---

## Case study: SushiFresh GmbH {#slide-sushifresh-pooling-intro}
::: slide-subtitle
Six stores, one new central plant
:::

::: {.callout-note appearance="simple"}

## Initial situation
**SushiFresh GmbH** is a fictitious regional producer of fresh sushi boxes in the Rhine-Main area, supplying **six stores** daily. In Chapter 5, the decision was made to consolidate production at **one central site in Wiesbaden-Erbenheim**. This chapter asks: how much safety stock is needed — decentralised vs. centralised?
:::

```{r sushifresh-filialtabelle}
#| echo: false
#| warning: false
#| message: false

sf_filialen <- data.frame(
  Filiale = c("Wiesbaden", "Mainz", "Frankfurt", "Rüsselsheim", "Darmstadt", "Offenbach"),
  mu      = c(120, 150, 280, 90, 110, 95),
  sigma   = c(25, 30, 55, 20, 24, 22)
)

kbl(sf_filialen,
    col.names = c("Store", "μ (boxes/day)", "σ"),
    align = c("l", "r", "r")) |>
  kable_styling(bootstrap_options = c("condensed", "striped"), full_width = TRUE) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(3, bold = TRUE, background = "#FFF3CD")
```

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}
Total demand: **845 boxes/day**. Frankfurt (bold) is the largest **and** most volatile store.
:::
:::

---

## The nature of uncertainty {#slide-pooling-1}
::: slide-subtitle
Implied uncertainty in supply chains [@chopra2025supply]
:::

::: {.callout-note appearance="minimal" icon=false}
**Implied uncertainty** denotes the uncertainty a supply chain **must** cope with — determined by the combination of **demand uncertainty** and the market's **service requirements** [@chopra2025supply, Ch. 2].
:::

::: columns
::: {.column width="52%"}

::: {.fragment .fade-up}
::: smaller-text
**Two sources of implied uncertainty**

| Dimension | Low | High |
|---|---|---|
| **Demand uncertainty** | Salt, electricity | Fashion, electronics |
| **Supply variability** | Concrete, coal | Semiconductors, fishing |

$\rightarrow$ High implied uncertainty requires **responsive** supply chains with buffer capacity.
:::
:::

:::
::: {.column width="46%"}

::: {.fragment .fade-up}

```{r implied-uncertainty-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3.5

# Implied uncertainty spectrum according to Chopra & Meindl
produkte <- data.frame(
  label     = c("Salt", "Soup", "Beer", "Clothing",
                "PC", "Fashion", "Phone chips"),
  iu_score  = c(0.05, 0.18, 0.30, 0.52, 0.65, 0.82, 0.94),
  kategorie = c("Functional", "Functional", "Functional",
                "Hybrid", "Hybrid", "Innovative", "Innovative")
)

farben <- c("Functional" = "#52B788",
            "Hybrid"     = "#E9C46A",
            "Innovative" = "#E76F51")

fig_iu <- plot_ly(produkte,
  x    = ~iu_score,
  y    = rep(1, nrow(produkte)),
  type = "scatter", mode = "markers+text",
  color = ~kategorie, colors = farben,
  marker = list(size = 22, opacity = 0.9,
                line = list(color = "white", width = 1.5)),
  text        = ~label,
  textposition = "top center",
  textfont    = list(size = 12, family = "Arial", color = "#1F2933")
) %>%
  layout(
    xaxis = list(
      title    = "Implied uncertainty →",
      range    = c(-0.05, 1.05),
      tickvals = c(0, 0.5, 1),
      ticktext = c("Low", "Medium", "High"),
      fixedrange = TRUE
    ),
    yaxis = list(
      visible    = FALSE,
      range      = c(0.8, 1.5),
      fixedrange = TRUE
    ),
    legend = list(orientation = "h", y = -0.15, title = list(text = "Product type")),
    annotations = list(
      list(
        x = 0.12, y = 0.82, xref = "x", yref = "paper",
        text = "<b>Efficient SC</b><br>little buffer needed",
        showarrow = FALSE,
        font = list(size = 11, color = "#52B788", family = "Arial"),
        align = "center"
      ),
      list(
        x = 0.88, y = 0.82, xref = "x", yref = "paper",
        text = "<b>Responsive SC</b><br>buffers & flexibility",
        showarrow = FALSE,
        font = list(size = 11, color = "#E76F51", family = "Arial"),
        align = "center"
      )
    ),
    shapes = list(list(
      type = "line",
      x0 = 0.5, x1 = 0.5, y0 = 0, y1 = 1,
      xref = "x", yref = "paper",
      line = list(color = "#9CA3AF", dash = "dot", width = 1.2)
    )),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 30, r = 20, t = 20, b = 60),
    height = 330
  )

fig_iu
```

:::

:::
:::

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}

## The basic dilemma
- **Demand** is variable and hard to predict
- **Capacities** (plants, warehouses, staff) are fixed in advance
- $\rightarrow$ A mismatch between supply and demand is **structurally unavoidable**
:::
:::

---

## The nature of uncertainty {#slide-pooling-mismatch}
::: slide-subtitle
Mismatch: fluctuating demand vs. fixed capacity
:::

::: columns
::: {.column width="52%"}

::: {.callout-note appearance="minimal" icon=false}
Capacities (machines, staff, storage space) are determined **ex ante**. Demand materialises **ex post**. The result: a **systematic mismatch** in both directions.
:::

::: {.fragment .fade-up}
::: smaller-text

**Consequences of the mismatch**

| Scenario | Phenomenon | Consequence |
|---|---------|---------|
| $d > C$ | Overload | Lost sales, loss of revenue |
| $d < C$ | Underutilisation | Idle costs |
| $d \approx C$ | Equilibrium |  with a perfect forecast |

:::
:::

:::
::: {.column width="46%"}

::: {.fragment .fade-up}

```{r mismatch-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3.5

set.seed(12)
n_periods <- 24
mu_d <- 120
sigma_d <- 25
C <- 130

t_vals <- 1:n_periods
d_vals <- pmax(40, round(rnorm(n_periods, mu_d, sigma_d)))
prod_vals <- pmin(d_vals, C)
lost_vals <- pmax(0, d_vals - C)
idle_vals <- pmax(0, C - d_vals)

df_mm <- data.frame(
  t = t_vals, d = d_vals, prod = prod_vals,
  lost = lost_vals, idle = idle_vals
)

fig_mm <- plot_ly(df_mm, x = ~t)

# Capacity limit
fig_mm <- fig_mm %>% add_trace(
  x = c(1, n_periods), y = c(C, C),
  type = "scatter", mode = "lines",
  name = paste0("Capacity C = ", C),
  line = list(color = "#6B7280", width = 1.8, dash = "dash"),
  hoverinfo = "skip"
)

# Idle capacity (shaded)
fig_mm <- fig_mm %>% add_trace(
  x = ~t, y = ~idle,
  type = "bar", name = "Idle Capacity (C − d)",
  marker = list(color = "rgba(234,179,8,0.45)",
                line = list(color = "rgba(234,179,8,0.0)", width = 0)),
  base = ~prod
)

# Lost Sales (shaded above C)
fig_mm <- fig_mm %>% add_trace(
  x = ~t, y = ~lost,
  type = "bar", name = "Lost Sales (d − C)",
  marker = list(color = "rgba(220,38,38,0.55)",
                line = list(color = "rgba(220,38,38,0.0)", width = 0)),
  base = C
)

# Production
fig_mm <- fig_mm %>% add_trace(
  x = ~t, y = ~prod,
  type = "scatter", mode = "lines",
  name = "Production / sales",
  line = list(color = cols$top, width = 2)
)

# Demand
fig_mm <- fig_mm %>% add_trace(
  x = ~t, y = ~d,
  type = "scatter", mode = "lines+markers",
  name = "Demand d",
  line   = list(color = "#1D4ED8", width = 1.5),
  marker = list(color = "#1D4ED8", size = 5)
)

fig_mm <- fig_mm %>% layout(
  barmode = "overlay",
  xaxis = list(title = "Period", dtick = 3, fixedrange = TRUE),
  yaxis = list(title = "Units",
               range = c(40, max(d_vals) * 1.12),
               fixedrange = TRUE),
  legend = list(orientation = "h", y = -0.22,
                font = list(size = 11)),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 20, t = 10, b = 70),
  height = 370,
  width = 650
)

fig_mm
```

:::

:::
:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## Practical examples
- **Automotive industry:** plants produce only certain models for several years $\rightarrow$ utilisation depends on demand
- **Retail:** seasonal goods (Christmas items) are ordered too early, leftovers are sold off cheaply
- **Hospital:** operating theatre capacities are fixed daily, emergency volume is highly variable
:::
:::



---

## Dealing with uncertainty {#slide-cp-background}
::: slide-subtitle
Capacity pooling
:::

::: {.callout-note appearance="simple"}

## Initial situation (structure following Cachon & Terwiesch, Ch. 17)
In its **start-up phase**, SushiFresh initially operated only **2 pilot kitchens** for **2 pilot sites** (today's Wiesbaden and Mainz stores). Each kitchen is assigned exclusively to its own site – **no flexibility**. Capacity per kitchen is **90 boxes/day**. Demand is **uncertain** and is observed after capacities have been fixed.
:::

::: columns
::: {.column width="50%"}

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}

## Consequences of inflexibility
- If demand exceeds capacity $\rightarrow$ **shortage** (lost sales)
- If demand falls below capacity $\rightarrow$ **idle capacity**
- Both effects reduce **system performance**
:::
:::

:::
::: {.column width="48%"}

::: {.fragment .fade-up}

```{r cp-mismatch-setup}
#| echo: false
#| warning: false
#| message: false

# Empirical demand data from cp_data (months 1–12)
cp_data_bg <- data.frame(
  Monat = 1:12,
  dA    = c(78, 79, 92, 76, 95, 104, 87, 92, 91, 85, 96, 93),
  dB    = c(102, 92, 93, 87, 80,  97, 90, 76, 89, 86, 89, 93)
)
C_bg <- 90

make_mismatch_fig <- function(d_vals, C, werk_label, farbe_linie,
                               height = 290, width = 600) {
  t_vals    <- seq_along(d_vals)
  prod_vals <- pmin(d_vals, C)
  lost_vals <- pmax(0, d_vals - C)
  idle_vals <- pmax(0, C - d_vals)

  df <- data.frame(t = t_vals, d = d_vals, prod = prod_vals,
                   lost = lost_vals, idle = idle_vals)

  fig <- plot_ly(df, x = ~t)

  # Capacity limit
  fig <- fig %>% add_trace(
    x = c(1, max(t_vals)), y = c(C, C),
    type = "scatter", mode = "lines",
    name = paste0("C = ", C),
    line = list(color = "#6B7280", width = 1.6, dash = "dash"),
    hoverinfo = "skip",
    showlegend = TRUE
  )

  # Idle capacity (yellow)
  fig <- fig %>% add_trace(
    x = ~t, y = ~idle,
    type = "bar", name = "Idle Capacity (C − d)",
    marker = list(color = "rgba(234,179,8,0.45)",
                  line = list(color = "rgba(234,179,8,0.0)", width = 0)),
    base = ~prod
  )

  # Lost sales (red)
  fig <- fig %>% add_trace(
    x = ~t, y = ~lost,
    type = "bar", name = "Lost Sales (d − C)",
    marker = list(color = "rgba(220,38,38,0.55)",
                  line = list(color = "rgba(220,38,38,0.0)", width = 0)),
    base = C
  )

  # Production / sales
  fig <- fig %>% add_trace(
    x = ~t, y = ~prod,
    type = "scatter", mode = "lines",
    name = "Production",
    line = list(color = farbe_linie, width = 2)
  )

  # Demand line
  fig <- fig %>% add_trace(
    x = ~t, y = ~d,
    type = "scatter", mode = "lines+markers",
    name = "Demand d",
    line   = list(color = "#1D4ED8", width = 1.4),
    marker = list(color = "#1D4ED8", size = 5)
  )

  fig <- fig %>% layout(
    barmode = "overlay",
    title   = list(text = paste0("<b>", werk_label, "</b>"),
                   font = list(size = 13, color = "#1F2933", family = "Arial"),
                   x = 0.02, xanchor = "left"),
    xaxis   = list(title = "Month", dtick = 2, fixedrange = TRUE,
                   range = c(0.5, max(t_vals) + 0.5)),
    yaxis   = list(title = "Units",
                   range = c(60, max(d_vals) * 1.15),
                   fixedrange = TRUE),
    legend  = list(orientation = "h", y = -0.30,
                   font = list(size = 10)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 45, r = 10, t = 30, b = 65),
    height = height
  )

  fig
}
```

```{r cp-mismatch-werkA}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_mismatch_fig(
  d_vals      = cp_data_bg$dA,
  C           = C_bg,
  werk_label  = "Wiesbaden kitchen (pilot)",
  farbe_linie = cols$top,
  height      = 270,
  width       = 600
)
```

```{r cp-mismatch-werkB}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_mismatch_fig(
  d_vals      = cp_data_bg$dB,
  C           = C_bg,
  werk_label  = "Mainz kitchen (pilot)",
  farbe_linie = cols$mid,
  height      = 270,
  width       = 600
)
```

:::

:::
:::

---

## Capacity pooling {#slide-cp-original}
::: slide-subtitle
Original configuration – each kitchen supplies only its assigned store
:::

**Monthly demand and production, pilot phase ($C^{Wi} = C^{Ma} = 90$)**

::: columns
::: {.column width="25%"}

::: small-text

```{r cp-original-table}
#| echo: false
#| warning: false
#| message: false

C1 <- C2 <- 90

cp_data <- data.frame(
  Monat = 1:12,
  dA    = c(78, 79, 92, 76, 95,104, 87, 92, 91, 85, 96, 93),
  dB    = c(102, 92, 93, 87, 80, 97, 90, 76, 89, 86, 89, 93),
  xA    = c(78, 79, 90, 76, 90, 90, 87, 90, 90, 85, 90, 90),
  xB    = c(90, 90, 90, 87, 80, 90, 90, 76, 89, 86, 89, 90)
)

cp.means <- colMeans(cp_data)
cp.means[1] <- NA

cp_data <- rbind(cp_data,cp.means)

cp_data$Monat <- as.character(cp_data$Monat)

cp_data$Monat[13] <- "&#8960;"

kbl(cp_data,
    col.names = c("Month", "d<sub>t</sub><sup>Wi</sup>", "d<sub>t</sub><sup>Ma</sup>",
                  "x<sub>t</sub><sup>Wi</sup>", "x<sub>t</sub><sup>Ma</sup>"),
    align = c("c","r","r","r","r"),
    escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 20) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(13, bold = TRUE, background = "#2F3E4640")
```

:::
:::


::: {.column width="27%"}

```{r cp-original-graph}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3

# Original setting: C1->1->A, C2->2->B (separate lines, no pooling)
plants <- data.frame(
  id  = c("P1","P2"),
  lbl = c("1","2"),
  cl  = c("C<sup>Wi</sup>","C<sup>Ma</sup>"),
  x   = 0.35,
  y   = c(0.72, 0.28)
)

markets <- data.frame(
  id  = c("A","B"),
  lbl = c("A","B"),
  dj  = c("d<sup>Wi</sup>","d<sup>Ma</sup>"),
  x   = 0.72,
  y   = c(0.72, 0.28)
)

pxy <- setNames(plants[,  c("x","y")], c("x","y")); rownames(pxy) <- plants$id
mxy <- setNames(markets[, c("x","y")], c("x","y")); rownames(mxy) <- markets$id

r <- 0.055  # node radius in plot coordinates

edges <- data.frame(
  from = c("P1","P2"),
  to   = c("A","B"),
  lbl  = c("x<sup>Wi,Wi</sup>","x<sup>Ma,Ma</sup>"),
  dash = c(FALSE, FALSE),
  stringsAsFactors = FALSE
)

fig <- plot_ly(type = "scatter", mode = "markers")

for (i in seq_len(nrow(edges))) {
  x0 <- pxy[edges$from[i], "x"];  y0 <- pxy[edges$from[i], "y"]
  x1 <- mxy[edges$to[i],   "x"];  y1 <- mxy[edges$to[i],   "y"]
  dx <- x1 - x0;  dy <- y1 - y0;  d <- sqrt(dx^2 + dy^2)
  ax <- x0 + r * dx / d;  ay <- y0 + r * dy / d
  ex <- x1 - r * dx / d;  ey <- y1 - r * dy / d

  fig <- fig %>% add_annotations(
    x = ex, y = ey, ax = ax, ay = ay,
    xref = "x", yref = "y", axref = "x", ayref = "y",
    showarrow = TRUE, arrowhead = 2, arrowsize = 1,
    arrowwidth = 1.5, arrowcolor = "#4B4B4B", text = ""
  )

  if (nchar(edges$lbl[i]) > 0) {
    fig <- fig %>% add_annotations(
      x = (ax + ex) / 2, y = (ay + ey) / 2 + 0.06,
      xref = "x", yref = "y",
      text = edges$lbl[i], showarrow = FALSE,
      font = list(size = 13, color = "#1F2933", family = "Arial")
    )
  }
}

# Cost labels left of the plants
fig <- fig %>% add_annotations(
  x = plants$x - 0.12, y = plants$y,
  xref = "x", yref = "y",
  text = plants$cl, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "right"
)

# Plant nodes (hexagon)
fig <- fig %>% add_trace(
  x = plants$x, y = plants$y, text = plants$lbl,
  mode = "markers+text",
  marker = list(symbol = "hexagon", size = 42,
                color = "#BFDBF7", line = list(color = "#4B5563", width = 1.5)),
  textfont = list(size = 16, color = "#1F2933", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# Market nodes (circle)
fig <- fig %>% add_trace(
  x = markets$x, y = markets$y, text = markets$lbl,
  mode = "markers+text",
  marker = list(symbol = "circle", size = 42,
                color = "#2D9E8E", line = list(color = "#2F3E46", width = 1.5)),
  textfont = list(size = 16, color = "white", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# Demand labels right of the markets
fig <- fig %>% add_annotations(
  x = markets$x + 0.09, y = markets$y,
  xref = "x", yref = "y",
  text = markets$dj, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "left"
)

# Set labels at the top
fig <- fig %>% add_annotations(
  x = c(0.35, 0.72), y = c(1.00, 1.00),
  xref = "x", yref = "y",
  text = c("<b>Kitchen</b>", "<b>Store</b>"),
  showarrow = FALSE,
  font = list(size = 13, color = "#1F2933", family = "Arial")
)

fig <- fig %>% layout(
  xaxis = list(visible = FALSE, range = c(0, 1), fixedrange = TRUE),
  yaxis = list(visible = FALSE, range = c(0, 1.08),  fixedrange = TRUE),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 50, t = 10, b = 10),
  height = 250,
  width  = 400
)

fig
```
:::


::: {.column width="46%"}

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## System Evaluation

- **Expected demand:** $E(d^{Wi})=$ `r round(mean(cp_data$dA),1)`; $E(d^{Ma})=$ `r round(mean(cp_data$dB),1)`
- **Expected production:** $E(x^{Wi})=$ `r round(mean(cp_data$xA),1)`; $E(x^{Ma})=$ `r round(mean(cp_data$xB),1)`
- **Site utilisation:** $\frac{E(x^{Wi})}{C^{Wi}}=$ `r round(mean(cp_data$xA)/C1*100,1)`%; $\frac{E(x^{Ma})}{C^2}=$ `r round(mean(cp_data$xB)/C2*100,1)`%
- **Expected fill rate:** $\frac{E(x^{Wi})}{E(d^{Wi})}=$ `r round(mean(cp_data$xA/cp_data$dA)*100,1)`%; $\frac{E(x^{Ma})}{E(d^{Ma})}=$ `r round(mean(cp_data$xB/cp_data$dB)*100,1)`%
:::
:::


:::
:::

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}

## Problem
Utilisation and fill rate are high, yet there are **lost sales** and **underutilisation** $\rightarrow$ Can helping each other out (**capacity pooling**) improve these KPIs?
:::
:::


---

## Capacity pooling {#slide-cp-pooling-1side}
::: slide-subtitle
One-sided pooling – the Wiesbaden kitchen can supply both sites
:::

::: columns
::: {.column width="30%"}

::: small-text
```{r cp-pooling-table}
#| echo: false
#| warning: false
#| message: false

C1 <- 90; C2 <- 90

cp_pool_1 <- cp_data[-13,] %>%
  mutate(
    x1A = pmin(dA, C1),
    x1B = ifelse(dB > C2 & dA < C1,
                 pmin(C1 - dA, dB - C2), 0),
    x1  = x1A + x1B,
    x2B = pmin(C2, dB),
    xB =  x2B + x1B
  )

kbl(cp_pool_1[, c("Monat","dA","dB","x1A","x1B","x2B")],
    col.names = c("Month", "d<sub>t</sub><sup>Wi</sup>", "d<sub>t</sub><sup>Ma</sup>",
                  "x<sub>t</sub><sup>Wi,Wi</sup>", "x<sub>t</sub><sup>Wi,Ma</sup>", "x<sub>t</sub><sup>2B</sup>"),
    align = c("c","r","r","r","r","r"),
    escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 22) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  column_spec(4:5, background = "#E8F4F1") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```
:::
:::


::: {.column width="25%"}

```{r cp-pooling-graph}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3

# Pooling setting: C1->1->A (solid), C1->1->B (dashed), C2->2->B (solid)
plants <- data.frame(
  id  = c("P1","P2"),
  lbl = c("1","2"),
  cl  = c("C<sup>Wi</sup>","C<sup>Ma</sup>"),
  x   = 0.35,
  y   = c(0.72, 0.28)
)

markets <- data.frame(
  id  = c("A","B"),
  lbl = c("A","B"),
  dj  = c("d<sup>Wi</sup>","d<sup>Ma</sup>"),
  x   = 0.72,
  y   = c(0.72, 0.28)
)

pxy <- setNames(plants[,  c("x","y")], c("x","y")); rownames(pxy) <- plants$id
mxy <- setNames(markets[, c("x","y")], c("x","y")); rownames(mxy) <- markets$id

r <- 0.06  # node radius in plot coordinates

edges <- data.frame(
  from  = c("P1","P1","P2"),
  to    = c("A",  "B",  "B"),
  lbl   = c("x<sup>Wi,Wi</sup>", "x<sup>Wi,Ma</sup>", "x<sup>Ma,Ma</sup>"),
  dash  = c(FALSE, TRUE, FALSE),   # TRUE = dashed (pooling arrow)
  stringsAsFactors = FALSE
)

fig <- plot_ly(type = "scatter", mode = "markers")

for (i in seq_len(nrow(edges))) {
  x0 <- pxy[edges$from[i], "x"];  y0 <- pxy[edges$from[i], "y"]
  x1 <- mxy[edges$to[i],   "x"];  y1 <- mxy[edges$to[i],   "y"]
  dx <- x1 - x0;  dy <- y1 - y0;  d <- sqrt(dx^2 + dy^2)
  ax <- x0 + r * dx / d;  ay <- y0 + r * dy / d
  ex <- x1 - r * dx / d;  ey <- y1 - r * dy / d

  # Dashed arrow via line segments (Plotly does not support dash in add_annotations)
  if (edges$dash[i]) {
    # Segmented dashed line with 6 sections
    n_seg <- 10
    xs <- seq(ax, ex, length.out = n_seg + 1)
    ys <- seq(ay, ey, length.out = n_seg + 1)
    for (s in seq(1, n_seg, by = 2)) {
      fig <- fig %>% add_segments(
        x = xs[s], xend = xs[s+1],
        y = ys[s], yend = ys[s+1],
        line = list(color = "#4B4B4B", width = 1.8, dash = "solid"),
        hoverinfo = "skip", showlegend = FALSE
      )
    }
    # Arrowhead at the end point
    fig <- fig %>% add_annotations(
      x = ex, y = ey, ax = xs[n_seg], ay = ys[n_seg],
      xref = "x", yref = "y", axref = "x", ayref = "y",
      showarrow = TRUE, arrowhead = 2, arrowsize = 1,
      arrowwidth = 1.8, arrowcolor = "#4B4B4B", text = ""
    )
  } else {
    fig <- fig %>% add_annotations(
      x = ex, y = ey, ax = ax, ay = ay,
      xref = "x", yref = "y", axref = "x", ayref = "y",
      showarrow = TRUE, arrowhead = 2, arrowsize = 1,
      arrowwidth = 1.5, arrowcolor = "#4B4B4B", text = ""
    )
  }

  # Edge label
  if (nchar(edges$lbl[i]) > 0) {
    # x^1B label slightly offset (crossing arrow)
    lbl_off_x <- ifelse(edges$dash[i],  0.03, 0.00)
    lbl_off_y <- ifelse(edges$dash[i], -0.05, 0.06)
    fig <- fig %>% add_annotations(
      x = (ax + ex) / 2 + lbl_off_x,
      y = (ay + ey) / 2 + lbl_off_y,
      xref = "x", yref = "y",
      text = edges$lbl[i], showarrow = FALSE,
      font = list(size = 13, color = "#1F2933", family = "Arial")
    )
  }
}

# Cost labels left of the plants
fig <- fig %>% add_annotations(
  x = plants$x - 0.12, y = plants$y,
  xref = "x", yref = "y",
  text = plants$cl, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "right"
)

# Plant nodes (hexagon)
fig <- fig %>% add_trace(
  x = plants$x, y = plants$y, text = plants$lbl,
  mode = "markers+text",
  marker = list(symbol = "hexagon", size = 42,
                color = "#BFDBF7", line = list(color = "#4B5563", width = 1.5)),
  textfont = list(size = 16, color = "#1F2933", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# Market nodes (circle)
fig <- fig %>% add_trace(
  x = markets$x, y = markets$y, text = markets$lbl,
  mode = "markers+text",
  marker = list(symbol = "circle", size = 42,
                color = "#2D9E8E", line = list(color = "#2F3E46", width = 1.5)),
  textfont = list(size = 16, color = "white", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# Demand labels right of the markets
fig <- fig %>% add_annotations(
  x = markets$x + 0.09, y = markets$y,
  xref = "x", yref = "y",
  text = markets$dj, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "left"
)

# Set labels at the top
fig <- fig %>% add_annotations(
  x = c(0.35, 0.72), y = c(1.00, 1.00),
  xref = "x", yref = "y",
  text = c("<b>Kitchen</b>", "<b>Store</b>"),
  showarrow = FALSE,
  font = list(size = 13, color = "#1F2933", family = "Arial")
)

fig <- fig %>% layout(
  xaxis = list(visible = FALSE, range = c(0.2, .9), fixedrange = TRUE),
  yaxis = list(visible = FALSE, range = c(0.2, 1),  fixedrange = TRUE),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 50, t = 10, b = 10),
  height = 250,
  width  = 400
)

fig
```
:::


::: {.column width="44%"}

::: {.fragment .fade-up}
::: {.callout-note appearance="minimal" icon=false}
**Pooling rule** (if $d_t^{Ma} > C^{Ma}$ and $d_t^{Wi} < C^{Wi}$):

$$x_t^1 = d_t^{Wi} + \min(C^{Wi} - d_t^{Wi};\; d_t^{Ma} - C^{Ma})$$
$$x_t^{Ma,Ma} = C^{Ma}$$
:::
:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## System Evaluation (Pooling)

- **Production plant 1:** $E(x^{Wi})=$ `r round(mean(cp_pool_1$x1),1)`
- **Supply to market B:** $E(x^{Wi,Ma}+x^{Ma,Ma})=$ `r round(mean(cp_pool_1$xB),1)`
- **Utilisation plant 1:** $\frac{E(x^{Wi})}{C^{Wi}}=$ `r round(mean(cp_pool_1$x1)/C1*100,1)`%
- **Fill rate market B:** $\frac{E(x^{Wi,Ma}+x^{Ma,Ma})}{E(d^{Ma})}=$ `r round(mean(cp_pool_1$xB)/mean(cp_pool_1$dB)*100,1)`%
:::
:::

:::
:::

---

## Capacity pooling {#slide-cp-pooling-2side}
::: slide-subtitle
Two-sided pooling – both kitchens can supply both sites
:::


:::: columns
::: {.column width="30%"}

::: small-text
```{r cp-pooling-table2}
#| echo: false
#| warning: false
#| message: false

cp_pool <- cp_data[-13,] %>%
  mutate(
    x1A = pmin(dA, C1),
    x1B = ifelse(dB > C2 & dA < C1,
                 pmin(C1 - dA, dB - C2), 0),
    x1  = x1A + x1B,
    x2B = pmin(C2, dB),
    x2A = ifelse(dA > C1 & dB < C2,
                 pmin(C2 - dB, dA - C1), 0),
    x2  = x2A + x2B,
    xA = x2A + x1A,
    xB = x2B + x1B
  )

kbl(cp_pool[, c("Monat","dA","dB","x1A","x1B","x2B", "x2A")],
    col.names = c("Month", "d<sub>t</sub><sup>Wi</sup>", "d<sub>t</sub><sup>Ma</sup>",
                  "x<sub>t</sub><sup>Wi,Wi</sup>", "x<sub>t</sub><sup>Wi,Ma</sup>", "x<sub>t</sub><sup>Ma,Ma</sup>", "x<sub>t</sub><sup>Ma,Wi</sup>"),
    align = c("c","r","r","r","r","r","r"),
    escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed","striped"),
                full_width = TRUE, font_size = 22) |>
  column_spec(1, bold = TRUE, background = "#2F3E46", color = "white") |>
  column_spec(4:5, background = "#E8F4F1") |>
  column_spec(6:7, background = "#E8F4A1") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```
:::
:::


::: {.column width="25%"}

```{r cp-pooling-graph2}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3

# Pooling setting: C1->1->A (solid), C1->1->B (dashed), C2->2->B (solid)
plants <- data.frame(
  id  = c("P1","P2"),
  lbl = c("1","2"),
  cl  = c("C<sup>Wi</sup>","C<sup>Ma</sup>"),
  x   = 0.35,
  y   = c(0.72, 0.28)
)

markets <- data.frame(
  id  = c("A","B"),
  lbl = c("A","B"),
  dj  = c("d<sup>Wi</sup>","d<sup>Ma</sup>"),
  x   = 0.72,
  y   = c(0.72, 0.28)
)

pxy <- setNames(plants[,  c("x","y")], c("x","y")); rownames(pxy) <- plants$id
mxy <- setNames(markets[, c("x","y")], c("x","y")); rownames(mxy) <- markets$id

r <- 0.06  # node radius in plot coordinates

edges <- data.frame(
  from  = c("P1","P1","P2", "P2"),
  to    = c("A",  "B",  "B", "A"),
  lbl   = c("x<sup>Wi,Wi</sup>", "x<sup>Wi,Ma</sup>", "x<sup>Ma,Ma</sup>",  "x<sup>Ma,Wi</sup>"),
  dash  = c(FALSE, TRUE, FALSE, TRUE),   # TRUE = dashed (pooling arrow)
  stringsAsFactors = FALSE
)

fig <- plot_ly(type = "scatter", mode = "markers")

for (i in seq_len(nrow(edges))) {
  x0 <- pxy[edges$from[i], "x"];  y0 <- pxy[edges$from[i], "y"]
  x1 <- mxy[edges$to[i],   "x"];  y1 <- mxy[edges$to[i],   "y"]
  dx <- x1 - x0;  dy <- y1 - y0;  d <- sqrt(dx^2 + dy^2)
  ax <- x0 + r * dx / d;  ay <- y0 + r * dy / d
  ex <- x1 - r * dx / d;  ey <- y1 - r * dy / d

  # Dashed arrow via line segments (Plotly does not support dash in add_annotations)
  if (edges$dash[i]) {
    # Segmented dashed line with 6 sections
    n_seg <- 10
    xs <- seq(ax, ex, length.out = n_seg + 1)
    ys <- seq(ay, ey, length.out = n_seg + 1)
    for (s in seq(1, n_seg, by = 2)) {
      fig <- fig %>% add_segments(
        x = xs[s], xend = xs[s+1],
        y = ys[s], yend = ys[s+1],
        line = list(color = "#4B4B4B", width = 1.8, dash = "solid"),
        hoverinfo = "skip", showlegend = FALSE
      )
    }
    # Arrowhead at the end point
    fig <- fig %>% add_annotations(
      x = ex, y = ey, ax = xs[n_seg], ay = ys[n_seg],
      xref = "x", yref = "y", axref = "x", ayref = "y",
      showarrow = TRUE, arrowhead = 2, arrowsize = 1,
      arrowwidth = 1.8, arrowcolor = "#4B4B4B", text = ""
    )
  } else {
    fig <- fig %>% add_annotations(
      x = ex, y = ey, ax = ax, ay = ay,
      xref = "x", yref = "y", axref = "x", ayref = "y",
      showarrow = TRUE, arrowhead = 2, arrowsize = 1,
      arrowwidth = 1.5, arrowcolor = "#4B4B4B", text = ""
    )
  }

  # Edge label
  # x^1B label slightly offset (crossing arrow)
  lbl_off_x <- ifelse(i >= 3, ifelse(i==3, 0.03 , -0.02), 0.00)
  lbl_off_y <- ifelse(i >= 3, ifelse(i==3, -0.05 , -0.08), 0.06)
  fig <- fig %>% add_annotations(
    x = (ax + ex) / 2 + lbl_off_x,
    y = (ay + ey) / 2 + lbl_off_y,
    xref = "x", yref = "y",
    text = edges$lbl[i], showarrow = FALSE,
    font = list(size = 13, color = "#1F2933", family = "Arial")
  )
  
}

# Cost labels left of the plants
fig <- fig %>% add_annotations(
  x = plants$x - 0.12, y = plants$y,
  xref = "x", yref = "y",
  text = plants$cl, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "right"
)

# Plant nodes (hexagon)
fig <- fig %>% add_trace(
  x = plants$x, y = plants$y, text = plants$lbl,
  mode = "markers+text",
  marker = list(symbol = "hexagon", size = 42,
                color = "#BFDBF7", line = list(color = "#4B5563", width = 1.5)),
  textfont = list(size = 16, color = "#1F2933", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# Market nodes (circle)
fig <- fig %>% add_trace(
  x = markets$x, y = markets$y, text = markets$lbl,
  mode = "markers+text",
  marker = list(symbol = "circle", size = 42,
                color = "#2D9E8E", line = list(color = "#2F3E46", width = 1.5)),
  textfont = list(size = 16, color = "white", family = "Arial"),
  textposition = "middle center",
  hoverinfo = "skip", showlegend = FALSE
)

# Demand labels right of the markets
fig <- fig %>% add_annotations(
  x = markets$x + 0.09, y = markets$y,
  xref = "x", yref = "y",
  text = markets$dj, showarrow = FALSE,
  font = list(size = 14, color = "#1F2933", family = "Arial"),
  xanchor = "left"
)

# Set labels at the top
fig <- fig %>% add_annotations(
  x = c(0.35, 0.72), y = c(1.00, 1.00),
  xref = "x", yref = "y",
  text = c("<b>Kitchen</b>", "<b>Store</b>"),
  showarrow = FALSE,
  font = list(size = 13, color = "#1F2933", family = "Arial")
)

fig <- fig %>% layout(
  xaxis = list(visible = FALSE, range = c(0.2, .9), fixedrange = TRUE),
  yaxis = list(visible = FALSE, range = c(0.2, 1),  fixedrange = TRUE),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 50, t = 10, b = 10),
  height = 250,
  width  = 400
)

fig
```
:::


::: {.column width="44%"}

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## System Evaluation

- **Production Wiesbaden:** $E(x^{Wi})=$ `r round(mean(cp_pool$x1),1)`
- **Production Mainz:** $E(x^{Ma})=$ `r round(mean(cp_pool$x2),1)`
- **Supply to Wiesbaden:** $E(x^{Wi, Wi}+x^{Ma, Wi})=$ `r round(mean(cp_pool$xA),1)`
- **Supply to Mainz:** $E(x^{Wi,Ma}+x^{Ma, Ma})=$ `r round(mean(cp_pool$xB),1)`
- **Utilisation Wiesbaden:** $\frac{E(x^{Wi})}{C^{Wi}}=$ `r round(mean(cp_pool$x1)/C1*100,1)`% 
- **Utilisation plant 2:** $\frac{E(x^{Ma})}{C^{Ma}}=$ `r round(mean(cp_pool$x2)/C2*100,1)`% 
- **Fill rate Wiesbaden:** $\frac{E(x^{Wi,Wi}+x^{Ma,Wi})}{E(d^{Wi})}=$ `r round(mean(cp_pool$xA)/mean(cp_pool$dA)*100,1)`% 
- **Fill rate Mainz:** $\frac{E(x^{Ma,Ma}+x^{Wi,Ma})}{E(d^{Ma})}=$  `r round(mean(cp_pool$xB)/mean(cp_pool$dB)*100,1)`% 
:::
:::

:::
::::

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}
The supply system behaves as if the separation between markets/products and plants were removed $\rightarrow$ total capacity and total demand are decisive
:::
:::

---

## Capacity pooling {#slide-cp-kpis}
::: slide-subtitle
KPI comparison
:::

```{r cp-kpi-compute}
#| echo: false
#| warning: false
#| message: false


cp_data <- cp_data[-13,]

tot.demand <- c(sum(cp_data$dA), sum(cp_data$dB))

kpi_df <- data.frame(
  Kennzahl = c(
    "Ø Production Wiesbaden (units)",
    "Ø Production Mainz (units)",
    "Ø Sales Wiesbaden (units)",
    "Ø Sales Mainz (units)",
    "Lost sales Wiesbaden (units)",
    "Lost sales Mainz (units)",
    "Utilisation Wiesbaden (%)",
    "Utilisation Mainz (%)",
    "Sales ratio Wiesbaden (%)",
    "Sales ratio Mainz (%)"
  ),
  Dezentral = c(
    round(mean(cp_data$xA), 1),
    round(mean(cp_data$xB), 1),
    round(mean(cp_data$xA), 1),
    round(mean(cp_data$xB), 1),
    sum(cp_data$dA)-sum(cp_data$xA),
    sum(cp_data$dB)-sum(cp_data$xB),
    round(mean(cp_data$xA/C1*100), 1),
    round(mean(cp_data$xB/C2*100), 1),
    round(sum(cp_data$xA)/tot.demand[1]*100, 1),
    round(sum(cp_data$xB)/tot.demand[2]*100, 1)
  ),
  "Pooling einseitig" = c(
    round(mean(cp_pool_1$x1), 1),
    round(mean(cp_pool_1$x2B), 1),
    round(mean(cp_pool_1$xA), 1),
    round(mean(cp_pool_1$xB), 1),
    sum(cp_pool_1$dA)-sum(cp_pool_1$xA),
    sum(cp_pool_1$dB)-sum(cp_pool_1$xB),
    round(mean(cp_pool_1$xA/C1*100), 1),
    round(mean(cp_pool_1$xB/C2*100), 1),
    round(sum(cp_pool_1$xA)/tot.demand[1]*100, 1),
    round(sum(cp_pool_1$xB)/tot.demand[2]*100, 1)
  ),
  "Pooling zweiseitig" = c(
    round(mean(cp_pool$x1), 1),
    round(mean(cp_pool$x2), 1),
    round(mean(cp_pool$xA), 1),
    round(mean(cp_pool$xB), 1),
    sum(cp_pool$dA)-sum(cp_pool$xA),
    sum(cp_pool$dB)-sum(cp_pool$xB),
    round(mean(cp_pool$xA/C1*100), 1),
    round(mean(cp_pool$xB/C2*100), 1),
    round(sum(cp_pool$xA)/tot.demand[1]*100, 1),
    round(sum(cp_pool$xB)/tot.demand[2]*100, 1)
  )
)
```

::: columns
::: {.column width="50%"}

```{r cp-kpi-table}
#| echo: false
#| warning: false
#| message: false

kbl(kpi_df,
    col.names = c("KPI", "Decentralised", "One-sided pooling", "Two-sided pooling"),
    align = c("l", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("condensed", "striped"),
                full_width = TRUE, font_size = 20) |>
  column_spec(1) |>
  column_spec(2, background = "#F3E7AE") |>
  column_spec(3, bold = TRUE, background = "#E8F4F1") |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)
```


:::
::: {.column width="48%"}

::: {.fragment .fade-up}

```{r cp-kpi-bar}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4

metriken <- c("Utilisation\n Wiesbaden (%)","Utilisation\n Mainz (%)", "Sales ratio\n Wiesbaden (%)", "Sales ratio\n Mainz (%)")
dez_vals  <- kpi_df$Dezentral[c(7:10)]
pool1_vals <- kpi_df$Pooling.einseitig[c(7:10)]
pool2_vals <- kpi_df$Pooling.zweiseitig[c(7:10)]

fig_kpi <- plot_ly()

fig_kpi <- fig_kpi %>% add_trace(
  x = metriken, y = dez_vals,
  type = "bar", name = "Decentralised",
  marker = list(color = "#E9C46A",
                line = list(color = "#B8951A", width = 1)),
  text = paste0(dez_vals, "%"),
  textposition = "outside",
  textfont = list(size = 13, family = "Arial", color = "#1F2933")
)

fig_kpi <- fig_kpi %>% add_trace(
  x = metriken, y = pool1_vals,
  type = "bar", name = "One-sided pooling",
  marker = list(color = cols$top,
                line = list(color = "#01696f", width = 1)),
  text = paste0(pool1_vals, "%"),
  textposition = "outside",
  textfont = list(size = 13, family = "Arial", color = "#1F2933")
)

fig_kpi <- fig_kpi %>% add_trace(
  x = metriken, y = pool2_vals,
  type = "bar", name = "Two-sided pooling",
  marker = list(color = cols$mid,
                line = list(color = "#a16a61", width = 1)),
  text = paste0(pool2_vals, "%"),
  textposition = "outside",
  textfont = list(size = 13, family = "Arial", color = "#1F2933")
)

fig_kpi <- fig_kpi %>% layout(
  barmode = "group",
  xaxis = list(title = "", fixedrange = TRUE),
  yaxis = list(title = "Value (%)", fixedrange = TRUE,
               range = c(90, 102)),
  legend = list(orientation = "h", y = -0.18),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 20, t = 30, b = 60),
  height = 370
)

fig_kpi
```

:::

:::
:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## Benefits of capacity pooling
Capacity pooling improves **all** service KPIs at once: more sales, higher utilisation, less idle capacity – without additional production capacity.
:::
:::

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}

## Costs of capacity pooling
However, capacity pooling requires additional logistics capacity and, above all, technological flexibility
:::
:::

---

## Capacity pooling {#slide-chaining}
::: slide-subtitle
Chaining
:::

::: {.callout-note appearance="minimal"}
**Chaining**

Capacity pooling enables capacities to be shared across different markets. However, it requires technological and logistical flexibility. The more markets and products that need to be served, the more complex and costly capacity pooling becomes. So the question is: how can capacity pooling be organised for multiple products and markets?
:::

::::: {.smaller-text}
:::: columns

::: {.column width="33%"}
::: {.fragment .fade-up}
**Full chaining**

```{r cp-full-chain1}
#| echo: false
#| warning: false
#| message: false
#| out-width: "95%"
#| fig-height: 3.2

draw_network <- function(edges, plants, markets, r = 0.055, range.x = c(0,1), range.y = c(0,1), width = 120, height = 320 ) {
  fig <- plot_ly(type = "scatter", mode = "markers")

  for (i in seq_len(nrow(edges))) {
    p  <- plants[edges$from[i], ]
    m  <- markets[edges$to[i], ]
    dx <- m$x - p$x; dy <- m$y - p$y; d <- sqrt(dx^2+dy^2)
    ax <- p$x + r*dx/d; ay <- p$y + r*dy/d
    ex <- m$x - r*dx/d; ey <- m$y - r*dy/d

    if (edges$dash[i]) {
      n_seg <- 12
      xs <- seq(ax, ex, length.out = n_seg+1)
      ys <- seq(ay, ey, length.out = n_seg+1)
      for (s in seq(1, n_seg-1, by=2)) {
        fig <- fig %>% add_segments(
          x = xs[s], xend = xs[s+1], y = ys[s], yend = ys[s+1],
          line = list(color="#4B4B4B", width=1.6),
          hoverinfo="skip", showlegend=FALSE
        )
      }
      fig <- fig %>% add_annotations(
        x=ex, y=ey, ax=xs[n_seg], ay=ys[n_seg],
        xref="x", yref="y", axref="x", ayref="y",
        showarrow=TRUE, arrowhead=2, arrowsize=0.9,
        arrowwidth=1.6, arrowcolor="#4B4B4B", text=""
      )
    } else {
      fig <- fig %>% add_annotations(
        x=ex, y=ey, ax=ax, ay=ay,
        xref="x", yref="y", axref="x", ayref="y",
        showarrow=TRUE, arrowhead=2, arrowsize=0.9,
        arrowwidth=2.0, arrowcolor="#222222", text=""
      )
    }
  }

  # Plant nodes (hexagon)
  fig <- fig %>% add_trace(
    x=plants$x, y=plants$y, text=plants$lbl, mode="markers+text",
    marker=list(symbol="hexagon", size=36,
                color="#BFDBF7", line=list(color="#4B5563",width=1.5)),
    textfont=list(size=14, color="#1F2933", family="Arial"),
    textposition="middle center", hoverinfo="skip", showlegend=FALSE
  )

  # Market nodes (circle)
  fig <- fig %>% add_trace(
    x=markets$x, y=markets$y, text=markets$lbl, mode="markers+text",
    marker=list(symbol="circle", size=36,
                color="#2D9E8E", line=list(color="#2F3E46",width=1.5)),
    textfont=list(size=14, color="white", family="Arial"),
    textposition="middle center", hoverinfo="skip", showlegend=FALSE
  )

  fig %>% layout(
    xaxis=list(visible=FALSE, range=range.x, fixedrange=TRUE),
    yaxis=list(visible=FALSE, range=range.y, fixedrange=TRUE),
    plot_bgcolor="transparent", paper_bgcolor="transparent",
    margin=list(l=10,r=10,t=30,b=10), height=height, width=width
  )
}


plants  <- data.frame(id = paste0("P",1:4), lbl = as.character(1:4),
                      x = 0.20, y = c(0.85,0.58,0.38,0.12))
markets <- data.frame(id = c("A","B","C","D"), lbl = c("A","B","C","D"),
                      x = 0.78, y = c(0.85,0.58,0.38,0.12))
rownames(plants)  <- plants$id
rownames(markets) <- markets$id

r <- 0.055

edges <- data.frame(
  from = c("P1","P2","P3","P4",   "P1","P2","P3","P4",   "P1","P2","P3","P4",   "P1","P2","P3","P4"),
  to   = c("A","B","C","D",       "B","C","D","A",       "C","D","A","B",       "D","A","B","C"),
  dash = c(rep(FALSE,4),          rep(TRUE,4),          rep(TRUE,4),          rep(TRUE,4)),
  stringsAsFactors = FALSE
)

draw_network(edges, plants, markets, range.y = c(.05,.925), width = 300, height = 300)

```

Links: $n \cdot n = 16$

Technologies per plant: $n=4$

Markets per plant: $n=4$
:::
:::

::: {.column width="33%"}

::: {.fragment .fade-up}
**Short chaining**  

```{r chaining-short}
#| echo: false
#| warning: false
#| message: false
#| out-width: "95%"
#| fig-height: 3.2

# Solid: 1->A, 2->B, 3->C, 4->D
# Dashed (short chain, next neighbour +1): 1->B, 2->C, 3->D, 4->A  (wrap-around)
# Dashed cross-down: 2->A, 3->B  etc. (Cachon: "short" = neighbor skip)
edges <- data.frame(
  from = c("P1","P2","P3","P4",   "P1","P2","P3","P4"),
  to   = c("A","B","C","D",       "B","A","D","C"),
  dash = c(rep(FALSE,4),          rep(TRUE,4)),
  stringsAsFactors = FALSE
)

draw_network(edges, plants, markets, range.y = c(.05,.925), width = 300, height = 300)
```

Links: $2 \cdot n = 8$

Technologies per plant: $n=2$

Markets per plant: $n=2$
:::
:::


::: {.column width="33%"}

::: {.fragment .fade-up}
**Long chaining**  

```{r chaining-long}
#| echo: false
#| warning: false
#| message: false
#| out-width: "95%"
#| fig-height: 3.2

# Long SC alternative: Solid: 1->A, 2->B, 3->C, 4->D
# Dashed (long chain, skip +2): 1->C, 2->D, 3->A, 4->B
edges_long <- data.frame(
  from = c("P1","P2","P3","P4",   "P1","P2","P3","P4"),
  to   = c("A","B","C","D",       "B","C","D","A"),
  dash = c(rep(FALSE,4),          rep(TRUE,4)),
  stringsAsFactors = FALSE
)

draw_network(edges_long, plants, markets, range.y = c(.05,.925), width = 300, height = 300)
```

Links: $2 \cdot n = 8$

Technologies per plant: $n=2$

Markets per plant: $n=4$

:::
:::
::::
:::::

---

## Dealing with uncertainty {#slide-inv-pooling-intro}
::: slide-subtitle
Pooling over time - safety stocks 
:::

::: {.callout-note appearance="minimal" icon=false}

## Definition of safety stock (recap)

Safety stocks buffer demand fluctuations over time. Let $S$ be the initial inventory at the start of the risk period and $\mu$ the expected demand in the risk period; then the safety stock $sb$ is: $sb = S - \mu$

Example: the Wiesbaden-Erbenheim central plant periodically orders crates of fresh salmon fillet from the fish wholesaler.
:::

:::: columns
::: {.column width="33%"}

::: {.fragment .fade-up}
::: {.callout-note appearance="minimal" icon=false}

## Setting safety stocks

- *by cost*: safety stocks incur inventory holding costs and avoid shortage costs $\rightarrow$ cost trade-off 
- *Service readiness*: protection against the risk of being unable to deliver during the risk period ($\alpha$ service level)
- *Fill rate*: share of demand to be met within the risk period in the long run ($\beta$ service level)
:::
:::
:::


::: {.column width="65%"}

::: {.fragment .fade-up}

```{r ts-lager-graph}
#| echo: false
#| warning: false
#| message: false
#| out-width: "95%"
#| fig-height: 6

set.seed(42)
n   <- 24        # periods
lam <- 20        # mean demand (Poisson)
t   <- 4         # review interval
S   <- 135       # Order-up-to-Level
wbz <- 2         # replenishment lead time (periods)
l.ini <- 120     # initial inventory
sb <- S - (t + wbz) * lam

demand <- rpois(n, lambda = lam)

# --- Simulation ---
res <- tS.lager(d = demand, t = t, S = S, l.ini = l.ini, wbz = wbz)
res <- as.data.frame(res)

# --- Subset: periods 2 to 22 only ---
idx      <- 2:22
perioden <- seq_len(nrow(res))[idx]
res_sub  <- res[idx, ]

# --- Double arrows: length 3.95, placed side by side without gaps ---
arrow_len <- t - 0.05          # 3.95 (slightly shorter than review interval t=4)
x_start   <- perioden[1]
x_end     <- perioden[length(perioden)]

# Arrow start points: without gaps from x_start
arrow_x0 <- seq(x_start, x_end - arrow_len + 0.01, by = arrow_len)
arrow_x1 <- arrow_x0 + arrow_len

# y position of the arrows: just below the zero line
y_min_data <- min(pmin(0, res_sub$Lagerbestand), na.rm = TRUE)
y_arrow    <- y_min_data - abs(y_min_data) * 0.15 - 5

fig <- plot_ly()

# Inventory level
fig <- fig |> add_lines(
  x = perioden, y = res_sub$Lagerbestand,
  line = list(shape = "hvh"), name = "Inventory level"
)

# Safety stock
fig <- fig |> add_lines(
  x = perioden, y = sb,
  name = "Safety stock"
)

# Shortage fill
fig <- fig |> add_trace(
  x = perioden, y = pmin(0, res_sub$Lagerbestand),
  type = "scatter", mode = "lines",
  line = list(shape = "hvh", color = "transparent"),
  name = "Shortages",
  fill = "tozeroy", fillcolor = "#FF000050"
)

# --- Draw double arrows ---
for (i in seq_along(arrow_x0)) {
  x0 <- arrow_x0[i]
  x1 <- arrow_x1[i]
  xm <- (x0 + x1) / 2

  # Horizontal line (shaft of the double arrow)
  fig <- fig |> add_segments(
    x = x0, xend = x1,
    y = y_arrow, yend = y_arrow,
    line = list(color = "#333333", width = 1.5),
    showlegend = FALSE, hoverinfo = "skip"
  )

  # Left arrowhead (points left)
  fig <- fig |> add_annotations(
    x = x0, y = y_arrow,
    ax = x0 + 0.6, ay = y_arrow,
    xref = "x", yref = "y", axref = "x", ayref = "y",
    showarrow = TRUE, arrowhead = 2, arrowsize = 0.9,
    arrowwidth = 1.5, arrowcolor = "#333333", text = ""
  )

  # Right arrowhead (points right)
  fig <- fig |> add_annotations(
    x = x1, y = y_arrow,
    ax = x1 - 0.6, ay = y_arrow,
    xref = "x", yref = "y", axref = "x", ayref = "y",
    showarrow = TRUE, arrowhead = 2, arrowsize = 0.9,
    arrowwidth = 1.5, arrowcolor = "#333333", text = ""
  )

  # Label "Period" centred below the arrow
  fig <- fig |> add_annotations(
    x = xm, y = y_arrow,
    xref = "x", yref = "y",
    text = "Period", showarrow = FALSE,
    yshift = -16,
    font = list(size = 11, color = "#333333", family = "Arial")
  )
}

fig <- fig |> layout(
  legend = list(
    orientation = "h",
    x           = 0.5,
    y           = 1.12,
    xanchor     = "center",
    yanchor     = "top"
  ),
  xaxis = list(
    showticklabels = FALSE,     # no numbers on the x-axis
    showgrid       = FALSE,
    zeroline       = FALSE,
    range          = c(x_start - 0.3, x_end + 0.3)
  ),
  yaxis = list(
    title = "Inventory",
    range = c(y_arrow - 10, NA)  # extend y-axis downwards for arrows
  ),
  margin = list(b = 55),           # margin for labels below
  height= 550,
  width = 1000
)

fig
```

:::
:::
::::


---

## Safety stocks  {#slide-alpha-empirisch}
::: slide-subtitle
$\alpha$ service level and empirical distribution function
:::

::: {.callout-note appearance="minimal" icon=false}
Assume that a stockout risk of 5% is acceptable, so the $\alpha$ service level is 95%.
*Question*: which (minimum) initial inventory ensures this service level?
*Answer*:  Formally, $P(d \leq S) \geq \alpha$ must hold. For an empirical demand distribution $F(d)$ we therefore need $\hat{F}(S) \geq \alpha$ or $S \geq \hat{F}^{-1}(\alpha)$. So we are looking for an **(empirical) quantile**.
:::

:::: columns
::: {.column width="25%"}

```{r alpha-ecdf-table}
#| echo: false
#| warning: false
#| message: false

dA <- cp_data$dA
mu_A <- mean(dA)
dA_sorted <- unique(sort(dA))
dA_freq <- table(dA)
n_obs <- length(dA)
F_emp <- cumsum(dA_freq)/ n_obs
F_emp_pct <- paste0(round(F_emp * 100, 1), "%")

# Highlight α thresholds (75%, 90%, 95%)
alpha_rows <- sapply(c(0.75, 0.90, 0.95), function(a) which(F_emp >= a)[1])

ecdf_df <- data.frame(
  Rang = seq_len(length(dA_freq)),
  d    = dA_sorted,
  H    = as.numeric(dA_freq),
  F    = F_emp_pct
)

tbl <- kbl(ecdf_df,
    col.names = c("Rank", "d<sub>t</sub>", "Count", "F̂(d)"),
    align = c("c", "r","r", "r"),
    escape = FALSE,
    caption = paste0("μ = ", round(mu_A, 1), " | n = ", n_obs)) |>
  kable_styling(
    bootstrap_options = c("condensed", "striped"),
    full_width = TRUE,
    font_size = 22
  ) |>
  column_spec(1, color = "gray") |>
  column_spec(2, bold = TRUE) |>
  column_spec(3, bold = TRUE) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE)

# Colour-highlight α rows
tbl <- tbl |>
  row_spec(alpha_rows[1], background = "#FFF3CD", bold = TRUE) |>  # α=75%
  row_spec(alpha_rows[2], background = "#CCE5FF", bold = TRUE) |>  # α=90%
  row_spec(alpha_rows[3], background = "#FFD6CC", bold = TRUE)    # α=95%

tbl
```

:::

::: {.column width="74%"}
::: {.fragment .fade-up}
```{r alpha-ecdf-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4.5

mu_A <- mean(dA)
dA_sorted <- sort(dA)
n_obs <- length(dA_sorted)
F_emp <- seq_len(n_obs) / n_obs

# Correct right-continuous step function
x_ecdf <- c(min(dA_sorted) - 6, min(dA_sorted))
y_ecdf <- c(0, 0)
for (i in seq_len(n_obs)) {
  x_ecdf <- c(x_ecdf, dA_sorted[i], dA_sorted[i])
  y_ecdf <- c(y_ecdf, if (i == 1) 0 else F_emp[i - 1], F_emp[i])
  if (i < n_obs) {
    x_ecdf <- c(x_ecdf, dA_sorted[i], dA_sorted[i + 1])
    y_ecdf <- c(y_ecdf, F_emp[i], F_emp[i])
  }
}
x_ecdf <- c(x_ecdf, dA_sorted[n_obs], dA_sorted[n_obs] + 6)
y_ecdf <- c(y_ecdf, 1, 1)

alpha_show  <- c(0.75, 0.90, 0.95)
alpha_cols  <- c("#E9A000", "#2196F3", "#E76F51")
alpha_labels <- c("α = 75%", "α = 90%", "α = 95%")

fig <- plot_ly()

# ECDF steps
fig <- fig %>% add_trace(
  x = x_ecdf, y = y_ecdf,
  type = "scatter", mode = "lines",
  name = "F̂(d)",
  line = list(color = cols$top, width = 2.8),
  hovertemplate = "d = %{x}<br>F̂(d) = %{y:.3f}<extra></extra>"
)

# Observation points
fig <- fig %>% add_trace(
  x = dA_sorted, y = F_emp,
  type = "scatter", mode = "markers",
  name = "Observations",
  marker = list(color = cols$top, size = 9,
                line = list(color = "white", width = 1.5)),
  hovertemplate = "d = %{x}<br>F̂ = %{y:.3f}<extra></extra>"
)

# α lines + reading lines
for (j in seq_along(alpha_show)) {
  a     <- alpha_show[j]
  idx_a <- which(F_emp >= a)[1]
  S_a   <- dA_sorted[idx_a]
  sb_a  <- round(S_a - mu_A, 1)
  col_a <- alpha_cols[j]

  fig <- fig %>% add_trace(
    x = c(min(dA_sorted) - 6, S_a), y = c(a, a),
    type = "scatter", mode = "lines",
    name = alpha_labels[j],
    line = list(color = col_a, width = 1.8, dash = "dot"),
    hoverinfo = "skip", showlegend = TRUE
  )
  fig <- fig %>% add_trace(
    x = c(S_a, S_a), y = c(0, a),
    type = "scatter", mode = "lines",
    line = list(color = col_a, width = 1.8, dash = "dot"),
    hoverinfo = "skip", showlegend = FALSE
  )
  fig <- fig %>% add_annotations(
    x = S_a, y = -0.045,
    xref = "x", yref = "paper",
    text = paste0("<b>S=", S_a, "<br>sb=", sb_a, "</b>"),
    showarrow = FALSE,
    font = list(size = 16, color = col_a, family = "Arial"),
    xanchor = "center"
  )
}

# μ line
fig <- fig %>% add_trace(
  x = c(mu_A, mu_A), y = c(0, 1),
  type = "scatter", mode = "lines",
  name = paste0("μ = ", round(mu_A, 1)),
  line = list(color = "#6B7280", width = 1.4, dash = "dashdot"),
  hoverinfo = "skip"
)

fig <- fig %>% layout(
  xaxis = list(
    title = "Demand d<sub>t</sub><sup>Wi</sup>",
    fixedrange = TRUE,
    range = c(min(dA_sorted) - 7, max(dA_sorted) + 7)
  ),
  yaxis = list(
    title = "F̂(d)",
    fixedrange = TRUE,
    range = c(-0.05, 1.08),
    tickformat = ".0%"
  ),
  #legend = list(orientation = "h", y = -0.26, font = list(size = 16)),
  legend = list(
    orientation = "v",
    x           = 0.99,
    y           = 0.75,
    xanchor     = "right",
    yanchor     = "top"
  ),
  
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 15, t = 10, b = 85),
  height = 420,
  width = 1100
)

fig
```
:::

::: {.fragment .fade-up}
::: {.callout-note appearance="minimal" icon=false}
Safety stocks $sb=S-\mu$: $sb_{75\%}=4$, $sb_{90\%}=7$, $sb_{95\%}=15$
:::
:::

:::
::::

---

## Inventory pooling {#slide-dezentral-zentral-empirisch}
::: slide-subtitle
Safety stocks with decentralised and centralised inventory holding 
:::

:::: columns

<!-- ══════════════════ LEFT COLUMN: DECENTRALISED ══════════════════ -->
::: {.column width="50%"}

**Decentralised network**

```{r alpha-decentral-network}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 1.75

plants  <- data.frame(id = 1:2, lbl = as.character(1:2),
                      x = 0.20, y = c(0.85,0.48))
markets <- data.frame(id = c("A","B"), lbl = c("A","B"),
                      x = 0.78, y = c(0.85,0.48))
rownames(plants)  <- plants$id
rownames(markets) <- markets$id

r <- 0.055

edges <- data.frame(
  from = c("1","2"),
  to   = c("A","B"),
  dash = c(rep(FALSE,2)),
  stringsAsFactors = FALSE
)

draw_network(edges, plants, markets, range.y = c(.35,1), width = 300, height = 150)

```

```{r alpha-ecdf-dec}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3.75

dA <- cp_data$dA
dB <- cp_data$dB
mu_A <- mean(dA); mu_B <- mean(dB)

# Helper function: ECDF step points
make_ecdf_xy <- function(d) {
  ds <- sort(d); n <- length(ds); Fe <- seq_len(n) / n
  x <- c(min(ds)-5, min(ds)); y <- c(0, 0)
  for (i in seq_len(n)) {
    x <- c(x, ds[i], ds[i])
    y <- c(y, if(i==1) 0 else Fe[i-1], Fe[i])
    if (i < n) { x <- c(x, ds[i], ds[i+1]); y <- c(y, Fe[i], Fe[i]) }
  }
  x <- c(x, ds[n], ds[n]+5); y <- c(y, 1, 1)
  list(x=x, y=y, sorted=ds, F=Fe)
}

eA <- make_ecdf_xy(dA)
eB <- make_ecdf_xy(dB)

alpha <- 0.95
S_A <- eA$sorted[which(eA$F >= alpha)][1]
S_B <- eB$sorted[which(eB$F >= alpha)][1]
S_mid <- (S_A + S_B)/2

sb_A <- round(S_A - mu_A, 1)
sb_B <- round(S_B - mu_B, 1)
col_A <- cols$top     # Market A — main colour
col_B <- cols$mid     # Market B — secondary colour

fig_dec2 <- plot_ly()

# ECDF market A
fig_dec2 <- fig_dec2%>% add_trace(
  x=eA$x, y=eA$y, type="scatter", mode="lines",
  name="F̂<sub>WI</sub>(d)", line=list(color=col_A, width=2.5),
  hovertemplate="d=%{x}<br>F̂̂<sub>WI</sub>=%{y:.2f}<extra></extra>"
) %>% add_trace(
  x=eA$sorted, y=eA$F, type="scatter", mode="markers",
  showlegend=FALSE, marker=list(color=col_A, size=7,
  line=list(color="white", width=1)), hoverinfo="skip"
)

# ECDF market B
fig_dec2 <- fig_dec2%>% add_trace(
  x=eB$x, y=eB$y, type="scatter", mode="lines",
  name="F̂<sub>Ma</sub>(d)", line=list(color=col_B, width=2.5, dash="dash"),
  hovertemplate="d=%{x}<br>F̂̂<sub>Ma</sub>=%{y:.2f}<extra></extra>"
) %>% add_trace(
  x=eB$sorted, y=eB$F, type="scatter", mode="markers",
  showlegend=FALSE, marker=list(color=col_B, size=7,
  line=list(color="white", width=1)), hoverinfo="skip"
)

# α = 95% line
fig_dec2 <- fig_dec2%>% add_trace(
  x=c(min(c(dA,dB))-6, max(S_A, S_B)), y=c(alpha, alpha),
  type="scatter", mode="lines", name="α = 95%",
  line=list(color="#E76F51", width=1.6, dash="dot"), hoverinfo="skip"
)
# Vertical reading lines
for (vals in list(list(S=S_A, mu=mu_A, sb=sb_A, col=col_A, lbl="A"),
                  list(S=S_B, mu=mu_B, sb=sb_B, col=col_B, lbl="B"))) {
  fig_dec2 <- fig_dec2%>%
    add_trace(x=c(vals$S, vals$S), y=c(0, alpha), type="scatter", mode="lines",
              line=list(color=vals$col, width=1.3, dash="dot"),
              hoverinfo="skip", showlegend=FALSE) %>%
    add_annotations(x=ifelse(vals$S < S_mid, vals$S*.975,  vals$S*1.025) , y=0.05, xref="x", yref="paper",
      text=paste0("<b>S<sub>",vals$lbl,"</sub>=",vals$S,"<br>sb<sub>",vals$lbl,"</sub>=",vals$sb,"</b>"),
      showarrow=FALSE, font=list(size=15, color=vals$col, family="Arial"),
      xanchor="center")
}
# μ lines
for (vals in list(list(mu=mu_A, col=col_A, nm=paste0("μ<sub>Wi</sub>=",round(mu_A,1))),
                  list(mu=mu_B, col=col_B, nm=paste0("μ<sub>Ma</sub>=",round(mu_B,1))))) {
  fig_dec2 <- fig_dec2%>% add_trace(
    x=c(vals$mu, vals$mu), y=c(0,1), type="scatter", mode="lines",
    name=vals$nm, line=list(color=vals$col, width=1.2, dash="dashdot"),
    hoverinfo="skip"
  )
}

fig_dec2 <- fig_dec2%>% layout(
  xaxis=list(title="Demand d", fixedrange=F,
             range=c(min(c(dA,dB))-8, max(S_A,S_B)+8)),
  yaxis=list(title="F̂(d)", fixedrange=F, range=c(-0.06,1.08),
             tickformat=".0%"),
  #legend=list(orientation="h", y=-0.28, font=list(size=10)),
  legend = list(
    orientation = "v",
    x           = 0.99,
    y           = 0.75,
    xanchor     = "right",
    yanchor     = "top",
    size = 13
  ),
  plot_bgcolor="transparent", paper_bgcolor="transparent",
  margin=list(l=50, r=10, t=5, b=85), height=350, width = 750
)
fig_dec2
```

:::

<!-- ══════════════════ RIGHT COLUMN: CENTRALISED ══════════════════ -->
::: {.column width="50%"}

::: {.fragment .fade-up}

**Centralised network (pooling)**

```{r alpha-central-network}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 1.75

plants  <- data.frame(id = 0, lbl = as.character(0),
                      x = 0.20, y = c(0.61))
markets <- data.frame(id = c("A","B"), lbl = c("A","B"),
                      x = 0.78, y = c(0.85,0.48))
rownames(plants)  <- plants$id
rownames(markets) <- markets$id

r <- 0.055

edges <- data.frame(
  from = c("0"),
  to   = c("A","B"),
  dash = c(rep(FALSE,2)),
  stringsAsFactors = FALSE
)

draw_network(edges, plants, markets, range.y = c(.35,1), width = 300, height = 150)

```

```{r alpha-ecdf-central}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3.75

dTotal <- cp_data$dA + cp_data$dB
mu_T   <- mean(dTotal)
eT     <- make_ecdf_xy(dTotal)

S_T  <- eT$sorted[which(eT$F >= alpha)][1]
sb_T <- round(S_T - mu_T, 1)
col_T <- "#F4A261"  # central warehouse colour (orange)

fig_cen2 <- plot_ly()

# ECDF total demand
fig_cen2 <- fig_cen2%>% add_trace(
  x=eT$x, y=eT$y, type="scatter", mode="lines",
  name="F̂<sub>A+B</sub>(d)", line=list(color=col_T, width=2.8),
  hovertemplate="d=%{x}<br>F̂=%{y:.2f}<extra></extra>"
) %>% add_trace(
  x=eT$sorted, y=eT$F, type="scatter", mode="markers",
  showlegend=FALSE, marker=list(color=col_T, size=7,
  line=list(color="white", width=1)), hoverinfo="skip"
)

# α = 95% line
fig_cen2 <- fig_cen2%>% add_trace(
  x=c(min(dTotal)-8, S_T), y=c(alpha, alpha),
  type="scatter", mode="lines", name="α = 95%",
  line=list(color="#E76F51", width=1.6, dash="dot"), hoverinfo="skip"
) %>% add_trace(
  x=c(S_T, S_T), y=c(0, alpha), type="scatter", mode="lines",
  line=list(color=col_T, width=1.3, dash="dot"),
  hoverinfo="skip", showlegend=FALSE
) %>% add_annotations(
  x=S_T, y=0.05, xref="x", yref="paper",
  text=paste0("<b>S<sub>0</sub>=",S_T,"<br>sb<sub>0</sub>=",sb_T,"</b>"),
  showarrow=FALSE, font=list(size=15, color=col_T, family="Arial"),
  xanchor="center"
) %>% add_trace(
  x=c(mu_T, mu_T), y=c(0,1), type="scatter", mode="lines",
  name=paste0("μ<sub>Wi+Ma</sub>=",round(mu_T,1)),
  line=list(color="#6B7280", width=1.2, dash="dashdot"), hoverinfo="skip"
)

fig_cen2 <- fig_cen2%>% layout(
  xaxis=list(title="Total demand d<sup>Wi</sup>+d<sup>Ma</sup>",
             fixedrange=TRUE, range=c(min(dTotal)-10, S_T+12)),
  yaxis=list(title="F̂(d)", fixedrange=TRUE, range=c(-0.06,1.08),
             tickformat=".0%"),
#  legend=list(orientation="h", y=-0.28, font=list(size=10)),
legend = list(
    orientation = "v",
    x           = 0.99,
    y           = 0.75,
    xanchor     = "right",
    yanchor     = "top",
    size = 13
  ),
  plot_bgcolor="transparent", paper_bgcolor="transparent",
  margin=list(l=50, r=10, t=5, b=85), height=350, width = 750
)
fig_cen2
```

:::
:::

::::

::: {.fragment .fade-up}
::: {.callout-note appearance="minimal" icon="info"}
Safety stock savings: $sb_{decentral}=sb_{A}+sb_{B}=27.5$ vs. $sb_{0}=22.5$  $\rightarrow$ 5 units or 18%
:::
:::

## Inventory pooling {#slide-dezentral-zentral-parameterisch}
::: slide-subtitle
Safety stocks with a known demand distribution
:::

::: {.callout-note appearance="minimal" icon=false}

## Demand distribution

It is often assumed that demand for goods in a given time span is a random variable whose distribution parameters (such as mean and variance) are known or can be estimated. Real data often deviate from this — simulated using the example of **Frankfurt/Offenbach** with occasional, shared holiday peaks.

:::


```{r}
#| label: sushifresh-sim-data-prep
#| echo: false
#| warning: false
#| message: false

# ── Simulated daily demand Frankfurt & Offenbach ──────────────────────────────
# (reproducible via set.seed(); replaces real data with synthetic series
#  with occasional, shared "holiday peaks")

set.seed(616)
n_tage <- 250

dA <- pmax(30, round(rnorm(n_tage, mean = 280, sd = 55 * 0.7)))   # Frankfurt
dB <- pmax(10, round(rnorm(n_tage, mean = 95,  sd = 22 * 0.7)))   # Offenbach

spitzentage <- sample(seq_len(n_tage), size = 6)
dA[spitzentage] <- dA[spitzentage] + round(runif(6, 220, 420))
dB[spitzentage] <- dB[spitzentage] + round(runif(6, 70, 150))

prod.names <- c("Frankfurt", "Offenbach")
dpool <- dA + dB

# ── Estimate parameters ───────────────────────────────────────────────────────
muA  <- mean(dA, na.rm = TRUE);  sigA <- sd(dA, na.rm = TRUE)
muB  <- mean(dB, na.rm = TRUE);  sigB <- sd(dB, na.rm = TRUE)
mupool  <- mean(dpool, na.rm = TRUE);  sigpool <- sd(dpool, na.rm = TRUE)

n_perioden <- length(dA)
```


::: columns
::: {.column width="52%"}

```{r}
#| label: sushifresh-sim-ecdf-fit
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4.2

# ── Colours ───────────────────────────────────────────────────────────────────
colA <- cols$top
colB <- cols$atp
colpool <- cols$dash

# ── Helper function: ECDF step points ─────────────────────────────────────────
make_ecdf_xy <- function(d) {
  d <- d[!is.na(d)]
  ds <- sort(d); n <- length(ds); Fe <- seq_len(n) / n
  x <- c(min(ds) - 0.08 * diff(range(ds)), min(ds))
  y <- c(0, 0)
  for (i in seq_len(n)) {
    x <- c(x, ds[i], ds[i])
    y <- c(y, if (i == 1) 0 else Fe[i - 1], Fe[i])
    if (i < n) { x <- c(x, ds[i], ds[i + 1]); y <- c(y, Fe[i], Fe[i]) }
  }
  x <- c(x, ds[n], ds[n] + 0.08 * diff(range(ds)))
  y <- c(y, 1, 1)
  list(x = x, y = y, sorted = ds, F = Fe)
}

eA <- make_ecdf_xy(dA);  eB <- make_ecdf_xy(dB); ePool <- make_ecdf_xy(dpool)

# ── Normal distribution CDF curves ────────────────────────────────────────────
xgridA <- seq(min(dA, na.rm=TRUE) - 200, max(dA, na.rm=TRUE) + 200, length.out = 300)
xgridB <- seq(min(dB, na.rm=TRUE) - 200, max(dB, na.rm=TRUE) + 200, length.out = 300)
xgridpool <- seq(min(dpool, na.rm=TRUE) - 5000, max(dpool, na.rm=TRUE) + 200, length.out = 300)

# ── KS test ───────────────────────────────────────────────────────────────────
ksA <- ks.test(dA, "pnorm", muA, sigA)
ksB <- ks.test(dB, "pnorm", muB, sigB)
kspool <- ks.test(dpool, "pnorm", mupool, sigpool)

fmt_p <- function(p) if (p < 0.001) "<0.001" else paste0("= ", round(p, 3))

# ── Plot ──────────────────────────────────────────────────────────────────────
fig <- plotly::plot_ly()

fig <- fig |>
  plotly::add_trace(x = eA$x, y = eA$y, type = "scatter", mode = "lines",
            name = paste0("ECDF ",prod.names[1]),
            line = list(color = colA, width = 2.2),
            marker = list(color = colA, size = 6,
                          line = list(color = "white", width = 1))) |>
  plotly::add_trace(x = xgridA, y = pnorm(xgridA, muA, sigA),
            type = "scatter", mode = "lines",
            name = paste0("N(", round(muA), ", ", round(sigA), "²)"),
            line = list(color = colA, width = 2.2, dash = "dot")) 
  


fig <- fig |>
  plotly::add_trace(x = eB$x, y = eB$y, type = "scatter", mode = "lines+markers",
            name = paste0("ECDF ",prod.names[2]),
            line = list(color = colB, width = 2.2),
            marker = list(color = colB, size = 6,
                          line = list(color = "white", width = 1))) |>
  plotly::add_trace(x = xgridB, y = pnorm(xgridB, muB, sigB),
            type = "scatter", mode = "lines",
            name = paste0("N(", round(muB), ", ", round(sigB), "²)"),
            line = list(color = colB, width = 2.2, dash = "dot"))


# fig <- fig |>  
#   plotly::add_trace(x = ePool$x, y = ePool$y, type = "scatter", mode = "lines+markers",
#             name = "ECDF total demand",
#             line = list(color = colpool, width = 2.2), 
#             marker = list(color = colpool, size = 6, line = list(color = "white", width = 1))) |>
#   plotly::add_trace(x = xgridpool, y = pnorm(xgridpool, mupool, sigpool),
#             type = "scatter", mode = "lines",
#             name = paste0("N(", round(mupool), ", ", round(sigpool), "²) "),
#             line = list(color = colpool, width = 2.2, dash = "dot"))

fig <- fig |>
  plotly::add_annotations(x = 0.02, y = 0.93, xref = "paper", yref = "paper",
    text = paste0("<b>KS test Frankfurt:</b> p ", fmt_p(ksA$p.value)),
    showarrow = FALSE, font = list(size = 16, color = colA, family = "Arial"),
    xanchor = "left") |>
  plotly::add_annotations(x = 0.02, y = 0.83, xref = "paper", yref = "paper",
    text = paste0("<b>KS test Offenbach:</b> p ", fmt_p(ksB$p.value)),
    showarrow = FALSE, font = list(size = 16, color = colB, family = "Arial"),
    xanchor = "left") 
  # plotly::add_annotations(x = 0.02, y = 0.73, xref = "paper", yref = "paper",
  #   text = paste0("<b>KS test total demand:</b> p ", fmt_p(kspool$p.value)),
  #   showarrow = FALSE, font = list(size = 16, color = colpool, family = "Arial"),
  #   xanchor = "left")

fig |> plotly::layout(
  xaxis = list(title = "Daily demand (sushi boxes)", fixedrange = FALSE,
               tickformat = ",.0f", range = c(-100, 800), font = list(size = 14)),
  yaxis = list(title = "F(d)", fixedrange = TRUE,
               range = c(-0.05, 1.08), tickformat = ".0%", font = list(size = 14)),
  legend = list(orientation = "h", y = -0.28, font = list(size = 18)),
  plot_bgcolor  = "transparent", paper_bgcolor = "transparent",
  margin = list(l = 55, r = 15, t = 10, b = 90),
  height = 390, width = 750
)
```

:::
::: {.column width="46%"}

::: {.callout-note appearance="minimal" icon=false}
**Data basis:** simulated daily demand over 250 days
[(set.seed(616), see R code for details)]{.smaller}


- **ECDF** (solid) = empirical distribution
- **Normal distribution** (dotted) = parametric model
:::

::: {.fragment .fade-up}

```{r}
#| label: sushifresh-sim-fit-table
#| echo: false
#| warning: false
#| message: false

fitdf <- data.frame(
  Standort = c("Frankfurt", "Offenbach"),
  mu       = round(c(muA, muB)),
  sigma    = round(c(sigA, sigB)),
  KS_p     = c(round(ksA$p.value, 3), round(ksB$p.value, 3)),
  SB_95    = round(c(quantile(dA, 0.95) , quantile(dB, 0.95))),
  pSB_95    = round(c( qnorm(.95, muA, sigA), qnorm(.95, muB, sigB)))
)

kableExtra::kbl(fitdf,
    col.names = c("Store", "μ̂", "σ̂", "KS p-value", "emp. SS (α=95%)","param. SS"),
    align = c("l", "r", "r", "r", "r","r"), escape = FALSE,
    format.args = list(big.mark = ",")) |>
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "striped"),
    full_width = TRUE, font_size = 18) |>
  kableExtra::row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  #kableExtra::row_spec(3, bold = TRUE, background = "#E8F4F1") |>
  kableExtra::column_spec(5, bold = TRUE, color = "#02818a")
```

:::
:::
:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}
The shared holiday peaks produce a right-skewed distribution in **both** stores that deviates clearly from the normal distribution (KS test: p < 0.01 for both) — accordingly, the empirical safety stock turns out lower than the parametric one.
:::
:::


---

## Inventory pooling {#slide-pooling-normal-intro}
::: slide-subtitle
Normally distributed demand – safety stock for two sites
:::

::: {.callout-note appearance="minimal" icon=false}
Let $D_1 \sim \mathcal{N}(\mu_1, \sigma_1^2)$ and $D_2 \sim \mathcal{N}(\mu_2, \sigma_2^2)$ be the **normally distributed demands** at two sites with correlation $\rho = \text{Cor}(D_1, D_2)$.

The safety stock at site $i$ for an $\alpha$ service level is: $sb_i = z_\alpha \cdot \sigma_i$, so that in total: $sb_{dec} = z_\alpha \cdot (\sigma_1 + \sigma_2)$
:::

::: columns
::: {.column width="50%"}

::: {.fragment .fade-up}
::: {.callout-note appearance="simple"}

## Central warehouse: distribution of total demand

The aggregated demand $D_0 = D_1 + D_2$ is also normally distributed:

$$D_0 \sim \mathcal{N}\!\left(\mu_1 + \mu_2,\; \sigma_1^2 + \sigma_2^2 + 2\rho\,\sigma_1\sigma_2\right)$$

Hence the central safety stock is:

$$sb_{cen} = z_\alpha \cdot \sqrt{\sigma_1^2 + \sigma_2^2 + 2\rho\,\sigma_1\sigma_2}$$
:::
:::

:::
::: {.column width="48%"}

::: {.fragment .fade-up}
::: smaller-text
**Comparison: decentralised vs. centralised**

| Configuration | Safety stock |
|---|---|
| Decentralised (2 warehouses) | $z_\alpha \cdot (\sigma_1 + \sigma_2)$ |
| Centralised ($\rho = -1$) | $z_\alpha \cdot \|\sigma_1 - \sigma_2\|$ |
| Centralised ($\rho = 0$) | $z_\alpha \cdot \sqrt{\sigma_1^2 + \sigma_2^2}$ |
| Centralised ($\rho = +1$) | $z_\alpha \cdot (\sigma_1 + \sigma_2)$ |
$\rightarrow$ $\|\sigma_1 - \sigma_2\| \leq \sqrt{\sigma_1^2 + \sigma_2^2} \leq \sigma_1 + \sigma_2$
$\rightarrow$ **Pooling pays off as long as $\rho < 1$**
:::
:::

:::
:::

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}
The **smaller** the correlation $\rho$, the greater the savings potential from central pooling. (**risk-balancing effect**)
:::
:::

---

## Key result {#slide-sushifresh-kernergebnis}
::: slide-subtitle
Safety stock for all six SushiFresh stores ($\alpha = 95\%$)
:::

```{r sushifresh-kernergebnis-berechnung}
#| echo: false
#| warning: false
#| message: false

sf_mu    <- c(120, 150, 280, 90, 110, 95)
sf_sigma <- c(25, 30, 55, 20, 24, 22)
z_sf     <- qnorm(0.95)

sb_dez_sf <- z_sf * sum(sf_sigma)                     # decentralised: sum of individual SS
sb_zen_sf <- z_sf * sqrt(sum(sf_sigma^2))              # centralised: square root of sum of variances (rho=0)
einspar_sf <- round((1 - sb_zen_sf / sb_dez_sf) * 100, 1)

# Idealised comparison: 6 identical stores with average sigma
sigma_mean_sf <- mean(sf_sigma)
ideal_sf <- pooling_safety_stock(n = 6, sigma = sigma_mean_sf, z = z_sf)
```

::: columns
::: {.column width="55%"}

::: {.callout-note appearance="simple"}

## Verified with `qnorm()`

- $z_{0.95} =$ `r round(z_sf,3)`
- **Decentralised:** $sb_{dec}=z_\alpha\sum\sigma_i=$ `r round(sb_dez_sf,1)` boxes
- **Centralised (pooled, independent):** $sb_{cen}=z_\alpha\sqrt{\sum\sigma_i^2}=$ `r round(sb_zen_sf,1)` boxes
- **Savings from consolidation:** `r einspar_sf`%
:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}
**Cross-check** `pooling_safety_stock()`: for 6 *identical* stores with $\bar\sigma=$ `r round(sigma_mean_sf,1)` the savings would be `r round(ideal_sf$savings_pct)`% — **more** than the actual `r einspar_sf`%. The dominant, volatile Frankfurt site ($\sigma=55$) reduces the relative pooling advantage compared with the idealised case of equally sized sites.
:::
:::

:::
::: {.column width="43%"}

```{r sushifresh-kernergebnis-bar}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3.6

plot_ly() |>
  add_trace(
    x = c("Decentralised", "Centralised (Wiesbaden-Erbenheim)"),
    y = c(sb_dez_sf, sb_zen_sf),
    type = "bar",
    marker = list(color = c("#E9C46A", cols$top),
                  line = list(color = c("#B8951A", "#01696f"), width = 1)),
    text = paste0(round(c(sb_dez_sf, sb_zen_sf), 1), " boxes"),
    textposition = "outside",
    textfont = list(size = 14, family = "Arial", color = "#1F2933")
  ) |>
  layout(
    xaxis = list(title = "", fixedrange = TRUE),
    yaxis = list(title = "Safety stock (boxes)", fixedrange = TRUE,
                 range = c(0, sb_dez_sf * 1.25)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 50, r = 20, t = 20, b = 40),
    height = 300, width = 420
  )
```

::: {.fragment .fade-up}
::: {.callout-warning appearance="simple"}
**56.0% less safety stock** through centralisation at Wiesbaden-Erbenheim — at an identical service level.
:::
:::

:::
:::

---

## Inventory pooling {#slide-pooling-zeitreihen}
::: slide-subtitle
Empirical illustration 
:::


```{r pooling-sim-data}
#| echo: false
#| warning: false
#| message: false

library(MASS)

set.seed(43)
n    <- 12
mu1  <- 120; mu2 <- 150   # Wiesbaden, Mainz
sig1 <- 25;  sig2 <- 30
rhos <- c(-0.8, 0.0, 0.8)

# Helper function: generate bivariate sample
gen_biv <- function(rho, n, mu1, mu2, sig1, sig2) {
  Sigma <- matrix(c(sig1^2, rho*sig1*sig2,
                    rho*sig1*sig2, sig2^2), 2, 2)
  raw   <- mvrnorm(n, mu = c(mu1, mu2), Sigma = Sigma)
  raw   <- round(raw)
  data.frame(t = 1:n, D1 = raw[,1], D2 = raw[,2])
}

# Helper function: scatter plot
make_scatter_fig <- function(df, rho_label, col_pt, height = 300, width= 300) {
  rho_emp <- round(cor(df$D1, df$D2), 2)
  fig <- plot_ly() 
  
  fig <- fig |> add_trace(
    data = df,
    x = ~D1, y = ~D2,
    type = "scatter", mode = "markers+text",
    text = ~paste0("t=", t),
    textposition = "top right",
    textfont = list(size = 10, color = "#6B7280"),
    marker = list(size = 12, color = col_pt, opacity = 0.85,
                  line = list(color = "white", width = 1.5)),
    hovertemplate = "t=%{text}<br>D₁=%{x}<br>D₂=%{y}<extra></extra>"
  )
  # Regression line
  fit   <- lm(D2 ~ D1, data = df)
  x_seq <- seq(min(df$D1)-5, max(df$D1)+5, length.out = 40)
  y_fit <- predict(fit, newdata = data.frame(D1 = x_seq))
  fig <- fig |>
    add_trace(x = x_seq, y = as.numeric(y_fit),
              type = "scatter", mode = "lines",
              name = "Regression line",
              showlegend = FALSE,
              line = list(color = "#6B7280", width = 1.5, dash = "dot"))
  fig |> layout(
    title  = list(
      # text = paste0("<b>ρ = ", rho_label,
      #               "</b>  (empirical: ", rho_emp, ")"),
      font = list(size = 13, color = "#1F2933", family = "Arial"),
      x = 0.04, xanchor = "left"),
    xaxis  = list(title = "D₁", fixedrange = TRUE,
                  range = c(min(df$D1)-10, max(df$D1)+15)),
    yaxis  = list(title = "D₂", fixedrange = TRUE,
                  range = c(min(df$D2)-10, max(df$D2)+15)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 45, r = 10, t = 40, b = 50),
    height = height, width = width 
  )
}

# Helper function: time series plot
make_ts_fig <- function(df, rho_label, col1 = "red", col2 = "blue", col0 ="green", height = 300, width = 300) {
  df$D0 <- df$D1 + df$D2
  fig <- plot_ly(df, x = ~t)
  fig <- fig |>
    add_trace(y = ~D1, type = "scatter", mode = "lines+markers",
              name = "D₁", line = list(color = col1, width = 2),
              marker = list(color = col1, size = 7)) |>
    add_trace(y = ~D2, type = "scatter", mode = "lines+markers",
              name = "D₂", line = list(color = col2, width = 2, dash = "dash"),
              marker = list(color = col2, size = 7, symbol = "square")) |>
    add_trace(y = ~D0, type = "scatter", mode = "lines+markers",
              name = "D₀ = D₁+D₂",
              line = list(color = col0, width = 2.5),
              marker = list(color = col0, size = 8, symbol = "diamond"))
  fig |> layout(
    # title  = list(text = paste0("<b>ρ = ", rho_label, "</b>"),
    #               font = list(size = 14, color = "#1F2933", family = "Arial"),
    #               x = 0.04, xanchor = "left"),
    xaxis  = list(title = "Period t", dtick = 2, fixedrange = TRUE),
    yaxis  = list(title = "Demand (boxes)", fixedrange = TRUE,
                  range = c(0, 420)),
    legend = list(orientation = "h", y = +0.95, font = list(size = 12)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 45, r = 10, t = 35, b = 65),
    height = height, width = width 
  )
}

sim_data <- lapply(rhos, gen_biv,
                   n = n, mu1 = mu1, mu2 = mu2, sig1 = sig1, sig2 = sig2)
names(sim_data) <- paste0("rho_", gsub("\\.", "", rhos))

# Actual correlations
rho_actual <- sapply(sim_data, function(df) round(cor(df$D1, df$D2), 2))
```


::: {.small-text}

:::: columns

::: {.column width="32%"}
**Negative correlation** $\rho = -0.8$ 
```{r pooling-zeitreihen-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

gen.wid <- 380
gen.hei <- 260

make_ts_fig(sim_data$`rho_-08`, "-0.8", height = gen.hei, width = gen.wid)

```

```{r pooling-scatter-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_scatter_fig(df = sim_data$`rho_-08`, "-0.8", cols$top, height = gen.hei, width = gen.wid)

```

:::

::: {.column width="32%"}
::: {.fragment .fade-up}
**No correlation** $\rho = 0$ 
```{r pooling-zeitreihen-plot2}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_ts_fig(sim_data$rho_0, "0", height = gen.hei, width = gen.wid)

```

```{r pooling-scatter-plot2}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_scatter_fig(df = sim_data$rho_0, "0", cols$top, height = gen.hei, width = gen.wid)

```
:::
:::

::: {.column width="32%"}
::: {.fragment .fade-up}
**Positive correlation** $\rho = +0.8$ 
```{r pooling-zeitreihen-plot3}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_ts_fig(sim_data$rho_08, "+0.8", height = gen.hei, width = gen.wid)

```

```{r pooling-scatter-plot3}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.8

make_scatter_fig(df = sim_data$rho_08, "+0.8", cols$top, height = gen.hei, width = gen.wid)

```
:::
:::
::::
:::

::: {.fragment .fade-up}
::: {.callout-note appearance="minimal" icon=false}
With **negative correlation**, the fluctuations in $D_0$ dampen each other — total demand is **more stable** than the individual demands. With **positive correlation**, they reinforce each other.
:::
:::

---

## Inventory pooling {#slide-pooling-sb-vergleich}
::: slide-subtitle
Safety stock as a function of the correlation $\rho$
:::

```{r pooling-sb-berechnung}
#| echo: false
#| warning: false
#| message: false

alpha_sl <- 0.95
z_alpha  <- qnorm(alpha_sl)

# Theoretical values for the Wiesbaden (sigma1=25) and Mainz (sigma2=30) stores
rho_seq  <- seq(-1, 1, by = 0.01)
sig1 <- 25; sig2 <- 30

sb_dez <- z_alpha * (sig1 + sig2)         # constant
sb_zen <- z_alpha * sqrt(sig1^2 + sig2^2 + 2 * rho_seq * sig1 * sig2)
einspar <- round((1 - sb_zen / sb_dez) * 100, 1)

df_sb <- data.frame(rho = rho_seq, sb_dez = sb_dez,
                    sb_zen = sb_zen, einspar = einspar)

# Marker points for rho = -0.8, 0, +0.8
rho_marks <- c(-0.8, 0.0, 0.8)
mark_df   <- data.frame(
  rho     = rho_marks,
  sb_zen  = z_alpha * sqrt(sig1^2 + sig2^2 + 2*rho_marks*sig1*sig2),
  einspar = round((1 - z_alpha * sqrt(sig1^2 + sig2^2 +
              2*rho_marks*sig1*sig2) / sb_dez) * 100, 1)
)
```

::: columns
::: {.column width="55%"}

```{r pooling-sb-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4.5

fig_sb <- plot_ly()

# Decentralised SS (reference line)
fig_sb <- fig_sb |>
  add_trace(x = df_sb$rho, y = df_sb$sb_dez,
            type = "scatter", mode = "lines",
            name = "sb<sub>dec</sub> (sum)",
            line = list(color = "#6B7280", width = 2, dash = "dash"))

# Centralised SS
fig_sb <- fig_sb |>
  add_trace(x = df_sb$rho, y = df_sb$sb_zen,
            type = "scatter", mode = "lines",
            name = "sb<sub>cen</sub>(ρ)",
            line = list(color = cols$top, width = 3))

# Savings area
fig_sb <- fig_sb |>
  add_trace(x = df_sb$rho, y = df_sb$sb_zen,
            type = "scatter", mode = "lines",
            fill = "tonexty", fillcolor = "rgba(45,158,142,0.15)",
            line = list(color = "transparent"),
            showlegend = FALSE, hoverinfo = "skip")

# Marker points
farben_marks <- c("#E76F51", cols$mid, "#E9C46A")
for (i in seq_along(rho_marks)) {
  fig_sb <- fig_sb |>
    add_trace(x = mark_df$rho[i], y = mark_df$sb_zen[i],
              type = "scatter", mode = "markers",
              name = paste0("ρ = ", rho_marks[i]),
              marker = list(size = 13, color = farben_marks[i],
                            line = list(color = "white", width = 2))) |>
    add_annotations(
      x = mark_df$rho[i],
      y = mark_df$sb_zen[i] - 3.5,
      xref = "x", yref = "y",
      text = paste0("<b>", round(mark_df$sb_zen[i], 1),
                    "<br>(−", mark_df$einspar[i], "%)</b>"),
      showarrow = FALSE,
      font = list(size = 12, color = farben_marks[i], family = "Arial"),
      xanchor = "center"
    )
}

fig_sb |> layout(
  xaxis  = list(title = "Correlation ρ", fixedrange = TRUE,
                tickvals = seq(-1, 1, 0.25)),
  yaxis  = list(title = "Safety stock sb",
                fixedrange = TRUE, range = c(0, sb_dez * 1.08)),
  legend = list(orientation = "h", y = -0.22, font = list(size = 12)),
  plot_bgcolor  = "transparent",
  paper_bgcolor = "transparent",
  margin = list(l = 50, r = 15, t = 15, b = 70),
  height = 400, width = 700
)
```

:::
::: {.column width="43%"}

::: {.fragment .fade-up}
::: {.callout-tip appearance="simple"}

## Three scenarios (Frankfurt $\sigma_1=25$, Offenbach $\sigma_2=30$, $\alpha = 95\%$)

| $\rho$ | $sb_{cen}$ | Savings |
|:---:|---:|---:|
| $-0.8$ | `r round(mark_df$sb_zen[1],1)` | `r mark_df$einspar[1]`% |
| $\;\;\;0$ | `r round(mark_df$sb_zen[2],1)` | `r mark_df$einspar[2]`% |
| $+0.8$ | `r round(mark_df$sb_zen[3],1)` | `r mark_df$einspar[3]`% |
| $(+1)$ | `r round(sb_dez,1)` | 0% |

$sb_{dec} = z_\alpha \cdot (\sigma_1 + \sigma_2) =$ `r round(sb_dez,1)`
:::
:::


:::
:::

::: {.small-text}
:::: columns
::: {.column width="49%"}

::: {.fragment .fade-up}
::::: {.callout-warning appearance="simple"}

## Centralisation gain if $\sigma_1 \neq \sigma_2$

$$1-\frac{sb_{cen}}{sb_{dec}} = 1-\frac{\sqrt{\sigma_1^2 + \sigma_2^2 + 2\rho\,\sigma_1\sigma_2}}{\sigma_1 + \sigma_2}$$
:::::
::: 
:::

::: {.column width="49%"}

::: {.fragment .fade-up}
::::: {.callout-warning appearance="simple"}

## Centralisation gain if $\sigma_1 = \sigma_2$

$$1-\frac{sb_{cen}}{sb_{dec}} = 1-\sqrt{\frac{1+\rho}{2}}$$
:::::
:::
:::
::::
:::

---

## Inventory pooling {#slide-pooling-realdata}
::: slide-subtitle
Simulated daily demand: Frankfurt & Offenbach
:::

```{r pooling-real-data-prep}
#| echo: false
#| warning: false
#| message: false

# Reuse simulated daily demand from the previous section
# (Frankfurt/Offenbach, set.seed(616), incl. shared holiday peaks)
dA_r <- dA
dB_r <- dB
n_r  <- min(length(dA_r), length(dB_r))
dA_r <- dA_r[seq_len(n_r)]
dB_r <- dB_r[seq_len(n_r)]
dPool_r <- dA_r + dB_r

muA_r  <- mean(dA_r);  sigA_r  <- sd(dA_r)
muB_r  <- mean(dB_r);  sigB_r  <- sd(dB_r)
muP_r  <- mean(dPool_r); sigP_r <- sd(dPool_r)
rho_r  <- cor(dA_r, dB_r)

alpha_r <- 0.95
z_r     <- qnorm(alpha_r)

# Empirical safety stocks
SB_A_emp  <- quantile(dA_r,    alpha_r) - muA_r
SB_B_emp  <- quantile(dB_r,    alpha_r) - muB_r
SB_dez_emp <- SB_A_emp + SB_B_emp
SB_zen_emp <- quantile(dPool_r, alpha_r) - muP_r

# Parametric safety stocks (normal distribution)
SB_A_par   <- z_r * sigA_r
SB_B_par   <- z_r * sigB_r
SB_dez_par <- SB_A_par + SB_B_par
SB_zen_par <- z_r * sqrt(sigA_r^2 + sigB_r^2 + 2 * rho_r * sigA_r * sigB_r)

einspar_emp <- round((1 - SB_zen_emp / SB_dez_emp) * 100, 1)
einspar_par <- round((1 - SB_zen_par / SB_dez_par) * 100, 1)

eA_r    <- make_ecdf_xy(dA_r)
eB_r    <- make_ecdf_xy(dB_r)
ePool_r <- make_ecdf_xy(dPool_r)
```

:::: columns
::: {.column width="50%"}

```{r pooling-real-ts}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.6

df_ts <- data.frame(
  t  = seq_len(n_r),
  D1 = dA_r,
  D2 = dB_r,
  D0 = dPool_r
)

fig_ts <- plot_ly(df_ts, x = ~t)

fig_ts <- fig_ts |>
  add_trace(y = ~D1, type = "scatter", mode = "lines",
            name = "Frankfurt",
            line   = list(color = cols$top, width = 1.6)) |>
  add_trace(y = ~D2, type = "scatter", mode = "lines",
            name = "Offenbach",
            line   = list(color = cols$mid, width = 1.6, dash = "dash")) |>
  add_trace(y = ~D0, type = "scatter", mode = "lines",
            name = "D₀ = Frankfurt + Offenbach",
            line   = list(color = cols$dash, width = 2)) |>
  layout(
    xaxis  = list(title = "Day", fixedrange = TRUE),
    yaxis  = list(title = "Sales (boxes)", fixedrange = TRUE),
    legend = list(orientation = "h", x= 0.1, y = 0.88, font = list(size = 14)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 50, r = 10, t = 10, b = 70),
    height = 240, width = 680
  )

fig_ts
```

::: {.fragment .fade-up}
```{r pooling-real-ecdf}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.6

xg_A    <- seq(min(dA_r)    - 200, max(dA_r)    + 200, length.out = 300)
xg_B    <- seq(min(dB_r)    - 200, max(dB_r)    + 200, length.out = 300)
xg_Pool <- seq(min(dPool_r) - 300, max(dPool_r) + 300, length.out = 300)

col_A_r    <- cols$top
col_B_r    <- cols$mid
col_Pool_r <- cols$dash

plot_ly() |>
  # ECDF Frankfurt
  add_trace(x = eA_r$x, y = eA_r$y, type = "scatter", mode = "lines+markers",
            name = "ECDF Frankfurt",
            line = list(color = col_A_r, width = 2.2),
            marker = list(color = col_A_r, size = 6,
                          line = list(color = "white", width = 1))) |>
  # Normal distribution Frankfurt
  add_trace(x = xg_A, y = pnorm(xg_A, muA_r, sigA_r),
            type = "scatter", mode = "lines",
            name = paste0("N(", round(muA_r), ", ", round(sigA_r), "\u00b2)"),
            line = list(color = col_A_r, width = 2, dash = "dot")) |>
  # ECDF Offenbach
  add_trace(x = eB_r$x, y = eB_r$y, type = "scatter", mode = "lines+markers",
            name = "ECDF Offenbach",
            line = list(color = col_B_r, width = 2.2, dash = "dash"),
            marker = list(color = col_B_r, size = 6, symbol = "square",
                          line = list(color = "white", width = 1))) |>
  # Normal distribution Offenbach
  add_trace(x = xg_B, y = pnorm(xg_B, muB_r, sigB_r),
            type = "scatter", mode = "lines",
            name = paste0("N(", round(muB_r), ", ", round(sigB_r), "\u00b2)"),
            line = list(color = col_B_r, width = 2, dash = "dot")) |>
  # ECDF total demand (pool)
  add_trace(x = ePool_r$x, y = ePool_r$y, type = "scatter", mode = "lines+markers",
            name = "ECDF D\u2080 (Pool)",
            line = list(color = col_Pool_r, width = 2.2),
            marker = list(color = col_Pool_r, size = 6, symbol = "diamond",
                          line = list(color = "white", width = 1))) |>
  # Normal distribution pool
  add_trace(x = xg_Pool, y = pnorm(xg_Pool, muP_r, sigP_r),
            type = "scatter", mode = "lines",
            name = paste0("N(", round(muP_r), ", ", round(sigP_r), "\u00b2)"),
            line = list(color = col_Pool_r, width = 2, dash = "dot")) |>
  
  layout(
    xaxis  = list(title = "Daily demand (boxes)", fixedrange = FALSE,
                  tickformat = ",.0f"),
    yaxis  = list(title = "F(d)", fixedrange = TRUE,
                  range = c(-0.05, 1.08), tickformat = ".0%"),
    legend = list(orientation = "h", x = .1, y = 0.38, font = list(size = 11)),
    plot_bgcolor  = "transparent", paper_bgcolor = "transparent",
    margin = list(l = 50, r = 10, t = 5, b = 75), height = 240, width = 660
  )
```
:::

::: {.fragment .fade-up}
```{r pooling-real-scatter}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 2.5

fit_r   <- lm(dB_r ~ dA_r)
x_seq_r <- seq(min(dA_r) - 5, max(dA_r) + 5, length.out = 80)
y_fit_r <- predict(fit_r, newdata = data.frame(dA_r = x_seq_r))

fig_sc <- plot_ly() |>
  add_trace(
    x = dA_r, y = dB_r,
    type = "scatter", mode = "markers",
    name = "Observations",
    marker = list(size = 9, color = cols$top, opacity = 0.8,
                  line = list(color = "white", width = 1.2)),
    hovertemplate = "Frankfurt: %{x}<br>Offenbach: %{y}<extra></extra>"
  ) |>
  add_trace(
    x = x_seq_r, y = as.numeric(y_fit_r),
    type = "scatter", mode = "lines",
    name = "Regression line",
    line = list(color = "#6B7280", width = 1.6, dash = "dot"),
    hoverinfo = "skip"
  ) |>
  add_annotations(
    x = 0.97, y = 0.95, xref = "paper", yref = "paper",
    text = paste0("<b>ρ̂ = ", round(rho_r, 2), "</b>"),
    showarrow = FALSE,
    font = list(size = 16, color = "#1F2933", family = "Arial"),
    xanchor = "right"
  ) |>
  layout(
    xaxis  = list(title = "Daily sales Frankfurt", fixedrange = TRUE),
    yaxis  = list(title = "Sales Offenbach", fixedrange = TRUE),
    legend = list(orientation = "h", x = .1, y = 0.88, font = list(size = 14)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 50, r = 10, t = 10, b = 70),
    height = 240, width = 680
  )

fig_sc
```
:::

:::

::: {.column width="50%"}

::: {.fragment .fade-up}

```{r pooling-real-sb-table}
#| echo: false
#| warning: false
#| message: false

sb_df <- data.frame(
  Kennzahl = c(
    "μ̂", "σ̂",
    "SS decentralised (emp.)", "SS centralised (emp.)", "Savings (emp.)",
    "SS decentralised (norm.)", "SS centralised (norm.)", "Savings (norm.)"
  ),
  Frankfurt = c(round(muA_r, 1), round(sigA_r, 1),
            round(SB_A_emp, 1), "—", "—",
            round(SB_A_par, 1), "—", "—"),
  Offenbach = c(round(muB_r, 1), round(sigB_r, 1),
            round(SB_B_emp, 1), "—", "—",
            round(SB_B_par, 1), "—", "—"),
  Pool  = c(round(muP_r, 1), round(sigP_r, 1),
            round(SB_dez_emp, 1), round(SB_zen_emp, 1),
            paste0(einspar_emp, "%"),
            round(SB_dez_par, 1), round(SB_zen_par, 1),
            paste0(einspar_par, "%"))
)

kbl(sb_df,
    col.names = c("KPI", "Frankfurt", "Offenbach", "Total / pool"),
    align = c("l", "r", "r", "r"),
    escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed", "striped"),
                full_width = TRUE, font_size = 18) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(1:2, background = "#F8F9FA") |>
  row_spec(3:5, background = "#E8F4F1") |>
  row_spec(6:8, background = "#FFF3CD") |>
  row_spec(c(5, 8), bold = TRUE)
```

:::

::: {.fragment .fade-up}

```{r pooling-real-sb-bar}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 3.0

kategorien <- c("Empirical", "Normal distribution")
sb_dez_vals <- c(round(SB_dez_emp, 1), round(SB_dez_par, 1))
sb_zen_vals <- c(round(SB_zen_emp, 1), round(SB_zen_par, 1))

fig_bar <- plot_ly() |>
  add_trace(
    x = kategorien, y = sb_dez_vals,
    type = "bar", name = "Decentralised (SS₁ + SS₂)",
    marker = list(color = "#E9C46A",
                  line = list(color = "#B8951A", width = 1)),
    text  = round(sb_dez_vals, 1), textposition = "outside",
    textfont = list(size = 13, family = "Arial", color = "#1F2933")
  ) |>
  add_trace(
    x = kategorien, y = sb_zen_vals,
    type = "bar", name = "Centralised (pooling)",
    marker = list(color = cols$top,
                  line = list(color = "#01696f", width = 1)),
    text  = round(sb_zen_vals, 1), textposition = "outside",
    textfont = list(size = 13, family = "Arial", color = "#1F2933")
  ) |>
  layout(
    barmode = "group",
    xaxis   = list(title = "", fixedrange = TRUE),
    yaxis   = list(title = "Safety stock (units)",
                   fixedrange = TRUE,
                   range = c(0, max(sb_dez_vals) * 1.25)),
    legend  = list(orientation = "h", y = -0.28, font = list(size = 11)),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent",
    margin = list(l = 50, r = 20, t = 20, b = 70),
    height = 270, width = 680
  )

fig_bar
```

:::

::: {.fragment .fade-up}
::: {.callout-note appearance="minimal" icon=false}
Correlation $\hat{\rho} =$ `r round(rho_r, 2)` $\rightarrow$ pooling savings, empirical: **`r einspar_emp`%**, parametric: **`r einspar_par`%**
:::
:::

:::
::::


## Aside: normal distribution {#slide-exkurs-normal-intro}
::: slide-subtitle
Probability density function & cumulative distribution function
:::

::: {.callout-note appearance="minimal" icon=false}

## Formal definition

The **density function** $f(x)$ gives the relative frequency with which a value $x$ occurs, and the **cumulative distribution function** $F(x)$ is the cumulative probability:

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}}\,\exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right) \qquad F(x) = P(X \leq x) = \int_{-\infty}^{x} f(t)\,dt$$

:::


:::: columns
::: {.column width="50%"}

```{r exkurs-normal-pdf-cdf}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4.8

mu_ex    <- 0
sig_ex   <- 1
x_ex     <- seq(-4, 4, length.out = 400)
y_pdf_ex <- dnorm(x_ex, mu_ex, sig_ex)
y_cdf_ex <- pnorm(x_ex, mu_ex, sig_ex)

# Shaded area: μ ± σ
x_shade <- seq(-1, 1, length.out = 200)
y_shade <- dnorm(x_shade, mu_ex, sig_ex)

plot_ly() |>
  # ── PDF ──────────────────────────────────────────────────────────
  add_trace(x = x_ex, y = y_pdf_ex, type = "scatter", mode = "lines",
            name = "f(x) — density",
            line = list(color = cols$top, width = 2.8),
            yaxis = "y") |>
  # Shading μ ± σ (approx. 68%)
  add_trace(
    x = c(x_shade, rev(x_shade)),
    y = c(rep(0, 200), rev(y_shade)),
    type = "scatter", mode = "none",
    fill = "toself",
    fillcolor = "rgba(1,105,111,0.18)",
    name = "\u00b11\u03c3 \u2248 68%",
    yaxis = "y", showlegend = TRUE) |>
  # CDF on second y-axis
  add_trace(x = x_ex, y = y_cdf_ex, type = "scatter", mode = "lines",
            name = "F(x) — CDF",
            line = list(color = cols$dash, width = 2.4, dash = "dash"),
            yaxis = "y2") |>
  # Quantile marker: F(1) ≈ 0.841
  add_segments(x = 1, xend = 1, y = 0, yend = 1,
               line = list(color = cols$mid, width = 1.4, dash = "dot"),
               yaxis = "y2", showlegend = FALSE) |>
  add_annotations(x = 1.08, y = 0.84, yref = "y2",
    text = "<b>F(1) ≈ 0.841</b>",
    showarrow = FALSE, xanchor = "left",
    font = list(size = 12, color = cols$mid, family = "Arial")) |>
  layout(
    xaxis  = list(title = "x", fixedrange = TRUE,
                  tickvals = list(-3,-2,-1,0,1,2,3),
                  ticktext = list("μ−3σ","μ−2σ","μ−σ","μ","μ+σ","μ+2σ","μ+3σ")),
    yaxis  = list(title = "f(x) — density function", fixedrange = TRUE,
                  range = c(0, 0.46), side = "left"),
    yaxis2 = list(title = "F(x) — distribution function", overlaying = "y",
                  side = "right", range = c(0, 1.05),
                  tickformat = ".0%", fixedrange = TRUE),
    legend = list(orientation = "h", y = -0.28, font = list(size = 11)),
    plot_bgcolor  = "transparent", paper_bgcolor = "transparent",
    margin = list(l = 55, r = 65, t = 5, b = 70), height = 420, width = 680
  )
```

:::

::: {.column width="50%"}


::: {.callout-tip appearance="minimal" icon=false}
**Symmetry property and 68–95–99.7 rule:**

$$F(-z) = 1 - F(z)$$

| Range of values | Probability |
|:--------|:--------:|
| $\mu \pm 1\sigma$ | ≈ 68.3% |
| $\mu \pm 2\sigma$ | ≈ 95.4% |
| $\mu \pm 3\sigma$ | ≈ 99.7% |

:::

:::
::::

---

## Aside: standard normal distribution {#slide-exkurs-stdnorm}
::: slide-subtitle
z-transformation & standardisation
:::

::: {.callout-note appearance="minimal" icon=false}
The **z-transformation** converts any normal distribution $X \sim \mathcal{N}(\mu, \sigma^2)$ into the **standard normal distribution** $Z \sim \mathcal{N}(0, 1)$:

$$Z = \frac{X - \mu}{\sigma} \quad \Rightarrow P(X \leq x) = P\!\left(Z \leq \frac{x-\mu}{\sigma}\right) = \Phi(z)$$
$\rightarrow$ all $\mathcal{N}(\mu, \sigma^2)$ can be reduced to *one* table — the **standard normal distribution table** $\Phi(z)$
:::

:::: columns
::: {.column width="52%"}

```{r exkurs-stdnorm-plot}
#| echo: false
#| warning: false
#| message: false
#| out-width: "99%"
#| fig-height: 4.8

z_seq    <- seq(-4, 4, length.out = 400)
y_z_pdf  <- dnorm(z_seq)
y_z_cdf  <- pnorm(z_seq)

# Shadings
make_shade <- function(a, b) {
  xs <- seq(a, b, length.out = 150)
  list(x = c(xs, rev(xs)), y = c(rep(0, 150), rev(dnorm(xs))))
}
sh1 <- make_shade(-1.96, 1.96)   # 95%
sh2 <- make_shade(min(z_seq), -1.96)
sh3 <- make_shade( 1.96,  max(z_seq))

plot_ly() |>
  add_trace(x = sh1$x, y = sh1$y, type = "scatter", mode = "none",
            fill = "toself", fillcolor = "rgba(1,105,111,0.20)",
            name = "95%  (±1.96)", showlegend = TRUE) |>
  add_trace(x = sh2$x, y = sh2$y, type = "scatter", mode = "none",
            fill = "toself", fillcolor = "rgba(193,58,46,0.20)",
            name = "Rejection region", showlegend = TRUE) |>
  add_trace(x = sh3$x, y = sh3$y, type = "scatter", mode = "none",
            fill = "toself", fillcolor = "rgba(193,58,46,0.20)",
            showlegend = FALSE) |>
  add_trace(x = z_seq, y = y_z_pdf, type = "scatter", mode = "lines",
            name = "φ(z) — standard density",
            line = list(color = cols$top, width = 2.8)) |>
  # Vertical markers ± 1.96
  add_segments(x = 1.96,  xend = 1.96,  y = 0, yend = dnorm(1.96),
               line = list(color = cols$mid, dash = "dot", width = 1.4),
               showlegend = FALSE) |>
  add_segments(x = -1.96, xend = -1.96, y = 0, yend = dnorm(1.96),
               line = list(color = cols$mid, dash = "dot", width = 1.4),
               showlegend = FALSE) |>
  add_annotations(x = 2.05,  y = 0.08, text = "<b>z = 1.96</b>",
    showarrow = FALSE, xanchor = "left",
    font = list(size = 11, color = cols$mid, family = "Arial")) |>
  add_annotations(x = -2.05, y = 0.08, text = "<b>z = −1.96</b>",
    showarrow = FALSE, xanchor = "right",
    font = list(size = 11, color = cols$mid, family = "Arial")) |>
  layout(
    xaxis  = list(title = "z = (x − μ) / σ", fixedrange = TRUE,
                  tickvals = list(-3,-2,-1,0,1,2,3)),
    yaxis  = list(title = "φ(z)", fixedrange = TRUE, range = c(0, 0.44)),
    legend = list(orientation = "h", y = -0.28, font = list(size = 11)),
    plot_bgcolor  = "transparent", paper_bgcolor = "transparent",
    margin = list(l = 50, r = 15, t = 5, b = 70), height = 420, width = 680
  )
```

:::

::: {.column width="48%"}

::: {.callout-note appearance="minimal" icon=false}
**Important quantiles of $\mathcal{N}(0,1)$:**

| Service level $\alpha$ | $z_\alpha = \Phi^{-1}(\alpha)$ |
|:--------------------:|:------------------------------:|
| 90.0% | 1.282 |
| 95.0% | 1.645 |
| 97.5% | 1.960 |
| 99.0% | 2.326 |
| 99.9% | 3.090 |

:::

:::
::::

---

## Aside: quantile/distribution table {#slide-exkurs-ztable}
::: slide-subtitle
Reading $\Phi(z)$ and its inverse $z_\alpha = \Phi^{-1}(\alpha)$
:::

:::: columns
::: {.column width="58%"}

```{r exkurs-ztable}
#| echo: false
#| warning: false
#| message: false

# Interactive z table (rows: z integer + one decimal place,
#                      columns: second decimal place 0.00 … 0.09)
z_rows <- seq(0.0, 3.0, by = 0.1)
z_cols <- seq(0.00, 0.09, by = 0.01)

z_mat <- outer(z_rows, z_cols, function(a, b) {
  sprintf("%.4f", pnorm(a + b))
})

df_ztab <- as.data.frame(z_mat)
colnames(df_ztab) <- sprintf("+%.2f", z_cols)
rownames(df_ztab) <- sprintf("%.1f", z_rows)

# Highlight rows: z = 1.6 (≈ 95%) and z = 2.3 (≈ 99%)
highlight_rows <- which(rownames(df_ztab) %in% c("1.6", "2.3"))

kbl(df_ztab,
    caption = "Φ(z) — standard normal distribution (z ≥ 0)",
    align = "r", escape = FALSE) |>
  kable_styling(bootstrap_options = c("condensed", "striped", "hover"),
                full_width = TRUE, font_size = 12) |>
  row_spec(0, background = "#2F3E46", color = "white", bold = TRUE) |>
  row_spec(highlight_rows,
           background = "#E8F4F1", bold = TRUE) |>
  column_spec(6,  bold = TRUE, color = "#01696f") |>   # column +0.04 (→ 1.645)
  column_spec(4,  bold = TRUE, color = "#8B6914")       # column +0.02 (→ 2.326)
```

:::

::: {.column width="42%"}

::: {.callout-note appearance="minimal" icon=false}
**Reading the table:**  
Wanted: $P(X \leq x)$ for $X \sim \mathcal{N}(\mu, \sigma^2)$

1. Compute $z = (x - \mu)/\sigma$  
2. Row = first decimal place of $z$  
3. Column = second decimal place of $z$  
4. Table value = $\Phi(z)$
:::

::: {.callout-tip appearance="minimal" icon=false}
**Inventory example:**  
$D \sim \mathcal{N}(500, 80^2)$, service level $\alpha = 95\%$

$$z_{0.95} = \Phi^{-1}(0.95) = 1.645$$

$$\Rightarrow\; \text{SS} = z_\alpha \cdot \sigma = 1.645 \cdot 80 \approx \mathbf{132\;\text{units}}$$
:::


:::
::::
