## Uncertainty in Supply Chains

- Risk pooling
- **Newsvendor model**
- Bullwhip effect

```{r}
#| include: false

source("../R/scm_functions.R")

library(ggplot2)
library(dplyr)
library(scales)
library(plotly)
library(tidyverse)

# Example parameters ------------------------------------------------------

mu <- 1000
sigma <- 200
p <- 20
c <- 8
v <- 2
g <- 0

underage_cost <- p - c + g
overage_cost <- c - v
critical_ratio <- underage_cost / (underage_cost + overage_cost)
q_star <- qnorm(critical_ratio, mean = mu, sd = sigma)

# Local helper function for the visualisation ----------------------------

plot_newsvendor_distribution <- function(
    mean_demand,
    sd_demand,
    order_quantity,
    critical_ratio = NULL,
    title = "Newsvendor model: demand and order quantity"
) {
  stopifnot(
    is.finite(mean_demand),
    is.finite(sd_demand),
    sd_demand > 0,
    is.finite(order_quantity)
  )

  x_min <- min(0, mean_demand - 4 * sd_demand)
  x_max <- max(order_quantity, mean_demand + 4 * sd_demand)

  demand <- tibble(
    demand = seq(x_min, x_max, length.out = 1000),
    density = dnorm(demand, mean = mean_demand, sd = sd_demand)
  )

  left_area <- demand |> filter(demand <= order_quantity)
  right_area <- demand |> filter(demand >= order_quantity)

  ggplot(demand, aes(x = demand, y = density)) +
    geom_area(
      data = left_area,
      fill = "#00843D",
      alpha = 0.25
    ) +
    geom_area(
      data = right_area,
      fill = "#E85D04",
      alpha = 0.20
    ) +
    geom_line(
      linewidth = 1.1,
      colour = "#1F2937"
    ) +
    geom_vline(
      xintercept = mean_demand,
      linetype = "dashed",
      colour = "#6B7280"
    ) +
    geom_vline(
      xintercept = order_quantity,
      linewidth = 1.2,
      colour = "#00843D"
    ) +
    annotate(
      "text",
      x = order_quantity,
      y = max(demand$density) * 0.95,
      label = paste0(
        "Q* = ",
        format(
          round(order_quantity, 0),
          big.mark = ",",
          scientific = FALSE
        )
      ),
      hjust = -0.05,
      colour = "#00843D",
      fontface = "bold"
    ) +
    labs(
      title = title,
      subtitle = if (is.null(critical_ratio)) {
        NULL
      } else {
        paste0(
          "Critical fractile: ",
          number(
            critical_ratio,
            accuracy = 0.001,
            decimal.mark = "."
          )
        )
      },
      x = "Demand",
      y = "Density"
    ) +
    theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```

------------------------------------------------------------------------------

## Learning objectives {#newsvendor-learning-objectives}

After this unit you will be able to …

- classify the **newsvendor problem** as a one-off ordering decision under
  uncertain demand,
- determine overage and underage costs,
- derive and interpret the critical fractile rule,
- calculate an optimal order quantity for a given demand
  distribution,
- explain the influence of key model parameters on the ordering
  decision.


**Further reading**

- **Introduction to the newsvendor model:** [@thonemann2023operations, Chapter 5].
- **Single-period inventory models and critical fractile logic:** [@cachon2024matching, Chapter 13].
- **Newsvendor models in the context of inventory management:** [@silver2017inventory, Chapter 9].


------------------------------------------------------------------------------

## Case example {#newsvendor-fallbeispiel-intro}
::: slide-subtitle
Sushi boxes in the supermarket
:::

Every day, a supermarket sources freshly produced sushi boxes from
**SushiFresh GmbH**, a regional sushi producer in the Rhine-Main region.

```{r}
#| fig-height: 2
#| fig-width: 15
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Interactive timeline of the daily sushi box decision"

library(plotly)
library(tibble)

timeline <- tibble(
  zeitpunkt = c(0, 6, 24, 30),
  label = c(
    "Morning, 7 am<br>Fill sushi boxes",
    "8 am<br>Supermarket sales begin",
    "7 pm<br>Collection of leftovers",
    "8 pm<br>Dispose of leftovers"
  ),
  tooltip = c(
    "<b>7 am</b><br>Sushi boxes are freshly prepared.",
    "<b>8 am</b><br>Sales in the supermarket begin.",
    "<b>7 pm</b><br>Unsold boxes are collected.",
    "<b>8 pm</b><br>Remaining stock is discounted, donated or disposed of."
  )
)

plot_ly(
  data = timeline,
  x = ~zeitpunkt,
  y = ~rep(0, nrow(timeline)),
  type = "scatter",
  mode = "markers",
  marker = list(
    size = 18,
    color = "#00843D",
    line = list(
      color = "white",
      width = 2
    )
  ),
  text = ~tooltip,
  hoverinfo = "text",
  showlegend = FALSE
) |>
  add_segments(
    x = 0,
    xend = 30.5,
    y = 0,
    yend = 0,
    line = list(
      color = "#6B7280",
      width = 2
    ),
    inherit = FALSE,
    hoverinfo = "none",
    showlegend = FALSE
  ) |>
  add_segments(
    x = 6,
    xend = 24,
    y = 0,
    yend = 0,
    line = list(
      color = "#00843D",
      width = 5
    ),
    inherit = FALSE,
    hoverinfo = "none",
    showlegend = FALSE
  ) |>
  add_annotations(
    x = timeline$zeitpunkt,
    y = 0,
    text = timeline$label,
    showarrow = FALSE,
    yshift = -65,
    font = list(
      size = 18,
      color = "#1F2937"
    ),
    align = "center"
  ) |>
  add_annotations(
    x = 15,
    y = 0,
    text = "Selling period",
    showarrow = FALSE,
    yshift = 35,
    font = list(
      size = 18,
      color = "#00843D"
    )
  ) |>
  layout(
    margin = list(
      l = 10,
      r = 10,
      t = 20,
      b = 80
    ),
    xaxis = list(
      range = c(-5, 35),
      visible = FALSE,
      fixedrange = TRUE
    ),
    yaxis = list(
      range = c(-.5, .75),
      visible = FALSE,
      fixedrange = TRUE
    ),
    hoverlabel = list(
      bgcolor = "white",
      bordercolor = "#00843D",
      font = list(
        size = 16,
        color = "#1F2937"
      )
    ),
    width = 1500,
    height = 150
  ) |>
  config(
    displayModeBar = FALSE,
    responsive = TRUE
  )
```

- The boxes are prepared in the morning and delivered before the store opens.
- Demand is uncertain until the evening. Replenishment during the day
  is not possible.
- Unsold sushi boxes may not be sold at regular price on the
  following day.

::: {.fragment .fade-up}
> **Decision question:** How many sushi boxes should the supplier
> prepare?
:::


------------------------------------------------------------------------------

## Case example {#newsvendor-fallbeispiel-parameters}
::: slide-subtitle
Cost and revenue structure
:::

The following simplified values apply to a sushi box:

| Quantity | Meaning | Value |
|-------------------|----------------------|----:|
| Selling price $p$ | Revenue per sushi box sold | €10 |
| Production and delivery costs $c$ | Cost per sushi box prepared | €5 |
| Salvage value $v$ | Revenue from evening discount or disposal | €2 |

**In general, typically: $p > c > v$**

::: {.vspace-2}
:::

::: {.fragment .fade-up}
$\rightarrow$ **Contribution margin per sushi box sold**: $c^u=p-c = 10-5=5$ €
:::

::: {.vspace-2}
:::

::: {.fragment .fade-up}
$\rightarrow$ **Salvage loss per unsold sushi box**: $c^o=c-v=5-2=3$ €
:::

------------------------------------------------------------------------------

## Case example {#newsvendor-fallbeispiel-perspectives}
::: slide-subtitle
Two possible situations from the supplier's perspective
:::

::::: columns
::: {.column width="50%"}
**Too many boxes:**

::: incremental
- less demand $D$ than sushi boxes produced $q$ $\rightarrow$ excess stock $q-D$
- quantity sold: $\min(D,q) = D$
- contribution margin from sales: $D \cdot c^u$
- additional salvage costs: $(q-D) \cdot c^o$
:::

::: {.vspace-2}
:::

::: {.fragment .fade-up}
**Total contribution margin:** $P(D,q)=  D \cdot c^u - (q - D) \cdot c^o$
:::

:::

::: {.column width="50%"}
::: {.fragment .fade-up}
**Too few boxes:**

::: incremental
- more demand $D$ than sushi boxes produced $q$ $\rightarrow$ shortage $D-q$
- quantity sold: $\min(D,q) = q$
- total contribution margin: $q \cdot c^u$
- opportunity costs from lost contribution margin: $(D-q) \cdot c^u$
:::


::: {.vspace-2}
:::

::: {.fragment .fade-up}
**Total contribution margin:** $P(D,q) = q \cdot c^u$
:::
:::
:::
:::::

------------------------------------------------------------------------------

## Case example {#newsvendor-fallbesipiel-data}
::: slide-subtitle
Possible demand & production quantity
:::

As a simplified miniature example (e.g. for a small store or
a test sale), SushiFresh first estimates four possible demand
scenarios from sales history, day of the week, weather forecast
and current promotions:

| Demand $D$ | Probability $P(D=d)$ | Cumulative probability $F(d)$ |
|----------:|-------------------:|-----------------------:|
| 30 sushi boxes | 0.20 | 0.20 |
| 40 sushi boxes | 0.35 | 0.55 |
| 50 sushi boxes | 0.30 | 0.85 |
| 60 sushi boxes | 0.15 | 1.00 |

> Which quantity should the supplier prepare in the morning: 30, 40, 50 or
> 60 sushi boxes?

------------------------------------------------------------------------------

## Case example {#newsvendor-fallbesipiel-lsg}
::: slide-subtitle
Expected result
:::


::::: columns
::: {.column width="55%"}

**Set $q=40$ sushi boxes**

::: {.smaller-text}
| Demand $D$ | Prob. $P(D=d)$ | Excess stock $q-D$ |  Shortage $q-D$ |  Contri-\n bution $P$ |
|------:|------:|------:|-----:|------:|
| 30 sushi boxes | 0.20 | 10 |  0 | €120
| 40 sushi boxes | 0.35 | 0  |  0 | €200
| 50 sushi boxes | 0.30 | 0  | 10 | €200
| 60 sushi boxes | 0.15 | 0  | 20 | €200
:::

::: {.vspace-2}
:::

::: {.fragment .fade-up}
**Expected total contribution margin:** $\begin{align} E(P|q=40)= & 0.2 \cdot 120 + 0.35 \cdot 200 + \\
& 0.30 \cdot 200 + 0.15 \cdot 200= 184 \end{align}$
:::

:::


::: {.column width="45%"}

::: {.fragment .fade-up}
**$E(P|q)$ for all $q$**

```{r}
#| fig-height: 4.75
#| fig-width: 5
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Expected contribution margin for different order quantities of sushi boxes"

# Demand distribution from the sushi case example
demand <- c(30, 40, 50, 60)
probability <- c(0.20, 0.35, 0.30, 0.15)

# Parameters of the case example
p <- 10  # Selling price per sushi box
c <- 5  # Production and delivery costs per sushi box
v <- 2  # Salvage value per unsold sushi box

# Expected contribution margin, i.e. expected profit per period
expected_profit <- function(order_quantity) {
  realised_profit <-
    p * pmin(order_quantity, demand) +
    v * pmax(order_quantity - demand, 0) -
    c * order_quantity
    #(p-c) * pmax(demand - order_quantity, 0)

  sum(probability * realised_profit)
}

profit_table <- tibble(
  bestellmenge = demand,
  erwarteter_deckungsbeitrag = sapply(demand, expected_profit)
) |>
  mutate(
    optimal = erwarteter_deckungsbeitrag == max(erwarteter_deckungsbeitrag),
    hover_text = paste0(
      "<b>Order quantity: ", bestellmenge, " sushi boxes</b>",
      "<br>Expected contribution margin: ",
      number(
        erwarteter_deckungsbeitrag,
        accuracy = 0.01,
        decimal.mark = ".",
        big.mark = ","
      ),
      " EUR",
      ifelse(
        optimal,
        "<br><span style='color:#E85D04'><b>Optimal order quantity</b></span>",
        ""
      )
    )
  )

optimal_row <- profit_table |>
  filter(optimal)

plot_ly(
  data = profit_table,
  x = ~bestellmenge,
  y = ~erwarteter_deckungsbeitrag,
  type = "scatter",
  mode = "lines+markers",
  text = ~hover_text,
  hoverinfo = "text",
  line = list(
    color = "#00843D",
    width = 4
  ),
  marker = list(
    color = "#00843D",
    size = 13,
    line = list(
      color = "white",
      width = 2
    )
  ),
  showlegend = FALSE
) |>
  add_markers(
    data = optimal_row,
    x = ~bestellmenge,
    y = ~erwarteter_deckungsbeitrag,
    text = ~hover_text,
    hoverinfo = "text",
    marker = list(
      color = "#E85D04",
      size = 19,
      line = list(
        color = "white",
        width = 3
      )
    ),
    inherit = FALSE,
    showlegend = FALSE
  ) |>
  add_annotations(
    x = optimal_row$bestellmenge,
    y = optimal_row$erwarteter_deckungsbeitrag,
    text = paste0(
      "<b>Optimum:</b> ",
      optimal_row$bestellmenge,
      " boxes"
    ),
    showarrow = TRUE,
    arrowhead = 2,
    ax = 55,
    ay = -45,
    font = list(
      size = 17,
      color = "#E85D04"
    ),
    arrowcolor = "#E85D04",
    bgcolor = "rgba(255,255,255,0.9)",
    bordercolor = "#E85D04",
    borderwidth = 1
  ) |>
  layout(
    margin = list(
      l = 80,
      r = 25,
      t = 25,
      b = 70
    ),
    xaxis = list(
      title = list(
        text = "Order quantity q"
      ),
      tickmode = "array",
      tickvals = profit_table$bestellmenge,
      ticktext = profit_table$bestellmenge,
      fixedrange = TRUE,
      zeroline = FALSE
    ),
    yaxis = list(
      title = list(
        text = "Expected contribution margin E(P(Q)) in €"
      ),
      tickprefix = "",
      ticksuffix = " EUR",
      separatethousands = TRUE,
      fixedrange = TRUE,
      zeroline = FALSE
    ),
    hoverlabel = list(
      bgcolor = "white",
      bordercolor = "#00843D",
      font = list(
        size = 16,
        color = "#1F2937"
      )
    ),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor = "rgba(0,0,0,0)",
    width = 600,
    height = 450
  ) |>
  config(
    displayModeBar = FALSE,
    responsive = TRUE
  )
```
:::

::: {.fragment .fade-up}
$\rightarrow$ maximum expected contribution margin at $q=50$ with € `r round(expected_profit(50),2)` per day
:::

:::

:::::

------------------------------------------------------------------------------

## Newsvendor model {#newsvendor-objective}
::: slide-subtitle
Derivation of the objective function
:::

::::: columns
::: {.column width="50%"}
**Overage** $q > D$

$$P(D,q) = D \cdot c^u - (q-D)^+\cdot c^o$$

$$\Downarrow$$
:::

::: {.column width="50%"}
**Underage** $q < D$

$$P(D,q) = q \cdot c^u$$

$$\Downarrow$$
:::
:::::

::: {.small-text}

::: {.fragment .fade-up}
$$\begin{align} P(D,q) =& \min(D,q) \cdot c^u - (q-D)^+ \cdot c^o \\
= & (D-(D-q)^+) \cdot c^u - (q-D)^+ \cdot c^o \\
= & \underset{\text{contribution margin}}{D \cdot c^u} - \underset{\text{ shortage costs}}{(D-q)^+ \cdot c^u} - \underset{\text{excess stock costs}}{(q-D)^+ \cdot c^o}
\end{align}
$$
where $(q-D)^+ = \max(0,q-D)$, $(D-q)^+ = \max(0,D-q)$ and $\min(D,q) = D - (D-q)^+$
:::
:::

------------------------------------------------------------------------------

## Newsvendor model {#newsvendor-formal-model}
::: slide-subtitle
Formal model
:::

::: {.callout-note }

## Problem description

Find the order quantity $q$ that maximises expected profit or minimises expected costs:
$$ \underset{q}{\max} E(P(q)) = \mu \cdot c^u - E((D-q)^+) \cdot c^u - E((q-D)^+) \cdot c^o \; \Leftrightarrow \; \underset{q}{\min} E(C(q)) =  E((D-q)^+) \cdot c^u + E((q-D)^+) \cdot c^o $$
$\rightarrow$ minimise the weighted average of expected shortage $E((D-q)^+)$ and expected excess stock $E((q-D)^+)$
:::


::: {.columns}
::: {.column width="50%"}

::: {.fragment .fade-up}

**Discrete demand** [(cf. previous slide)](#newsvendor-fallbesipiel-lsg)

```{r newsvendor-discrete-plot}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3.5

# Values from the slide “Case example: possible demand & production quantity”
d_disc <- c(30, 40, 50, 60)
p_disc <- c(0.20, 0.35, 0.30, 0.15)

stopifnot(length(d_disc) == length(p_disc), abs(sum(p_disc) - 1) < 1e-10)

loss_disc <- function(q) {
  sum(p_disc * pmax(d_disc - q, 0))
}

sales_disc <- function(q) {
  sum(p_disc * pmax(q - d_disc,0))
}

df_exp_disc <- data.frame(
  q = d_disc,
  expected_loss = vapply(d_disc, loss_disc, numeric(1)),
  expected_sales = vapply(d_disc, sales_disc, numeric(1))
)

plot_ly(
  df_exp_disc,
  x = ~q,
  y = ~expected_loss,
  type = "scatter",
  mode = "lines+markers",
  name = "E[(D - q)<sup>+</sup>] - exp. shortage",
  line = list(color = "#D55E00", width = 3),
  marker = list(color = "#D55E00", size = 10),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage = %{y:.2f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales,
    name = "E[(q-D)<sup>+</sup>] - exp. excess stock",
    line = list(color = "#0072B2", width = 3),
    marker = list(color = "#0072B2", size = 10),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock = %{y:.2f}<extra></extra>"
  ) |>
  add_lines(
    y = p_disc,
    x = d_disc,
    name = "Density f(D)",
    mode = "lines+markers",
    yaxis = "y2",
    line = list(color = "#15a202", width = 1, dash = "dash"),
    marker = list(color = "#15a202", size = 5),
    hovertemplate = paste0(
      "Demand = %{x:.0f}",
      "<br>Density f(D) = %{y:.2f}",
      "<extra></extra>")
  ) |>
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "Expected quantity", rangemode = "tozero"),
    yaxis2 = list(
      title = "Density f(D)",
      overlaying = "y",
      side = "right",
      rangemode = "tozero",
      showgrid = FALSE,
      zeroline = FALSE,
      tickfont = list(color = "#15A202"),
      titlefont = list(color = "#15A202")
    ),
    margin = list(l = 70, r = 70, t = 45, b = 45),
    legend = list(orientation = "h", x = 0.5, y = 0.85),
    width = 650,
  height = 320
  )


```
:::
:::

::: {.column width="50%"}
::: {.fragment .fade-up}
**Continuous demand (Frankfurt store)** $D \sim N(280,55^2)$

```{r newsvendor-continuous-plot}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3.5

mu <- 280
sigma <- 55
q_show_cont <- 280

q_grid_cont <- seq(mu - 4 * sigma, mu + 4 * sigma, length.out = 500)

# For D ~ N(mu, sigma²):
# E[(D-q)+] = sigma * phi(z) + (mu-q) * [1-Phi(z)], z=(q-mu)/sigma
loss_norm <- Vectorize(function(s, mu = 0, sigma = 1){
  integrate(function(y) (y-s)*dnorm(y, mean=mu, sd = sigma), lower = s, upper=Inf)$value
})

# E[(q-D)+]
sales_norm <- Vectorize(function(s, mu = 0, sigma = 1){
  integrate(function(y) (s-y)*dnorm(y, mean=mu, sd = sigma), lower = -Inf, upper=s)$value
})

df_exp_cont <- data.frame(
  q = q_grid_cont,
  density = dnorm(q_grid_cont, mean = mu, sd = sigma),
  expected_loss = loss_norm(q_grid_cont, mu, sigma),
  expected_sales = sales_norm(q_grid_cont, mu, sigma)
)

plot_ly(
  df_exp_cont,
  x = ~q,
  y = ~expected_loss,
  type = "scatter",
  mode = "lines",
  name = "E(D - q)<sup>+</sup> - exp. shortage",
  line = list(color = "#D55E00", width = 3),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage = %{y:.1f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales,
    name = "E(q - D)<sup>+</sup> - exp. excess stock",
    line = list(color = "#0072B2", width = 3),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock = %{y:.1f}<extra></extra>"
  ) |>
  add_lines(
    y = ~density,
    name = "Density f(D)",
    yaxis = "y2",
    line = list(color = "#15a202", width = 1, dash = "dash"),
    hovertemplate = paste0(
      "Demand = %{x:.1f}",
      "<br>Density f(D) = %{y:.4f}",
      "<extra></extra>"
    )
  ) |>
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "expected quantity", rangemode = "tozero"),
    yaxis2 = list(
      title = "Density f(D)",
      overlaying = "y",
      side = "right",
      rangemode = "tozero",
      showgrid = FALSE,
      zeroline = FALSE,
      tickfont = list(color = "#15A202"),
      titlefont = list(color = "#15A202")
    ),
    margin = list(l = 70, r = 70, t = 45, b = 45),
    legend = list(orientation = "h", x = 0.5, y = 0.85),
    width = 650,
  height = 320
  )

```
:::
:::
:::


------------------------------------------------------------------------------

## Newsvendor model {#newsvendor-formal-solution}
::: slide-subtitle
Formal solution
:::

::: {.callout-note appearance="minimal" icon=false}
The optimal order quantity $q^*$ is where the slopes of the expected shortage costs $E((D-q)^+) \cdot c^u$ and the excess stock costs $E((q-D)^+)\cdot c^o$ are equal:
$$ \left(1 - F(q^*)\right) \cdot c^u \overset{!}{=} F(q^*) \cdot c^o \quad \Longrightarrow \quad q^* = F^{-1} \left( \frac{c^u}{c^o + c^u} \right)$$
:::


::: {.small-text}
::: {.columns}
::: {.column width="50%"}

::: {.fragment .fade-up}

**Discrete demand** [(distribution parameters cf. previous slide)](#newsvendor-fallbesipiel-lsg) with $c^u = 5$ and $c^o=3$

```{r newsvendor-discrete-plot2}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3.5

# Values from the slide “Case example: possible demand & production quantity”
d_disc <- c(30, 40, 50, 60)
p_disc <- c(0.20, 0.35, 0.30, 0.15)
cu = 5
co = 3

df_exp_disc$cost <- cu * df_exp_disc$expected_loss + co * df_exp_disc$expected_sales

q.star <- df_exp_disc$q[which.min(df_exp_disc$cost)]

plot_ly(
  df_exp_disc,
  x = ~q,
  y = ~expected_loss*cu,
  type = "scatter",
  mode = "lines+markers",
  name = "E[(D - q)<sup>+</sup>] - exp. shortage costs",
  line = list(color = "#D55E00", width = 3),
  marker = list(color = "#D55E00", size = 10),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage costs = %{y:.2f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales*co,
    name = "E[(q-D)<sup>+</sup>] - exp. excess stock costs",
    line = list(color = "#0072B2", width = 3),
    marker = list(color = "#0072B2", size = 10),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock costs = %{y:.2f}<extra></extra>"
  ) |>
  add_lines(
    y = ~cost,
    name = "E(P(D)) - exp. total costs",
    mode = "lines+markers",
    line = list(color = "#15a202", width = 3),
    marker = list(color = "#15a202", size = 10),
    hovertemplate = paste0(
      "Demand = %{x:.0f}",
      "<br>Density f(D) = %{y:.2f}",
      "<extra></extra>")
  ) |>
  add_lines(
    y = c(0,max(df_exp_disc$cost)),
    x = c(q.star,q.star),
    name = "opt. order quantity",
    mode = "lines",
    line = list(color = "#00000050", width = 2, dash = "dot"),
    hovertemplate = "opt. order quantity q* = %{x:.0f}"
  )|>
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "Expected costs", rangemode = "tozero"),
    margin = list(l = 70, r = 70, t = 15, b = 45),
    legend = list(orientation = "h", x = 0.3, y = 0.95),
    width = 650,
  height = 320
  )


```
:::
:::

::: {.column width="50%"}
::: {.fragment .fade-up}
**Continuous demand (Frankfurt store)** $D \sim N(280,55^2)$ with $c^u = 5$ and $c^o=3$

```{r newsvendor-continuous-plot2}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3.5


df_exp_cont$cost <- cu * df_exp_cont$expected_loss + co * df_exp_cont$expected_sales

q.star <- df_exp_cont$q[which.min(df_exp_cont$cost)]

plot_ly(
  df_exp_cont,
  x = ~q,
  y = ~expected_loss*cu,
  type = "scatter",
  mode = "lines",
  name = "c<sup>u</sup> &#183; E(D - q)<sup>+</sup> - exp. shortage costs",
  line = list(color = "#D55E00", width = 3),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage costs = %{y:.1f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales*co,
    name = "c<sup>o</sup> &#183; E(q - D)<sup>+</sup> - exp. excess stock costs",
    line = list(color = "#0072B2", width = 3),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock costs = %{y:.1f}<extra></extra>"
  ) |>
  add_lines(
    y = ~cost,
    name = "E(C(D)) - exp. total costs",
    mode = "lines+markers",
    line = list(color = "#15a202", width = 3),
    hovertemplate = paste0(
      "Demand = %{x:.0f}",
      "<br>Density f(D) = %{y:.2f}",
      "<extra></extra>")
  ) |>
  add_lines(
    y = c(0,max(df_exp_cont$cost)),
    x = c(q.star,q.star),
    name = "opt. order quantity",
    mode = "lines",
    line = list(color = "#00000050", width = 2, dash = "dot"),
    hovertemplate = "opt. order quantity q* = %{x:.0f}"
  ) |>
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "expected quantity", rangemode = "tozero"),
    margin = list(l = 50, r = 20, t = 15, b = 45),
    legend = list(orientation = "h", x = 0.5, y = 0.85),
    width = 650,
  height = 320
  )

```
:::
:::
:::
:::

------------------------------------------------------------------------------

## Newsvendor model {#newsvendor-critical-ratio}
::: slide-subtitle
Critical ratio
:::

::: callout-important

## Critical ratio

The optimal order quantity is the demand quantile whose cumulative probability equals the ratio of shortage costs to
total costs. The critical ratio thus determines the willingness to take risks

$$ CR = \frac{c^u}{c^o + c^u}$$
:::

::: {.small-text}
::: {.columns}

::: {.column width="32%"}
::: {.fragment .fade-up}
**Case 1: $c^u < c^o$**

::: incremental
- Shortage costs are lower than excess stock costs
- Excess stock should be avoided
- $\rightarrow$ opt. order quantity $q^*$ **smaller** than the median of the demand distribution
:::

:::
:::

::: {.column width="32%"}
::: {.fragment .fade-up}
**Case 2: $c^u = c^o$**

::: incremental
- Shortage costs = excess stock costs
- Shortages and excess stock equally important
- $\rightarrow$ opt. order quantity $q^*$ **equals** the median of the demand distribution
:::

:::
:::

::: {.column width="32%"}
::: {.fragment .fade-up}
**Case 3: $c^u > c^o$**

::: incremental
- Shortage costs are higher than excess stock costs
- Shortages should be avoided
- $\rightarrow$ opt. order quantity $q^*$ **larger** than the median of the demand distribution
:::
:::
:::

:::
:::

------------------------------------------------------------------------------

## Newsvendor model {#newsvendor-normal-distribution}
::: slide-subtitle
Optimal solution under normal distribution
:::

::: {.callout-note}

## Normally distributed demand

For normally distributed demand with $D \sim (\mu,\sigma^2)$, the optimal order quantity can be calculated by transforming to the standard normal distribution. For the optimal order quantity and costs:
$$z^* = \Phi^{-1} \left( \frac{c^u}{c^o + c^u} \right) \qquad q^*=\mu + \sigma \cdot z^* \qquad E(C(q^*))=(c^u+c^o)\cdot \sigma \cdot \varphi\left( z^* \right) $$
:::

**Case example** Frankfurt store with $N(280,55^2)$


::: {.small-text}
::: {.columns}

::: {.column width="32%"}
::: {.fragment .fade-up}
**Case 1: $c^u=1 < 5=c^o$**


```{r newsvendor-continuous-plot2a}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3.25

cu <- 1
co <- 5
df_exp_cont$cost <- cu * df_exp_cont$expected_loss + co * df_exp_cont$expected_sales

q.star <- df_exp_cont$q[which.min(df_exp_cont$cost)]

plot_ly(
  df_exp_cont,
  x = ~q,
  y = ~expected_loss*cu,
  type = "scatter",
  mode = "lines",
  name = "c<sup>u</sup> &#183; E(D - q)<sup>+</sup> - exp. shortage costs",
  line = list(color = "#D55E00", width = 3),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage costs = %{y:.1f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales*co,
    name = "c<sup>o</sup> &#183; E(q - D)<sup>+</sup> - exp. excess stock costs",
    line = list(color = "#0072B2", width = 3),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock costs = %{y:.1f}<extra></extra>"
  ) |>
  add_lines(
    y = ~cost,
    name = "E(C(D)) - exp. total costs",
    mode = "lines+markers",
    line = list(color = "#15a202", width = 3),
    hovertemplate = paste0(
      "Demand = %{x:.0f}",
      "<br>Density f(D) = %{y:.2f}",
      "<extra></extra>")
  ) |>
  add_lines(
    y = c(0,max(df_exp_cont$cost)),
    x = c(q.star,q.star),
    name = "opt. order quantity",
    mode = "lines",
    line = list(color = "#00000050", width = 2, dash = "dot"),
    hovertemplate = "opt. order quantity q* = %{x:.0f}"
  ) |> 
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "expected quantity", rangemode = "tozero"),
    margin = list(l = 50, r = 20, t = 15, b = 45),
    legend = list(orientation = "h", x = 0.5, y = 0.85),
    width = 500,
  height = 300
  )

```

$q^* = 280 + 55 \cdot$ `r round(qnorm(cu/(cu+co)),2)`=`r round(mu + sigma*qnorm(cu/(cu+co)),1)`
:::
:::

::: {.column width="32%"}
::: {.fragment .fade-up}
**Case 2: $c^u = 3 = c^o$**


```{r newsvendor-continuous-plot2b}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 3
#| fig-height: 3.25

cu <- 3
co <- 3
df_exp_cont$cost <- cu * df_exp_cont$expected_loss + co * df_exp_cont$expected_sales

q.star <- df_exp_cont$q[which.min(df_exp_cont$cost)]

plot_ly(
  df_exp_cont,
  x = ~q,
  y = ~expected_loss*cu,
  type = "scatter",
  mode = "lines",
  name = "c<sup>u</sup> &#183; E(D - q)<sup>+</sup> - exp. shortage costs",
  line = list(color = "#D55E00", width = 3),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage costs = %{y:.1f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales*co,
    name = "c<sup>o</sup> &#183; E(q - D)<sup>+</sup> - exp. excess stock costs",
    line = list(color = "#0072B2", width = 3),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock costs = %{y:.1f}<extra></extra>"
  ) |>
  add_lines(
    y = ~cost,
    name = "E(C(D)) - exp. total costs",
    mode = "lines+markers",
    line = list(color = "#15a202", width = 3),
    hovertemplate = paste0(
      "Demand = %{x:.0f}",
      "<br>Density f(D) = %{y:.2f}",
      "<extra></extra>")
  ) |>
  add_lines(
    y = c(0,max(df_exp_cont$cost)),
    x = c(q.star,q.star),
    name = "opt. order quantity",
    mode = "lines",
    line = list(color = "#00000050", width = 2, dash = "dot"),
    hovertemplate = "opt. order quantity q* = %{x:.0f}"
  ) |> 
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "expected quantity", rangemode = "tozero"),
    margin = list(l = 50, r = 20, t = 15, b = 45),
    legend = list(orientation = "h", x = 0.5, y = 0.85),
    width = 500,
  height = 300
  )

```
$q^* = 280 + 55 \cdot$ `r round(qnorm(cu/(cu+co)),2)`=`r round(mu + sigma*qnorm(cu/(cu+co)),1)`
:::
:::

::: {.column width="32%"}
::: {.fragment .fade-up}
**Case 3: $c^u=5 > 1=c^o$**


```{r newsvendor-continuous-plot2c}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 3
#| fig-height: 3.25

cu <- 5
co <- 1
df_exp_cont$cost <- cu * df_exp_cont$expected_loss + co * df_exp_cont$expected_sales

q.star <- df_exp_cont$q[which.min(df_exp_cont$cost)]

plot_ly(
  df_exp_cont,
  x = ~q,
  y = ~expected_loss*cu,
  type = "scatter",
  mode = "lines",
  name = "c<sup>u</sup> &#183; E(D - q)<sup>+</sup> - exp. shortage costs",
  line = list(color = "#D55E00", width = 3),
  hovertemplate = "Order quantity q = %{x:.1f}<br>exp. shortage costs = %{y:.1f}<extra></extra>"
) |>
  add_lines(
    y = ~expected_sales*co,
    name = "c<sup>o</sup> &#183; E(q - D)<sup>+</sup> - exp. excess stock costs",
    line = list(color = "#0072B2", width = 3),
    hovertemplate = "Order quantity q = %{x:.1f}<br>exp. excess stock costs = %{y:.1f}<extra></extra>"
  ) |>
  add_lines(
    y = ~cost,
    name = "E(C(D)) - exp. total costs",
    mode = "lines+markers",
    line = list(color = "#15a202", width = 3),
    hovertemplate = paste0(
      "Demand = %{x:.0f}",
      "<br>Density f(D) = %{y:.2f}",
      "<extra></extra>")
  ) |>
  add_lines(
    y = c(0,max(df_exp_cont$cost)),
    x = c(q.star,q.star),
    name = "opt. order quantity",
    mode = "lines",
    line = list(color = "#00000050", width = 2, dash = "dot"),
    hovertemplate = "opt. order quantity q* = %{x:.0f}"
  ) |> 
  layout(
    xaxis = list(title = "Order quantity q"),
    yaxis = list(title = "expected quantity", rangemode = "tozero"),
    margin = list(l = 50, r = 20, t = 15, b = 45),
    legend = list(orientation = "h", x = 0.5, y = 0.85),
    width = 500,
  height = 300
  )

```
$q^* = 280 + 55 \cdot$ `r round(qnorm(cu/(cu+co)),2)`=`r round(mu + sigma*qnorm(cu/(cu+co)),1)`
:::
:::

:::
::: 


------------------------------------------------------------------------------

## Newsvendor model {#newsvendor-formal-applications}
::: slide-subtitle
Typical applications
::: 

::: {.callout-note}

## Prerequisites

The newsvendor model is used particularly often when demand cannot be smoothed through inventory holding and shortage costs can be determined well
:::

Typical areas of application include, e.g. 

- seasonal and promotional items
- fresh products with a short shelf life
- tickets for events or transport
- ...

::: {.fragment .fade-up}
::: {.callout-important}
Beyond that, it is the basis for calculating cost-optimal safety stocks in inventory systems.
:::
:::
