

```{r}
#| include: false

library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)
library(plotly)
library(knitr)
library(kableExtra)
library(xml2)
library(datawizard)

# Load functions from scm_functions.R
source("../R/scm_functions.R")

set.seed(1234)

farbe_nachfrage <- "#1F4E79"
farbe_bestellung <- "#C00000"
farbe_bestand <- "#70AD47"
farbe_hintergrund <- "#F2F2F2"

plotly_layout <- function(p,
                          title = NULL,
                          x_title = NULL,
                          y_title = NULL,
                          showlegend = TRUE) {
  p |>
    layout(
      title = list(text = title, x = 0),
      xaxis = list(title = x_title),
      yaxis = list(title = y_title),
      showlegend = showlegend,
      legend = list(
        orientation = "h",
        x = 0,
        y = -0.15
      ),
      margin = list(l = 70, r = 30, b = 70, t = 70),
      paper_bgcolor = "white",
      plot_bgcolor = "white"
    )
}
```

## Uncertainty in supply chains  {#outline}

- Risk pooling
- Newsvendor model
- **Bullwhip effect**

------------------------------------------------------------------------------

## Learning objectives {#lernziele}

After this unit you will be able to …

- explain the bullwhip effect along a supply chain
- assess the main **causes and consequences** of the bullwhip effect
- determine the order-up-to level $S$ and safety stock $SB$
- classify measures to reduce the bullwhip effect

**Further reading**

- **Fundamental causes of and countermeasures to the bullwhip effect:** [@thonemann2023operations, Chapter 8, especially 8.2].
- **Empirical study of information transmission:** [@bray2012information].
- **Measurement, aggregation and impact of bullwhip effects:** [@yao2020empirically].

------------------------------------------------------------------------------

## Case example: sushi boxes {#fallbeispiel-intro}
::: slide-subtitle
Continued
::: 

::: {.callout-note}

## The SushiFresh supply chain

**SushiFresh GmbH** is a regional producer of fresh sushi boxes and supplies six supermarket stores in the Rhine-Main area. For production, SushiFresh depends on daily deliveries of fresh fish fillets supplied by the fish wholesaler **Nord-Fisch KG**. The sushi supply chain is therefore structured as follows:

$$
\text{Supermarket stores}
\mathrel{
  \mathop{\leftrightharpoons}
  \limits^{\text{Delivery}}_{\text{Order}}
}
\text{SushiFresh GmbH}
\mathrel{
  \mathop{\leftrightharpoons}
  \limits^{\text{Delivery}}_{\text{Order}}
}
\text{Nord-Fisch KG}
$$
:::


::: {.smaller-text}
::: columns
::: {.column width="30%"}

**Supermarket stores**

- observe end-customer demand (e.g. Frankfurt store: $\mu=280$, $\sigma=55$)
- order sushi boxes from SushiFresh daily 
:::

::: {.column width="30%"}
**SushiFresh GmbH**

- receives orders from the stores (= demand)
- orders fish for the sushi boxes from Nord-Fisch KG and keeps a stock of fish in reserve

:::

::: {.column width="30%"}
**Nord-Fisch KG**

- receives orders from SushiFresh GmbH (= demand)
- sources raw fish, produces the fish fillets and stores them in order to deliver quickly
:::
:::
:::

::: {.fragment .fade-up}
::: {.callout-important}
SushiFresh and Nord-Fisch each hold a (safety) stock of fish fillets to avoid shortages $\rightarrow$ How high should it be?
:::
:::

-----------------------------------------------------------------------------

## Case example: SushiFresh GmbH {#fallbeispiel-ordering-stocks}
::: slide-subtitle
Situation of SushiFresh GmbH
:::

::: {.callout-note}

## Ordering process for fish fillets

SushiFresh GmbH decides daily on its production quantity of sushi boxes for the Frankfurt store (cf. newsvendor model slides). Based on sushi box demand, let $d_t$ be the demand for fish fillet units on day $t$ (Frankfurt store: $\mu=280$, $\sigma=55$). SushiFresh can order replenishment from Nord-Fisch KG every day. Let the order quantity on day $t$ be $q_t$. Ordered fish fillets are delivered $l=2$ days later. Let the physically available quantity of fish fillet units be $i_t$. If no fish fillets are in stock, the demand is backordered ($b_t$) and produced or delivered later.
:::

::: {.fragment .fade-up}
**Sushi box inventories**

- the **physical inventory** $i_t$ is calculated as $i_t = \max(0,i_{t-1} + q_{t-2} - d_t)$
- the **net inventory** $n_t$ also accounts for shortages $n_t = i_{t-1} + q_{t-2} - d_t$
- the **inventory position** $p_t$ also accounts for open orders $p_t = n_{t-1} + q_{t-1} (+q_t)$
:::


------------------------------------------------------------------------------

## Case example: SushiFresh GmbH {#fallbeispiel-stock-overview}
::: slide-subtitle
Inventory of SushiFresh GmbH (Frankfurt store)
:::

:::: columns

::: {.column width="70%"}

```{r}
#| fig-width: 13
#| fig-height: 7

# Parameters and sample data: Frankfurt store (mu=280, sigma=55)
# Replenishment lead time: two periods (raw material sourcing + production/delivery)
# An order o_t arrives at the beginning of period t + 2.

dem <- c(234, 256, 344, 410, 278, 322, 212, 256, 278, 234, 256, 278)

i.ini <- 990
S.par <- 990

res <- tS.lager(d = dem, t = 1, S = S.par, l.ini = i.ini, wbz = 2)

#simuliere_tS(bedarf = dem, t=1, S = 30, anfangsbestand = 40, wbz = 2)

daten_bestand <- res |>
  as_tibble() |> 
  rename(Nettobestand= Lagerbestand, 'Disponibler_Bestand' = dis.LB, Bestellmenge = Bestellung, Liefermenge = Lieferung) |> 
  mutate('Physischer_Bestand' = pmax(Nettobestand,0),
         Periode = 1:nrow(res),
         offene_Bestellungen = cumsum(Bestellmenge)-cumsum(Liefermenge)
         ) |> 
  add_row(Periode = 0, Nettobestand = i.ini, 'Disponibler_Bestand' = i.ini, Bestellmenge = 0, Liefermenge = 0, 'Physischer_Bestand' = i.ini, offene_Bestellungen = 0, .before = 1) |>
  mutate(Nettobestand_vor_Bedarf = lag(Nettobestand))


```

```{r}
#| fig-width: 10
#| fig-height: 7
p_bestand <- plot_ly()

p_bestand <- p_bestand |>
  add_trace(
    data = daten_bestand,
    x = ~(Periode-.3),
    y = ~(-Bedarf),
    base = ~Nettobestand_vor_Bedarf,
    name = "Demand d[t]",
    type = "bar",
    width = 0.4,
    marker = list(
      color = "rgba(89, 89, 89, 0.35)",
      line = list(
        color = "rgba(89, 89, 89, 0.65)",
        width = 1
      )
    ),
    hovertemplate = paste0(
      "%{customdata}<br>"
    ),
    customdata = ~Bedarf
  ) |>
  add_trace(
    data = daten_bestand,
    x = ~(Periode-0.15),
    y = ~Liefermenge,
    base = ~(Nettobestand_vor_Bedarf-Bedarf),
    name = "Delivery quantity q[t-2]",
    type = "bar",
    width = 0.2,
    marker = list(
      color = "rgba(89, 89, 0, 0.35)",
      line = list(
        color = "rgba(89, 89, 0, 0.65)",
        width = 1
      )
    ),
    hovertemplate = paste0(
      "%{customdata}<br>"
    ),
    customdata = ~Liefermenge
  ) |>
  add_trace(
    data = daten_bestand,
    x = ~(Periode+0.2),
    y = ~(offene_Bestellungen-Bestellmenge),
    base = ~Nettobestand,
    name = "Open orders q[t-1]",
    type = "bar",
    width = 0.2,
    marker = list(
      color = "rgba(0, 89, 0, 0.35)",
      line = list(
        color = "rgba(0, 89, 0, 0.65)",
        width = 1
      )
    ),
    hovertemplate = paste0(
      "%{customdata}<br>"
    ),
    customdata = ~(offene_Bestellungen-Bestellmenge)
  ) |>
  add_trace(
    data = daten_bestand,
    x = ~(Periode+0.2),
    y = ~Bestellmenge,
    base = ~(Nettobestand+offene_Bestellungen-Bestellmenge),
    name = "Order quantity q[t]",
    type = "bar",
    width = 0.2,
    marker = list(
      color = "rgba(0, 89, 45, 0.35)",
      line = list(
        color = "rgba(0, 89, 45, 0.65)",
        width = 1
      )
    ),
    hovertemplate = paste0(
      "%{customdata}<br>"
    ),
    customdata = ~Bestellmenge
  ) |>
  add_trace(
    data = daten_bestand,
    x = ~Periode,
    y = ~Physischer_Bestand,
    name = "Physical inventory i[t]",
    type = "scatter",
    mode = "lines+markers",
    line = list(
      shape = "hvh",
      color = "#1F4E79",
      width = 4
    ),
    marker = list(
      color = "#1F4E79",
      size = 9
    ),
    hovertemplate = paste0(
      "Period %{x}<br>",
      "%{y}<extra></extra>"
    )
  ) |>
  add_trace(
    data = daten_bestand,
    x = ~Periode,
    y = ~Nettobestand,
    name = "Net inventory n[t]",
    type = "scatter",
    mode = "lines+markers",
    line = list(
      shape = "hvh",
      color = "#C00000",
      width = 4
    ),
    marker = list(
      color = "#C00000",
      size = 9
    ),
    hovertemplate = paste0(
      "Period %{x}<br>",
      "%{y}<extra></extra>"
    )
  ) |>
  add_trace(
    data = daten_bestand,
    x = ~Periode,
    y = ~Disponibler_Bestand,
    name = "Inventory position p[t]",
    type = "scatter",
    mode = "lines+markers",
    line = list(
      shape = "hvh",
      color = "#D97B29",
      width = 4,
      dash = "dash"
    ),
    marker = list(
      color = "#D97B29",
      size = 9,
      symbol = "diamond"
    ),
    hovertemplate = paste0(
      "Period %{x}<br>",
      "%{y}<extra></extra>"
    )
  ) |>
  layout(
    xaxis = list(
      title = "Period t",
      tickmode = "linear",
      tick0 = 1,
      dtick = 1
    ),
    yaxis = list(
      title = "Quantity",
      zeroline = TRUE,
      zerolinecolor = "#7C8792",
      zerolinewidth = 2
    ),
    barmode = "overlay",
    bargap = 0.25,
    legend = list(
      orientation = "h",
      x = 0,
      y = -0.22
    ),
    margin = list(
      l = 75,
      r = 35,
      b = 100,
      t = 80
    ),
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    hovermode = "x unified",
    width = 1100,
    hieght = 480
  )

p_bestand
```
:::

::: {.column width="30%"}

::: {.smaller-text}
**Demand**

```{r}
daten_bestand |>
  dplyr::rename(Nachfrage = Bedarf) |>
  dplyr::select(Periode, Nachfrage) |>
  knitr::kable(
    col.names = c("Period", "Demand"),
    align = "rr",
    digits = 0
  )
```

**Initial inventory** $i_0 =$ `r round(i.ini)`
**Order-up-to level** $S =$ `r round(S.par)`
:::
:::
::::

------------------------------------------------------------------------------

## Case example: SushiFresh GmbH {#fallbeispiel-order-up-to-level}
::: slide-subtitle
Setting the order-up-to level
:::

::: {.callout-note}

## Ordering process & order-up-to level $S$

Setting the order-up-to level $S$ determines the order quantity and the safety stock:

$$ q_t = S - p_t \qquad \qquad sb_t = S - \mu \cdot (l+1) $$
:::

::: {.small-text}
:::: columns

::: {.column width="50%"}
::: {.fragment .fade-up}

**for a given $\alpha$ service level** 

Extending the single-period formula (cf. pooling lecture) to multiple periods:

$$ S = \mu \cdot (l+1) + \sqrt{l+1} \cdot \sigma \cdot z_{\alpha}$$

:::
:::

::: {.column width="50%"}
::: {.fragment .fade-up}
**for a given shortage cost rate $c^u$ and inventory holding cost rate $c^o$ ** 

Applying the newsvendor formula:

$$ S = \mu \cdot (l+1) + \sqrt{l+1} \cdot \sigma \cdot z_{\frac{c^u}{c^u+c^o}}$$

:::
:::
::::
:::

::: {.fragment .fade-up}
::: {.callout-important}
Important: the parameters $\mu$ and $\sigma$ are the expected daily **demand (rate)** (demand per period) and the expected **deviation** (standard deviation). Both parameters are **estimated** on the basis of empirical customer demand.
::: 
:::


------------------------------------------------------------------------------

## Case example: SushiFresh GmbH {#fallbeispiel-order-decisions}
::: slide-subtitle
Order decisions and their effects
:::

::: {.callout-note}

## Estimating the demand parameters $\mu$ and $\sigma$

The parameters of the demand distribution are determined using statistical estimation methods (estimators), such as the (moving) average and the empirical standard deviation: 

$\hat{\mu} = \frac{1}{T} \sum_{\tau=t-T}^{t-1} d_\tau \qquad \qquad \hat{\sigma} = \sqrt{ \frac{1}{T-1} \sum_{\tau=t-T}^{t-1} (d_\tau - \hat{\mu})^2 } \rightarrow \;\text{forecast error}$ 
:::

::: {.smaller-text}

```{r}
#| echo: false
#| results: asis

# ---------------------------------------------------------------
# Parameters of the inventory and forecasting policy
# ---------------------------------------------------------------

n_perioden <- length(dem)

fenster <- 3L
wbz <- 2L
servicegrad <- 0.95
z_alpha <- qnorm(servicegrad)

anfangsbestand <- i.ini

# The order-up-to level is redetermined in every period.
# This illustrates the direct effect of parameter estimation
# on the order decision.
bestellintervall <- 1L

# ---------------------------------------------------------------
# Rolling parameter estimation
# For the decision in period t, only the demands of the
# previous periods are used:
# mu_hat[t]    = mean(d[t-fenster], ..., d[t-1])
# sigma_hat[t] = sd(d[t-fenster], ..., d[t-1])
# In the first periods, all observations available up to that
# point are used. With only one observation,
# sigma_hat is conservatively set to 0.
# ---------------------------------------------------------------

mu_hat <- numeric(n_perioden)
sigma_hat <- numeric(n_perioden)
n_beobachtungen <- integer(n_perioden)

for (t in seq_len(n_perioden)) {

  index_vorperiode <- seq_len(t - 1L)

  if (length(index_vorperiode) == 0L) {
    beobachtungen <- dem[2]
  } else {
    index_fenster <- tail(
      index_vorperiode,
      min(fenster, length(index_vorperiode))
    )

    beobachtungen <- dem[index_fenster]
  }

  n_beobachtungen[t] <- length(beobachtungen)
  mu_hat[t] <- mean(beobachtungen)

  sigma_hat[t] <- if (length(beobachtungen) > 1L) {
    sd(beobachtungen)
  } else {
    1
  }
}

# ---------------------------------------------------------------
# Dynamic order-up-to level
# With daily review, R = 1.
# The protection period is therefore:
# P = R + L = 1 + wbz
# ---------------------------------------------------------------

schutzperiode <- 1L + wbz

sicherheitsbestand <- z_alpha * sigma_hat * sqrt(schutzperiode)

S_t <- round(schutzperiode * mu_hat + sicherheitsbestand)

# ---------------------------------------------------------------
# Simulation of daily order decisions
# Sequence within period t:
# 1. Delivery q[t-wbz] arrives.
# 2. Demand dem[t] is realised.
# 3. Physical inventory and net inventory are updated.
# 4. Open orders are taken into account.
# 5. Stock is replenished up to the dynamic order-up-to level S_t.
# ---------------------------------------------------------------

liefermenge <- numeric(n_perioden)
bestellmenge <- numeric(n_perioden)
offene_bestellungen <- numeric(n_perioden)

physischer_bestand <- numeric(n_perioden)
nettobestand <- numeric(n_perioden)
disponibler_bestand <- numeric(n_perioden)

netto_vorperiode <- anfangsbestand

for (t in seq_len(n_perioden)) {

  liefermenge[t] <- if (t > wbz) {
    bestellmenge[t - wbz]
  } else {
    0
  }

  nettobestand[t] <- netto_vorperiode +
    liefermenge[t] -
    dem[t]

  physischer_bestand[t] <- max(
    0,
    nettobestand[t]
  )

  offene_bestellungen[t] <- if (wbz == 0L) {
    0
  } else {
    index_offen <- seq.int(
      from = max(1L, t - wbz + 1L),
      to = t - 1L
    )

    if (length(index_offen) == 0L || t == 1L) {
      0
    } else {
      sum(bestellmenge[index_offen])
    }
  }

  disponibler_bestand[t] <- nettobestand[t] +
    offene_bestellungen[t]

  bestellmenge[t] <- max(
    0,
    S_t[t] - disponibler_bestand[t]
  )

  netto_vorperiode <- nettobestand[t]
}

daten_order_decisions <- tibble(
  Periode = seq_len(n_perioden),
  Bedarf = dem,
  Anzahl_Beobachtungen = n_beobachtungen,
  Mittelwert_Schaetzung = mu_hat,
  Standardabweichung_Schaetzung = sigma_hat,
  Sicherheitsbestand = sicherheitsbestand,
  Zielbestand = S_t,
  Liefermenge = liefermenge,
  Physischer_Bestand = physischer_bestand,
  Nettobestand = nettobestand,
  Offene_Bestellungen = offene_bestellungen,
  Disponibler_Bestand = disponibler_bestand,
  Bestellmenge = bestellmenge
)

sd_bedarf <- sd(daten_order_decisions$Bedarf)

sd_bestellung <- sd(daten_order_decisions$Bestellmenge)

bullwhip_mass <- var(daten_order_decisions$Bestellmenge) /
  var(daten_order_decisions$Bedarf)

sd_vergleich <- tibble(
  Groesse = c(
    "End-customer demand d[t]",
    "Order quantity q[t]"
  ),
  Standardabweichung = c(
    sd_bedarf,
    sd_bestellung
  ),
  Farbe = c(
    farbe_nachfrage,
    farbe_bestellung
  )
)

daten_order_decisions |>
  dplyr::select(Periode, Bedarf, Mittelwert_Schaetzung,Standardabweichung_Schaetzung, Zielbestand, Bestellmenge) |> 
  dplyr::transmute(
    Day = Periode,
    'Demand d[t]' = round(Bedarf,1),
    `Mean μ̂[t]` = round(Mittelwert_Schaetzung,1),
    `Standard deviation σ̂[t]` = round(Standardabweichung_Schaetzung, 1),
    `Order-up-to level S[t]` = round(Zielbestand, 1),
    `Order q[t]` = round(Bestellmenge, 1)
  ) |>
  t() |> 
  row_to_colnames(row = 1) |> 
  kable_zeilenweise(
    align = c("l", rep("r", n_perioden)  ),
    digits = 1,
    table.attr = 'class="table table-sm table-striped"'
  )
```

:::

::: {.fragment .fade-up}
::: {.callout-important}

## Observation

The variability of demand $\sigma_d=$ `r round(sd_vergleich$Standardabweichung[1],1)` is considerably smaller than the standard deviation of the order quantities $\sigma_q=$ `r round(sd_vergleich$Standardabweichung[2],1)` $\rightarrow$ **additional uncertainty** is generated for Nord-Fisch KG
::: 
:::


------------------------------------------------------------------------------

##  Bullwhip effect
::: slide-subtitle
Definition & illustration
:::

::: {.callout-important}

## Definition [@cachon2024matching]

The bullwhip effect is present in a supply chain if the variability of demand at one level of the supply chain is greater than the variability of demand at the next downstream level in the supply chain, where variability is  measured with the coefficient of variation.

::: 

:::: columns

::: {.column width="60%"}

**Typical patterns**

```{r}
#| fig-width: 12
#| fig-height: 6

perioden <- 1:30

nachfrage <- c(
  rep(4, 5),
  rep(6, 25)
)

bestellung_handel <- c(
  rep(4, 4), 6, 8, 9, 8, 7, 6, 5, 4,
  4, 5, 6, 7, 7, 6, 5, 5, 5, 5,
  6, 6, 6, 6, 6, 6, 6, 6
)

bestellung_grosshandel <- c(
  rep(4, 5), 5, 8, 11, 13, 12, 10, 8,
  6, 4, 3, 3, 4, 5, 7, 8, 8, 7,
  6, 5, 5, 5, 6, 6, 6, 6 
)

daten_bullwhip <- tibble(
  Periode = perioden,
  Endkundennachfrage = nachfrage,
  Handel = bestellung_handel,
  Großhandel = bestellung_grosshandel
)


p_sushi_bullwhip <- daten_bullwhip |>
  plot_ly() |> 
  add_trace(
    x = ~Periode,
    y = ~Endkundennachfrage,
    name = "End-customer demand",
    type = "scatter",
    mode = "lines+markers",
    line = list(
      colour = "#1F4E79",
      width = 4
    ),
    marker = list(
      colour = "#1F4E79",
      size = 8
    )
  ) |>
  add_trace(
    x = ~Periode,
    y = ~Handel,
    name = "Order quantity retailer",
    type = "scatter",
    mode = "lines+markers",
    line = list(
      colour = "#ED7D31",
      width = 3
    ),
    marker = list(
      colour = "#ED7D31",
      size = 7
    )
  ) |>
  add_trace(
    x = ~Periode,
    y = ~Großhandel,
    type = "scatter",
    mode = "lines+markers",
    name = "Order quantity wholesaler",
    line = list(
      colour = "#C00000",
      width = 3
    ),
    marker = list(
      colour = "#C00000",
      size = 7
    )
  ) |>
  layout(
    xaxis = list(
      title = "Period",
      dtick = 1
    ),
    yaxis = list(
      title = "Order quantity",
      range = c(0,15)
    ),
    legend = list(
      orientation = "h",
      x = 0,
      y = 0.1
    ),
    margin = list(
      l = 70,
      r = 30,
      b = 50,
      t = 70
    ),
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    width = 1200,
    height = 480  
  )

p_sushi_bullwhip
```

:::

::: {.column width="39%"}
::: {.small-text}
**Upstream effects**

- “Amplification” of order quantities 
- Inventory volatility increases
- more frequent shortages
:::
:::

::::
------------------------------------------------------------------------------

##  Bullwhip effect
::: slide-subtitle
Typical observations
:::

::: {.small-text}
- usually $q_t \neq d_t$, since the order quantity contains several components $$
q_t=\text{expected demand}+\text{inventory correction}+\text{safety stock correction}.$$
- suppliers cannot tell whether fluctuations in order quantities indicate long-term shifts (i.e. end-customer demand is rising) or short-term inventory corrections $\rightarrow$  estimating future demand is difficult $\rightarrow$ **forecast errors increase** $\rightarrow$ order quantities fluctuate more
- hence the fluctuations of order quantities often increase with every stage of the value chain: $\operatorname{Var}(d_t)< \operatorname{Var}(q_t^{i})<\operatorname{Var}(q_t^{i+1})<\operatorname{Var}(q_t^{i+2})< ...$
:::

::: {.fragment .fade-up}
::: {.callout-important}

## Measuring the bullwhip effect

Measures of the bullwhip effect are the variance ratios $B=\frac{\operatorname{Var}(q_t)}{\operatorname{Var}(d_t)}>1$ and $B=\frac{\operatorname{Var}(q^{i}_t)}{\operatorname{Var}(q^{i-1}_t)}>1$ 

$\rightarrow$ the larger $B$, the stronger the bullwhip effect

::: 
:::


------------------------------------------------------------------------------

## Bullwhip effect

::: slide-subtitle
Empirical evidence [@bray2012information; @chen2012bullwhip; @yao2020empirically]
:::

::: columns
::: {.column width="48%"}

::: {.callout-note appearance="minimal" icon="false"}

### $2/3$ of companies

In a sample of **4,689 listed US companies**,
around two thirds showed a positive amplification of
procurement variability relative to sales variability.

:::

::: {.callout-important appearance="minimal" icon="false"}

### $+15.8\%$ on average

The mean bullwhip amplification was $15.8\%$ of
demand variability; the median was $6.7\%$.

:::

::: {.callout-warning appearance="minimal" icon="false"}

### Order of magnitude

On average, the quarterly standard deviation of upstream
orders exceeded that of observed demand
by around **USD 20 million**.

:::

:::

::: {.column width="48%"}

::: {.callout-tip appearance="minimal" icon="false"}

### Real performance consequences [@yao2020empirically]

A stronger bullwhip effect is associated with a higher inventory ratio and
at the same time more stockouts $\rightarrow$ safety stocks are insufficient

:::

::: {.callout-note appearance="minimal" icon="false"}

### Aggregation conceals the problem

Metrics aggregated across products, stores or time
underestimate the bullwhip effect compared with measurement at
SKU level.

:::

::: {.callout-important appearance="minimal" icon="false"}

### Management message

It is not only end-customer demand that creates volatility:

$$
\text{demand signal}
+
\text{local ordering decisions}
=
\text{order variance}
$$

:::

:::
:::


------------------------------------------------------------------------------

##  Bullwhip effect
::: slide-subtitle
Causes of the bullwhip effect
:::

::: {.small-text}

| Cause | Mechanism |
|---|---------|
| Local forecasts | Orders are interpreted as actual demand |
| Forecast updating | Small changes in demand lead to adjustments of order-up-to levels |
| Long lead times | Decisions must take effect far into the future |
| Order batching | Infrequent large orders create artificial demand peaks |
| Price promotions | Forward buying distorts observed demand |
| Rationing | Actors strategically order more than their actual demand |
| Lack of transparency | End-customer demand is not visible to all stages |
| Local systems of objectives | Stages optimise individually rather than system-wide |

:::

------------------------------------------------------------------------------

##  Bullwhip effect
::: slide-subtitle
Typical consequences
:::

::: columns
::: {.column width="50%"}

**Operational consequences**

- Fluctuating production schedules
- Unstable capacity utilisation
- Express shipments and special measures
- High inventories and yet shortages
- Lower delivery capability
- Greater planning uncertainty

:::

::: {.column width="50%"}
**Economic consequences**

- Higher capital tie-up
- Higher warehousing costs
- Higher obsolescence risks
- Higher set-up and production costs
- Higher transport costs
- Losses in revenue and contribution margin

:::
:::

------------------------------------------------------------------------------


##  Bullwhip effect
::: slide-subtitle
Countermeasures
:::

::: {.small-text}
| Lever | Measure | Effect on inventories and variance |
|--|-----|-----|
| Information flow | Share point-of-sale data | $\sigma$ of forecast errors decreases |
| Planning | Joint forecasts and CPFR | Local forecast bias decreases |
| Ordering policy | Small, frequent orders | Order batching decreases |
| Lead time | Shorten information and delivery lead times | Risk period decreases |
| Prices | Stable prices instead of forward-buying incentives | Artificial demand peaks decrease |
| Allocation | Rationing based on historical quantities | Strategic over-ordering decreases |
| Control | Overall optimum instead of local targets | Order and inventory decisions are coordinated |
:::
